### 2011-11-18 trunk/paper/drinking.tex
added manuscript

\documentclass{bioinfo}
\copyrightyear{2005}
\pubyear{2005}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\usepackage{pifont}
% Load packages
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage[utf8]{inputenc} %unicode support
\urlstyle{rm}
\usepackage{natbib}
\renewcommand{\cite}{\citep}
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{pifont}
%\newcommand{\tick}{\ding{52}}
\newcommand{\tick}{\checkmark}
\newcommand{\tickNo}{\hspace{1pt}\ding{55}}

\begin{document}
\firstpage{1}

\title{Process-based phenotypes: an application to the Cellular
  Phenotype Ontology}

\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Gabriella
  Rustici$^2$ and Midori Harris(?)$^3$ and Georgios V. Gkoutos$^{1}$}

\address{$^{1}$Department of Genetics, University of Cambridge,
  Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
  $^{2}$EBI, Wellcome Trust Genome Campus, Hinxton, Cambridge,
  Cambridge CB2 3EH, UK}

\maketitle

\begin{abstract}
  Phenotypes important for...

  Include behavioral phenotypes, based on behavior ontology

  We identify a problem in representing phenotypes such as drinking
  behavior that...

  provide solution...
\end{abstract}

\section{Introduction}

problem of phenotypes...

why reasoning, why formal, success,...

solved: structures, qualities

open: process phenotypes very complex...  
physiology...
measurements...

\subsection{Problem statement 1}
Heart rate increased: frequency of CMC increased fluid flow increased
volume per contraction increased ...

what is the relation to the heart?

\subsection{Problem statement 2}
increased apoptosis

% \subsection{Problem statement 3}
% A phenotype such as {\em Increased drinking behavior} is used for the
% annotation of phenotypes in mouse and other species.

% Different meanings:
% \begin{itemize}
% \item amount of substance increased over fixed period (e.g., 24h)
% \item amount of substance per drinking act is increased
% \item amount of time spent drinking over fixed period is increased
% \item amount of time spent per drinking act is increased
% \item number of drinking acts per fixed period increased
% \item variety of substances that organism drinks is increased (over
%   24h, per drinking act)
% \item substance flow during drinking increased
% \end{itemize}

% more than just these aspects of ``drinking behavior'' can be increased

\subsection{Observations}
observation 1: there is a fundamental distinction between attribution
of qualities for a fixed period (e.g., 24h) and single {\em act}s of
drinking.

observation 2: each of the different possible ``increased'' drinking
behavior may be the result of different underlying causes and due to
this, distinguishing between them may be important.

observation 3: depending on the experiment performed, only some of the
actual qualities can be measured.

observation 4: some qualities can be inferred given others. For
example,...

\section{Methods}
\subsection{Formal ontology}
The ontology as an approach to semantic standardisation was proposed
more than a decade ago and since then has become the dominant
methodology used to semantically categorise phenodeviance.  The
biomedical research community has invested considerable amount of
effort and resources in the development and establishment of
ontologies that are becoming increasingly successful as information
management and integration tools in many disparate scientific fields
allowing interoperability and semantic information processing between
diverse biomedical resources and domains.

In computer science, an ontology is a specification of a
conceptualization of a domain of knowledge \cite{Gruber1995,
  Guarino1998}.  % In philosophy, on the other hand, an ontology is a
% system of categories that account for a particular view on the
% world. The philosophical sense of ``ontology'' is what computer
% science calls ``conceptualization''.  While a conceptualization is
% language-independent, its specification is based on a language. In
% this sense, ontologies specify some aspects of the meaning of terms
% in vocabularies.
Ontologies commonly distinguish between {\em classes} and {\em
  individuals} within a domain of knowledge. A class is an entity that
can have {\em instances}, while individuals are entities that cannot
be instantiated. Examples of individuals include the Eiffel tower or
the 2009 Ironman Triathlon in Hawaii, while examples of classes
include {\em Tower} or {\em Triathlon}. The Eiffel tower can be an
instance of the class {\em Tower}, and the 2009 Ironman Triathlon an
instance of {\em Triathlon}.  The meaning of classes is specified by
stating what must be true of their instances.

In addition to classes and individuals, ontologies often include {\em
  relations}. Relations are the ``the glue that holds things together,
the primary constituents of the facts that go to make up reality''
\cite{}.

In {\em formal} ontologies, the specification of classes and relations
follows the axiomatic-deductive method. Given a set of terms that are
used within a domain and whose meaning we wish to specify, we begin by
providing {\em explicit definitions} for some terms, potentially
introducing new terms. An explicit definition of a term $t$ is a
statement that can replace every occurrence of $t$ in any sentence.

Eventually, a set of {\em primitive terms} remains that are not
further defined. Using only the primitive terms, we can construct
complex sentences. Based on the intended meaning of the primitive
terms, we consider some of these sentences true and some of them false
in our domain. We select some of the true sentences as {\em axioms}
which provide the core of our ontology. Ideally, the axioms are chosen
so that all true sentences in the domain we intend to represent follow
by means of logical deduction from the axioms. More commonly, however,
only {\em some aspects} of the intended meaning are formally
represented while other aspects are omitted either due to limitations
in language expressivity or due to their irrelevance to the problem
for which an ontology is developed.

Based on the axioms and definitions, we can use deduction to infer
statements that logically follow from the axioms.  The process of
automatically deducing sentences from axioms is called {\em automated
  reasoning}. Automated reasoning allows users of an ontology to carry
our key activities: verifying the ontology's consistency, inferring
hidden knowledge and thereby performing powerful queries.  An ontology
is inconsistent if {\em every} statement can be inferred from the
ontology. In particular, an ontology is inconsistent if a statement
$\phi$ and its negation $\neg \phi$ can be inferred from the
ontology's axioms. Classes in an ontology may be unsatisfiable: a
class $C$ is unsatisfiable, if it is not possible that the class has
instances. Unsatisfiable classes are commonly the result of a
contradictory class definition. That a class is unsatisfiable is often
a special kind of unintended consequence that can be drawn from an
ontology.

Automated reasoning in the Web Ontology Language (OWL) can be employed
to automatically compute the generalization hierarchy underlying an
ontology as well as for verification of data consistency and complex
queries \cite{Hoehndorf2011incon, Hoehndorf2011models}.

Layered approach: although OWL has high expressivity, some things
cannot be expressed in OWL; some things should not be expressed in OWL
for specific applications due to tractability reasons

Most applications rely on graph structure of ontology...

\subsection{Processes, their participants and qualities}
Processes are entities that are temporally extended: processes have
temporal parts and {\em unfold} through time. Examples of classes that
may have processes as instances include {\em Drinking}, {\em
  Triathlon} or {\em Apoptosis}.

Processes commonly have participants\footnote{It may be argued that
  processes necessarily have participants or that processes may have
  no participants. However, these discussions do not affect our aims
  here and take no position in this regard.}. Participants of
processes are entities that are present at time points (i.e., they
have no temporal parts) and may undergo changes during the process.
We can further distinguish different {\em roles} of participation in a
process. For example, a runner may participate in a {\em Triathlon}
process as a {\em Runner (role)}, while a particular {\em goal line}
can participate in the same process in a different role.

Amongst the participants of some processes, we will distinguish
between {\em inputs} and {\em outputs} of processes. Chemical
reactions, for example, are processes with definite inputs (the {\em
  reactants}) and outputs ({\em the products}).

\subsection{Gene Ontology}
quick intro...

mention regulation relations...

regulation of X processes

\subsection{PATO and the EQ model}

\section{Results}
\subsection{Attributes of processes}
An attribute that can be directly linked to a process is its {\em
  duration}. A duration is an attribute that characterizes the
temporal extent of a process and is similar to {\em Length}, {\em
  Area} and {\em Volume} for one-, two- and three-dimensional
entities. Furthermore, processes may be attributed a {\em temporal
  location} with respect to a particular coordinate system.

However, the majority of attributes that characterize processes are
not based on genuine process attributes alone, but rather relate
attributes of process participants with the duration of a process. In
particular, a {\em rate} typically refers to an attribute of some
entity {\em with respect to an attribute of another entity}, and in
the context of processes, rates often refer to attributes of a process
participant with respect to the duration of the process. For example,
a {\em mass flow rate} refers to the {\em Mass} of a process
participant with respect to the duration of the process, i.e., how
much matter is moved (from one point to another) through the process.
As a more complex example, a {\em rate of change of position} refers
to the {\em distance} that an object is moved with respect to the
duration of the process.

Not all rates of a process depend on attributes of the process
participants. In particular, a {\em frequency of occurrence} or {\em
  event rate} refers to the number of occurrences of a type of process
during the reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process containing the heart
with a duration of one minute). Further attributes that depend on
types of processes with regard to a reference process are {\em
  distribution patterns}, i.e., how the occurrences of processes of a
particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time.

Related to distribution patterns are {\em changing qualities} of
processes. For example, the rate of heart beating may change ({\em
  increase} or {\em decrease}) throughout the course of the reference
process. A simple analysis of {\em increasing} ({\em decreasing})
rates would be that the rate of a heart beating within the first half
of a process if {\em lower} ({\em higher}) than in the second half of
the process. Mathematically, this process sub-division can be iterated
until we end up with infinitesimally small time intervals.

While some processes can be subdivided indefinitely without changing
the type of process, others cannot. A class $C$ of {\em continuous
  processes} is a class that has processes $p$ as instance such that
all temporal parts of $p$ are also instances of $C$. Examples include
{\em continuous movements} or {\em regulation} processes. On the other
hand, {\em discrete processes} can be subdivided into stages of
activity and stages of inactivity: a class $C$ of {\em discrete
  processes} is a class with processes $p$ as instance and there is a
part of $p$ that is not an instance of $C$. An example is {\em heart
  beating} which has periods of activity (a single heart beat) and
inactivity.

% In line with the GO, we call processes that cause

% [Regulation of X vs. X: regulation modulates the number/amount of
% occurrence...]

In English language, we may say that an object has a {\em frequency}
or {\em rate} as if these were attributes of objects at time
points. For example, that my heart {\em now} beats with a frequency of
80 beats per minute, or that a car is moving with a speed of 280
kilometres per minute {\em at a particular point in time} (e.g., as
observed with a speed camera) may be considered attributes of the
objects, not attributes of processes. In our analysis, rates as
attributes of objects can be explicitly defined using rates of
processes. For example, the heart beating frequency of a particular
heart $h$ at a time point $t$ is the frequency of heart beating
processes in which $h$ participates at an infinitesimally small
process around $t$ and within a reference process. The reference
process is necessary in order to obtain a value for a frequency or
rate even if the process in question is not occurring. For example, a
rate of cardiac muscle contraction can be determined even when no
cardiac muscle contraction is occurring in virtue of a reference
process with a rate of cardiac muscle contraction.

The reference process for an object at a time point $t$ is uniquely
determined for continuous processes (the infinitesimally small process
around $t$), and ambiguous for discrete processes.

A similar construction as for continuous processes can be made for
discrete processes. A rate of blood flow is immediately an attribute
of a single heart beat process. However, during a period of time, the
cumulative rate of blood flow can be observed as the sum of the blood
flow rates of individual heart beating processes.

\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{processpatterns.pdf}
  \caption{Overview over ...\label{fig:patterns}}
\end{figure}

Complex relations between Based on these definition patterns,
interdependencies between process attributes can be observed.

\begin{table*}
  \begin{tabular}{l|l|l|l}
    & Increased blood flow rate & Normal blood flow rate & Decreased
    blood flow rate \\
    Increased frequency &increased&increased&?\\
    Normal frequency &increased&normal&decreased\\
    Decreased frequency &?&decreased&decreased
  \end{tabular}
  \caption{\label{...}Total blood flow rate...}
\end{table*}

\subsection{Cellular Phenotype Ontology}

A domain that would greatly benefit from the development of a CPO is
Systems Microscopy, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
(Lock \& Stromblad, Exp. Cell Res. 2010).

If we consider some of the studies, which have been published in this
field in the last few years (Neumann et al, Nature 464: 722-727;
Schmitz et al, Nature Cell Biology 12: 886-893; Fuchs et al, Mol Sys
Biology 6: 370), the need for a CPO becomes evident.

In all three studies here considered live-cell imaging assays and RNAi
knockdown were used to generate phenotypic profiles that quantify the
cellular response to a given siRNA thus allowing identification of
hundreds of genes involved in diverse biological functions including
cell division, migration and survival.

In each study, several phenotypes were detected and described by the
authors without the use of ontologies (see table), making the
integration between datasets extremely difficult. For example, it is
evident that cell division phenotypes were observed in the three
datasets (e.g. 'Mitotic delay/arrest' and 'Prolonged mitotic exit' or
'Methaphase delay' and 'Methaphase cells') but the overlap between
such phenotypes is unclear.

Data integration is also complicated by the lack of standardization at
the level of data production and processing; all these issues are
currently being address by the different groups involved in the
Systems Microscopy Network of Excellence
(http://www.systemsmicroscopy.eu/) and the first step towards data
integration will be achieved by developing a CPO.

This ontology will be used to integrate phenotypes' definitions across
existing datasets and will then become an integrated part of the data
processing pipeline and used to annotate the data as it gets generated
(Conrad et al, Nature Methods 8: 246-249).

\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth,
  height=.33\textheight]{overview.pdf}
  \caption{Overview over ...\label{fig:overview}}
\end{figure*}

CPO contains both structural abnormalities and physiological
abnormalities.

Phenotypes in the CPO are either based on structural abnormalities or
abnormal physiology involving cells or cell components. Structural
abnormalities in the CPO are based on GO's Cellular Component (GO-CC)
hierarchy. GO-CC contains 2,918 classes for cell parts (including {\em
  Cell}) and extracellular components of cells. For each cellular
component class $C$ in the GO-CC, we create a new class labelled {\em
  $C$ phenotype} in the CPO. For example, for the class {\em
  Mitochondrion} ({\tt GO:0005739}) in the GO-CC, we create the class
{\em Mitochondrion phenotype}.

Amongst the structural phenotype classes, we first distinguish between
{\em normal} and {\em abnormal} phenotypes. An {\em Abnormal phenotype
  of $C$} is a phenotype of an organism that does not have a normal
$C$ as part, while a {\em Normal phenotype of $C$} represents the
state in which an organism has a {\em normal $C$} as part.

Amongst the abnormal phenotypes that we include for all cell
components listed in GO-CC, we distinguish {\em abnormal morphology}
and {\em abnormal physiology} phenotypes. An {\em Abnormal morphology
  of $C$} is either the (abnormal) absence of required parts of $C$,
the (abnormal) presence of additional parts, or abnormal qualities of
$C$ \cite{Hoehndorf2010phene}. For example, an absence of caveolae
({\tt MP:...}) would be a subclass of {\em Abnormal morphology of
  plasma membrane} in virtue of caveolae necessarily being part of the
{\em Plasma membrane} ({\tt GO:0005886}).

Abnormal physiology of a cell component refers to abnormal {\em
  functionality} of a cell component. We assume that a functionality
of a cell component is (the potential for) a process in which the cell
component is (causally) involved. We use the definitions of GO classes
that were created based on lexical decompositions of GO class labels
\cite{Mungall2010go, Bada2007a, Ogren2004} to identify the processes
in which cell components are involved. For example, the definition of
the GO class {\em Mitochondrial fission} ({\tt GO:0000266}) is
explicitly defined as an {\em Organelle fission} ({\tt GO:0048285})
that {\bf results-in-the-division-of} a {\em Mitochondrion} ({\tt
  GO:0005739}). Based on this definition, we make the assumption that
{\em Mitochondrial fission} is one of the functions of a {\em
  Mitochondrion} and that an {\em Abnormality of mitochondrial
  fission} is a subclass of an {\em Abnormality of mitochondrion
  physiology}.

Amongst abnormal physiology, we distinguish between abnormalities in a
{\em single occurrence} of a cell component's functioning and an
abnormal {\em pattern of multiple occurrences} of a cell component's
functioning. For example, abnormalities in cell division resulting in
{\em Aneuploidy} refer to abnormalities of functioning processes,
while an {\em increased rate of cell division} refers to an
abnormality in the pattern of occurrence of multiple cell division
processes. In the CPO, abnormalities in the pattern of occurrence of
$X$ are represented as abnormalities in the {\em regulation of $X$}.

Single occurrences of processes can be abnormal in multiple ways,
depending on the type of process.
%
First, common to all processes is the quality of {\em duration} and
consequently, each process can have an {\em abnormal} (increased or
decreased) duration. The second common type of abnormality are
abnormalities based on process participants in relation to the
duration of the process. These include all kinds of {\em rates} such
as {\em mass flow rate}, {\em energy flow rate} and {\em velocity}
(the rate of change of position). In each of these cases, an object
participates in a process and a quality (or change of quality) of that
object throughout the duration of the process is considered to form a
new quality. If the process has participants that are distinguished
into {\em inputs} and {\em outputs}, then a recurring pattern is that
the amount of inputs or outputs that participate in the process can be
{\em increased} or {\em decreased}.  Finally, some objects may be
divided into stages during which particular states of affairs obtain,
and a process may be abnormal in that these states of affairs do not
obtain at a particular stage. Notably, at the beginning and the end of
a process, pre- and post-conditions may obtain that are abnormally
changed in a process. For example, {\em Aneuploidy} -- an abnormality
during cell division at which the chromosomes do not separate properly
between the two cells -- may be considered the result of such an
abnormality.

We implement the first two types of abnormality in the CPO. First, as
a subclass of each {\em Abnormality of P} class, we create {\em
  Abnormal duration of P}, which in turn has {\em Increased duration
  of P} and {\em Decreased duration of P} as subclasses. Second, if we
are able to identify {\em inputs} $I(P)$ or {\em outputs} $O(P)$ of
the process $P$ in the formal definitions of the GO, we automatically
generate {\em Abnormality of $I(P)$ in $P$} as well as {\em
  Abnormality of $O(P)$ in $P$}.  The left side of Figure
\ref{fig:overview} illustrates the schema of classes we generate for
single process abnormalities.

The second type of abnormality in the CPO relate to abnormalities of
{\em multiple} occurrences of some process $X$. According to the GO,
{\em regulation} processes are processes that maintain or modify the
occurrence of processes of a particular type. Following this
convention, we call an abnormality of multiple occurrences of $X$ {\em
  abnormality of the regulation of $X$}.

A first kind of abnormality of regulatory processes are {\em abnormal
  temporal distribution patterns} of a process. The most common
abnormal distribution pattern

We use PATO's {\em frequency} class to define {\em Abnormal frequency
  of occurrence of $X$} as a subclass of an {\em Abnormality of the
  regulation of $X$}.

\subsection{Cell Phenotype Ontology: implementation and evaluation}
We were faced with two choices for implementing the CPO: we could
either implement a pre-composed ontology in which all classes and
their definitions are pre-generated according to the patterns we
define, or we could develop an annotation software that enables the
selection of our process phenotype patterns based on the current
structure of the GO.  To maximize the utility and compatibility of the
CPO, and to provide stable identifiers for its concepts, we selected
the first strategy and developed a software to automatically generate
a pre-composed ontology from the GO.

We developed a software that utilizes the OWL API \cite{Horridge2007}
in order to generate an OWL representation of the CPO. The software
requires three input files: a version of the GO on which to based the
generated CPO, a version of PATO that is used to define abnormal
qualities, and a copy of the GO cross-product definitions
\cite{Mungall2010go} that is used to relate cell components to the
processes in which they participate as well as identify the
participants, inputs and outputs of processes.

We automatically generate a unique numerical identifier for each class
in the CPO.  Since the CPO is based on the GO and need to be updated
with subsequent versions of the GO, we must ensure to keep identifiers
stable in subsequent versions of CPO. Therefore, we use the
identifiers for GO classes to generate CPO class identifiers.

In the CPO, identifiers are of the form {\tt CPO:XXGGGGGGG}, where
{\tt GGGGGGG} is the seven-digit identifier of the GO class on which
the CPO class is based, and {\tt XX} is a prefix that identifies the
type of pattern we applied to the GO class. For example, based on the
class {\em Apoptotis} ({\tt GO:0006915}), we generate the CPO classes
{\em Abnormality of Apoptosis}, {\em Abnormality of single occurrence
  of apoptosis} and {\em Abnormality of regulation of apoptosis}.
We use the prefixes {\tt 12}, {\tt 14} and {\tt 15} for each of the
corresponding phenotype patterns, and consequently generate the class
identifiers {\tt CPO:120006915}, {\tt CPO:140006915} and {\tt
  CPO:150006915}.

\section{Discussion}
\subsection{Implementation as pre-composed ontology}
Pre-composed or software support for pattern?

\subsection{Use of CPO}

\subsection{Future research: acts and intentional states}
activities/actions, intentional (goal-oriented) acts

feeling thirsty (internal state) as driving force behind drinking
behavior

non-measurable, inference only

\subsection{Formalization}
Formal, axiomatic system for process qualities, maybe based on PSL,
FOIS paper 2010

one example:...

\section{Conclusion}

% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section*{Authors contributions}

% % %%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section*{Acknowledgements}
% The authors declare that they have no competing interests.

% % %%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section*{Acknowledgements}
% \ifthenelse{\boolean{publ}}{\small}{}

% The research work in its first unrevised form was presented at the
% SMBM 2010, Hinxton, Cambridge, U.K.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The Bibliography %%
%%                                                         %%
%% Bmc_article.bst will be used to %%
%% create a .BBL file for submission, which includes %%
%% XML structured for BMC.  %%
%%                                                         %%
%%                                                         %%
%% Note that the displayed Bibliography will not %%
%% necessarily be rendered by Latex exactly as specified %%
%% in the online Instructions for Authors.  %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{plain} % Style BST file
\bibliography{../bibtex/lc.bib}

\end{document}

### 2011-11-18 trunk/paper/drinking.tex

@@
 
-In English language, we may say that an object has a {\em frequency}
-or {\em rate} as if these were attributes of objects at time
-points. For example, that my heart {\em now} beats with a frequency of
-80 beats per minute, or that a car is moving with a speed of 280
-kilometres per minute {\em at a particular point in time} (e.g., as
-observed with a speed camera) may be considered attributes of the
-objects, not attributes of processes. In our analysis, rates as
-attributes of objects can be explicitly defined using rates of
-processes. For example, the heart beating frequency of a particular
-heart $h$ at a time point $t$ is the frequency of heart beating
-processes in which $h$ participates at an infinitesimally small
-process around $t$ and within a reference process. The reference
-process is necessary in order to obtain a value for a frequency or
-rate even if the process in question is not occurring. For example, a
-rate of cardiac muscle contraction can be determined even when no
-cardiac muscle contraction is occurring in virtue of a reference
-process with a rate of cardiac muscle contraction.
+We may say that an object has a {\em frequency} or {\em rate} as if
+these were attributes of objects at time points. For example, that my
+heart {\em now} beats with a frequency of 80 beats per minute, or that
+a car is moving with a speed of 280 kilometres per minute {\em at a
+  particular point in time} (e.g., as observed with a speed camera)
+may be considered attributes of the objects, not attributes of
+processes. In our analysis, rates as attributes of objects can be
+explicitly defined using rates of processes. For example, the heart
+beating frequency of a particular heart $h$ at a time point $t$ is the
+frequency of heart beating processes in which $h$ participates at an
+infinitesimally small process around $t$ and within a reference
+process. The reference process is necessary in order to obtain a value
+for a frequency or rate even if the process in question is not
+occurring. For example, a rate of cardiac muscle contraction can be
+determined even when no cardiac muscle contraction is occurring in
+virtue of a reference process with a rate of cardiac muscle
+contraction.
 
@@
 
-\subsection{Cell Phenotype Ontology: implementation and evaluation}
+\subsection{Cell Phenotype Ontology: implementation}
 We were faced with two choices for implementing the CPO: we could
@@
 
-In the CPO, identifiers are of the form {\tt CPO:XXGGGGGGG}, where
-{\tt GGGGGGG} is the seven-digit identifier of the GO class on which
-the CPO class is based, and {\tt XX} is a prefix that identifies the
-type of pattern we applied to the GO class. For example, based on the
-class {\em Apoptotis} ({\tt GO:0006915}), we generate the CPO classes
-{\em Abnormality of Apoptosis}, {\em Abnormality of single occurrence
-  of apoptosis} and {\em Abnormality of regulation of apoptosis}.
-We use the prefixes {\tt 12}, {\tt 14} and {\tt 15} for each of the
-corresponding phenotype patterns, and consequently generate the class
-identifiers {\tt CPO:120006915}, {\tt CPO:140006915} and {\tt
-  CPO:150006915}.
+In the CPO, identifiers contain two components and are of the form
+{\tt CPO:XXGGGGGGG}, where {\tt GGGGGGG} is the seven-digit identifier
+of the GO class on which the CPO class is based, and {\tt XX} is a
+prefix that identifies the type of phenotype pattern that is applied
+to the GO class. For example, based on the class {\em Apoptosis} ({\tt
+  GO:0006915}), we generate the CPO classes {\em Abnormality of
+  Apoptosis}, {\em Abnormality of single occurrence of apoptosis} and
+{\em Abnormality of regulation of apoptosis}.  We use the prefixes
+{\tt 12}, {\tt 14} and {\tt 15} for each of the corresponding
+phenotype patterns, and consequently generate the class identifiers
+{\tt CPO:120006915}, {\tt CPO:140006915} and {\tt CPO:150006915}. As
+long as the GO maintains its identifier for the {\em Apoptosis} class,
+the identifiers in the CPO will remain stable even when it is
+regenerated.
+
+We use the labels of GO classes to automatically generate class labels
+for phenotype classes as well as textual definitions for classes in
+the CPO. For example, the label of the class for increased number of
+occurrences of {\em Apoptosis} is {\em Increased frequency of
+  occurrences of Apoptosis}, and its textual definition states:
+``An increased frequency of occurrences of Apoptosis is a phenotype of
+entities in which the number of occurrences of Apoptosis within a
+given time period is increased in comparison to a reference
+process.''
 

### 2011-11-21 paper/drinking.tex
edited affiliation

--- 2011-11-18 trunk/paper/drinking.tex
@@
 \author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Gabriella
-  Rustici$^2$ and Midori Harris(?)$^3$ and Georgios V. Gkoutos$^{1}$}
+  Rustici$^2$ and Midori A. Harris$^3$ and Georgios V. Gkoutos$^{1}$}
 
@@
   $^{2}$EBI, Wellcome Trust Genome Campus, Hinxton, Cambridge,
-  Cambridge CB2 3EH, UK}
+  Cambridge CB10 1SD, UK\\
+  $^{3}$Department of Biochemistry; University of Cambridge, 80 Tennis Court Road, Cambridge CB2 1GA, UK}
 

### 2011-11-21 trunk/paper/drinking.tex

@@
   $^{2}$EBI, Wellcome Trust Genome Campus, Hinxton, Cambridge,
-  Cambridge CB2 3EH, UK}
+  Cambridge CB10 1SD, UK\\
+  $^{3}$Department of Biochemistry; University of Cambridge, 80 Tennis
+  Court Road, Cambridge CB2 1GA, UK}
 
@@
 increased apoptosis
-
-% \subsection{Problem statement 3}
-% A phenotype such as {\em Increased drinking behavior} is used for the
-% annotation of phenotypes in mouse and other species.
-
-% Different meanings:
-% \begin{itemize}
-% \item amount of substance increased over fixed period (e.g., 24h)
-% \item amount of substance per drinking act is increased
-% \item amount of time spent drinking over fixed period is increased
-% \item amount of time spent per drinking act is increased
-% \item number of drinking acts per fixed period increased
-% \item variety of substances that organism drinks is increased (over
-%   24h, per drinking act)
-% \item substance flow during drinking increased
-% \end{itemize}
-
-% more than just these aspects of ``drinking behavior'' can be increased
 
@@
 A first kind of abnormality of regulatory processes are {\em abnormal
-  temporal distribution patterns} of a process. The most common
-abnormal distribution pattern
-
-We use PATO's {\em frequency} class to define {\em Abnormal frequency
-  of occurrence of $X$} as a subclass of an {\em Abnormality of the
-  regulation of $X$}.
+  temporal distribution patterns} of a process. In these
+abnormalities, the {\em way} in which processes of a particular kind
+are temporally distributed is abnormal.  The most common abnormal
+distribution pattern is an increased or decreased frequency, and we
+use PATO's {\em frequency} class to define {\em Abnormal frequency of
+  occurrence of $X$}.
+% as a subclass of an {\em Abnormality of the regulation of $X$}.
 

### 2011-11-22 trunk/paper/drinking.tex

@@
 conceptualization of a domain of knowledge \cite{Gruber1995,
-  Guarino1998}.  % In philosophy, on the other hand, an ontology is a
-% system of categories that account for a particular view on the
-% world. The philosophical sense of ``ontology'' is what computer
-% science calls ``conceptualization''.  While a conceptualization is
-% language-independent, its specification is based on a language. In
-% this sense, ontologies specify some aspects of the meaning of terms
-% in vocabularies.
-Ontologies commonly distinguish between {\em classes} and {\em
-  individuals} within a domain of knowledge. A class is an entity that
-can have {\em instances}, while individuals are entities that cannot
-be instantiated. Examples of individuals include the Eiffel tower or
-the 2009 Ironman Triathlon in Hawaii, while examples of classes
-include {\em Tower} or {\em Triathlon}. The Eiffel tower can be an
-instance of the class {\em Tower}, and the 2009 Ironman Triathlon an
-instance of {\em Triathlon}.  The meaning of classes is specified by
-stating what must be true of their instances.
+  Guarino1998}.  Ontologies commonly distinguish between {\em classes}
+and {\em individuals} within a domain of knowledge. A class is an
+entity that can have {\em instances}, while individuals are entities
+that cannot be instantiated. Examples of individuals include the
+Eiffel tower or the 2009 Ironman Triathlon in Hawaii, while examples
+of classes include {\em Tower} or {\em Triathlon}. The Eiffel tower
+can be an instance of the class {\em Tower}, and the 2009 Ironman
+Triathlon an instance of {\em Triathlon}.  The meaning of classes is
+specified by stating what must be true of their instances.
 
@@
   occurrence of $X$}.
-% as a subclass of an {\em Abnormality of the regulation of $X$}.
+% \begin{verbatim}
+% has-phenotype some (has-part-some (participates-in some 
+%   (regulates some X and has-quality some (frequency and towards some X))))
+% \end{verbatim}
+For example, an {\em Abnormal frequency of occurrence of apoptosis} is
+defined as an abnormality of {\em Regulation of apoptosis} ({\tt
+  GO:0042981}) with respect to the {\em frequency} ({\tt
+  PATO:0000044}) of {\em Apoptosis} ({\tt GO:0006915}) occurrences.
+
+A second kind of regulatory abnormalities are related to the {\em
+  onset} of a process. With respect to a reference process, a
+particular kind of process may be {\em delayed} ({\tt PATO:0000502})
+or {\em premature} ({\tt PATO:0000694}). For example,
+
+Further kinds include deviations from distribution patterns. For
+example, a kind of process that is normally {\em rhythmic} can be
+abnormal in that it is {\em arrhythmic}...
 

### 2011-11-22 trunk/paper/drinking.tex

@@
 
+There are further types of deviation from a distribution pattern. For
+example, a kind of process that is normally {\em rhythmic} can be
+abnormal in that it is {\em arrhythmic}. A typical example of this
+kind of process is {\em Heart beating} ({\tt GO:0060047}), in which
+{\em Cardiac muscle contraction} ({\tt GO:0060048}) processes occur in
+a rhythmic pattern. In {\em Cardiac dysrhythmia}, however, {\em
+  Cardiac muscle contraction} processes occur arrhythmically, and we
+consider this to be an abnormality of the regulation of {\em Cardiac
+  muscle contraction}.
+
 A second kind of regulatory abnormalities are related to the {\em
@@
 particular kind of process may be {\em delayed} ({\tt PATO:0000502})
-or {\em premature} ({\tt PATO:0000694}). For example,
-
-Further kinds include deviations from distribution patterns. For
-example, a kind of process that is normally {\em rhythmic} can be
-abnormal in that it is {\em arrhythmic}...
+or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
+  apoptosis} refers to an abnormality of {\em Regulation of apoptosis}
+in which apoptosis is induced later than normal.
 

### 2011-11-22 paper/drinking.tex

@@
   Cambridge CB10 1SD, UK\\
-  $^{3}$Department of Biochemistry; University of Cambridge, 80 Tennis Court Road, Cambridge CB2 1GA, UK}
+  $^{3}$Department of Biochemistry; University of Cambridge, 80 Tennis
+  Court Road, Cambridge CB2 1GA, UK}
 
@@
 
-% \subsection{Problem statement 3}
-% A phenotype such as {\em Increased drinking behavior} is used for the
-% annotation of phenotypes in mouse and other species.
-
-% Different meanings:
-% \begin{itemize}
-% \item amount of substance increased over fixed period (e.g., 24h)
-% \item amount of substance per drinking act is increased
-% \item amount of time spent drinking over fixed period is increased
-% \item amount of time spent per drinking act is increased
-% \item number of drinking acts per fixed period increased
-% \item variety of substances that organism drinks is increased (over
-%   24h, per drinking act)
-% \item substance flow during drinking increased
-% \end{itemize}
-
-% more than just these aspects of ``drinking behavior'' can be increased
-
 \subsection{Observations}
@@
 A first kind of abnormality of regulatory processes are {\em abnormal
-  temporal distribution patterns} of a process. The most common
-abnormal distribution pattern
+  temporal distribution patterns} of a process. In these
+abnormalities, the {\em way} in which processes of a particular kind
+are temporally distributed is abnormal.  The most common abnormal
+distribution pattern is an increased or decreased frequency, and we
+use PATO's {\em frequency} class to define {\em Abnormal frequency of
+  occurrence of $X$}.
+% \begin{verbatim}
+% has-phenotype some (has-part-some (participates-in some 
+%   (regulates some X and has-quality some (frequency and towards some X))))
+% \end{verbatim}
+For example, an {\em Abnormal frequency of occurrence of apoptosis} is
+defined as an abnormality of {\em Regulation of apoptosis} ({\tt
+  GO:0042981}) with respect to the {\em frequency} ({\tt
+  PATO:0000044}) of {\em Apoptosis} ({\tt GO:0006915}) occurrences.
 
-We use PATO's {\em frequency} class to define {\em Abnormal frequency
-  of occurrence of $X$} as a subclass of an {\em Abnormality of the
-  regulation of $X$}.
+There are further types of deviation from a distribution pattern. For
+example, a kind of process that is normally {\em rhythmic} can be
+abnormal in that it is {\em arrhythmic}. A typical example of this
+kind of process is {\em Heart beating} ({\tt GO:0060047}), in which
+{\em Cardiac muscle contraction} ({\tt GO:0060048}) processes occur in
+a rhythmic pattern. In {\em Cardiac dysrhythmia}, however, {\em
+  Cardiac muscle contraction} processes occur arrhythmically, and we
+consider this to be an abnormality of the regulation of {\em Cardiac
+  muscle contraction}. While these abnormalities are often highly
+informative in clinical diagnostics and biological investigations, we
+usually lack the necessary information that is required to
+automatically determine meaningful types of abnormal distribution
+patterns.
+
+A second kind of regulatory abnormalities is related to the {\em
+  onset} of a process. With respect to a reference process, a
+particular kind of process may be {\em delayed} ({\tt PATO:0000502})
+or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
+  apoptosis} refers to an abnormality of the {\em Regulation of
+  apoptosis} in which apoptosis is induced later than normal.
+We use
 

### 2011-11-22 paper/drinking.tex

@@
 consequently, each process can have an {\em abnormal} (increased or
-decreased) duration. The second common type of abnormality are
-abnormalities based on process participants in relation to the
-duration of the process. These include all kinds of {\em rates} such
-as {\em mass flow rate}, {\em energy flow rate} and {\em velocity}
-(the rate of change of position). In each of these cases, an object
-participates in a process and a quality (or change of quality) of that
-object throughout the duration of the process is considered to form a
-new quality. If the process has participants that are distinguished
-into {\em inputs} and {\em outputs}, then a recurring pattern is that
-the amount of inputs or outputs that participate in the process can be
-{\em increased} or {\em decreased}.  Finally, some objects may be
-divided into stages during which particular states of affairs obtain,
-and a process may be abnormal in that these states of affairs do not
-obtain at a particular stage. Notably, at the beginning and the end of
-a process, pre- and post-conditions may obtain that are abnormally
-changed in a process. For example, {\em Aneuploidy} -- an abnormality
-during cell division at which the chromosomes do not separate properly
-between the two cells -- may be considered the result of such an
-abnormality.
+decreased) duration. For example, a part of an organism may
+participate in an {\em Inflammatory response} ({\tt GO:0006954}) that
+lasts longer than normal, i.e., the organism has an {\em Increased
+  duration of inflammatory response} phenotype. We define such a
+phenotype as a phenotype of an organism which has a part that
+participates in {\em Inflammatory response}, and this {\em
+  Inflammatory response} process has an {\em Increased duration} ({\tt
+  PATO:0000498}).
+
+The second common type of abnormality are abnormalities based on
+process participants in relation to the duration of the process. These
+include all kinds of {\em rates} such as {\em mass flow rate}, {\em
+  energy flow rate} and {\em velocity} (the rate of change of
+position). In each of these cases, an object participates in a process
+and a quality (or change of quality) of that object throughout the
+duration of the process is considered to form a new quality. If the
+process has participants that are distinguished into {\em inputs} and
+{\em outputs}, then a recurring pattern is that the amount of inputs
+or outputs that participate in the process can be {\em increased} or
+{\em decreased}. For example, an {\em Increased rate of cytoplasmic
+  streaming} is defined as
+
+Finally, some objects may be divided into stages during which
+particular states of affairs obtain, and a process may be abnormal in
+that these states of affairs do not obtain at a particular
+stage. Notably, at the beginning and the end of a process, pre- and
+post-conditions may obtain that are abnormally changed in a
+process. For example, {\em Aneuploidy} -- an abnormality during cell
+division at which the chromosomes do not separate properly between the
+two cells -- may be considered the result of such an abnormality.
 
@@
   apoptosis} refers to an abnormality of the {\em Regulation of
-  apoptosis} in which apoptosis is induced later than normal.
-We use
+  apoptosis} in which apoptosis is induced later than normal.  We use
+the PATO quality {\em onset} ({\tt PATO:...}) and its children {\em
+  delayed} and {\em premature} to define these types of regulatory
+abnormality.
+
+Finally, a third kind of regulatory abnormality refers to abnormal
+rates with respect to a participant of the process that is being
+regulated.
 

### 2011-11-23 paper/drinking.tex

@@
 \documentclass{bioinfo}
+\usepackage{setspace}
+
 \copyrightyear{2005}
@@
 
+{\bf Keywords:} cell phenotype, process, ontology, Semantic Web
+
 \section{Introduction}
@@
 
-\section{Methods}
+\section{Materials and Methods}
 \subsection{Formal ontology}
@@
 
-\begin{table*}
-  \begin{tabular}{l|l|l|l}
-    & Increased blood flow rate & Normal blood flow rate & Decreased
-    blood flow rate \\
-    Increased frequency &increased&increased&?\\
-    Normal frequency &increased&normal&decreased\\
-    Decreased frequency &?&decreased&decreased
-  \end{tabular}
-  \caption{\label{...}Total blood flow rate...}
-\end{table*}
-
 \subsection{Cellular Phenotype Ontology}
@@
 {\em decreased}. For example, an {\em Increased rate of cytoplasmic
-  streaming} is defined as
+  streaming} is defined as...
 
@@
   delayed} and {\em premature} to define these types of regulatory
-abnormality.
+abnormality. Similarly, we use PATO's {\em offset ({\tt PATO:...})
+  quality and its children to characterize regulatory abnormalities in
+  which a process ends prematurely or too late.
+
+\begin{table*}
+  \begin{tabular}{l|l|l|l}
+    & Increased blood flow rate & Normal blood flow rate & Decreased
+    blood flow rate \\
+    Increased frequency &increased&increased&?\\
+    Normal frequency &increased&normal&decreased\\
+    Decreased frequency &?&decreased&decreased
+  \end{tabular}
+  \caption{\label{...}Total blood flow rate...}
+\end{table*}
 
@@
 rates with respect to a participant of the process that is being
-regulated.
+regulated. For example, a cytoplasmic flow rate can
 
@@
 occurrences of {\em Apoptosis} is {\em Increased frequency of
-  occurrences of Apoptosis}, and its textual definition states:
-``An increased frequency of occurrences of Apoptosis is a phenotype of
+  occurrences of Apoptosis}, and its textual definition states: ``An
+increased frequency of occurrences of Apoptosis is a phenotype of
 entities in which the number of occurrences of Apoptosis within a
-given time period is increased in comparison to a reference
-process.''
+given time period is increased in comparison to a reference process.''
 
@@
 
-\section{Conclusion}
+\subsection{Conclusion}
+
+\section{Acknowledgements}
 
@@
 % % %%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section*{Acknowledgements}
 % The authors declare that they have no competing interests.
@@
 \bibliographystyle{plain} % Style BST file
-\bibliography{../bibtex/lc.bib}
+\bibliography{lc}
 

### 2011-11-23 trunk/paper/drinking.tex

@@
 consequently, each process can have an {\em abnormal} (increased or
-decreased) duration. The second common type of abnormality are
-abnormalities based on process participants in relation to the
-duration of the process. These include all kinds of {\em rates} such
-as {\em mass flow rate}, {\em energy flow rate} and {\em velocity}
-(the rate of change of position). In each of these cases, an object
-participates in a process and a quality (or change of quality) of that
-object throughout the duration of the process is considered to form a
-new quality. If the process has participants that are distinguished
-into {\em inputs} and {\em outputs}, then a recurring pattern is that
-the amount of inputs or outputs that participate in the process can be
-{\em increased} or {\em decreased}.  Finally, some objects may be
-divided into stages during which particular states of affairs obtain,
-and a process may be abnormal in that these states of affairs do not
-obtain at a particular stage. Notably, at the beginning and the end of
-a process, pre- and post-conditions may obtain that are abnormally
-changed in a process. For example, {\em Aneuploidy} -- an abnormality
-during cell division at which the chromosomes do not separate properly
-between the two cells -- may be considered the result of such an
-abnormality.
+decreased) duration. For example, a part of an organism may
+participate in an {\em Inflammatory response} ({\tt GO:0006954}) that
+lasts longer than normal, i.e., the organism has an {\em Increased
+  duration of inflammatory response} phenotype. We define such a
+phenotype as a phenotype of an organism which has a part that
+participates in {\em Inflammatory response}, and this {\em
+  Inflammatory response} process has an {\em Increased duration} ({\tt
+  PATO:0000498}).
+
+The second common type of abnormality are abnormalities based on
+process participants in relation to the duration of the process. These
+include all kinds of {\em rates} such as {\em mass flow rate}, {\em
+  energy flow rate} and {\em velocity} (the rate of change of
+position). In each of these cases, an object participates in a process
+and a quality (or change of quality) of that object throughout the
+duration of the process is considered to form a new quality. If the
+process has participants that are distinguished into {\em inputs} and
+{\em outputs}, then a recurring pattern is that the amount of inputs
+or outputs that participate in the process can be {\em increased} or
+{\em decreased}. For example, an {\em Increased rate of cytoplasmic
+  streaming} can be defined as an increased amount of inputs or an
+increased amount of outputs of a {\em cytoplasmic streaming} process.
+
+Finally, some objects may be divided into stages during which
+particular states of affairs obtain, and a process may be abnormal in
+that these states of affairs do not obtain at a particular
+stage. Notably, at the beginning and the end of a process, pre- and
+post-conditions may obtain that are abnormally changed in a
+process. For example, {\em Aneuploidy} -- an abnormality during cell
+division at which the chromosomes do not separate properly between the
+two cells -- may be considered the result of such an abnormality.
 
@@
 consider this to be an abnormality of the regulation of {\em Cardiac
-  muscle contraction}.
+  muscle contraction}. While these abnormalities are often highly
+informative in clinical diagnostics and biological investigations, we
+usually lack the necessary information that is required to
+automatically determine meaningful types of abnormal distribution
+patterns.
 
-A second kind of regulatory abnormalities are related to the {\em
+A second kind of regulatory abnormalities is related to the {\em
   onset} of a process. With respect to a reference process, a
@@
 or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
-  apoptosis} refers to an abnormality of {\em Regulation of apoptosis}
-in which apoptosis is induced later than normal.
+  apoptosis} refers to an abnormality of the {\em Regulation of
+  apoptosis} in which apoptosis is induced later than normal.  We use
+the PATO quality {\em onset} ({\tt PATO:...}) and its children {\em
+  delayed} and {\em premature} to define these types of regulatory
+abnormality. Similarly, we use PATO's {\em offset ({\tt PATO:...})
+  quality and its children to characterize regulatory abnormalities in
+  which a process ends prematurely or too late.
+
+\begin{table*}
+  \begin{tabular}{l|l|l|l}
+    & Increased blood flow rate & Normal blood flow rate & Decreased
+    blood flow rate \\
+    Increased frequency &increased&increased&?\\
+    Normal frequency &increased&normal&decreased\\
+    Decreased frequency &?&decreased&decreased
+  \end{tabular}
+  \caption{\label{...}Total blood flow rate...}
+\end{table*}
+
+Finally, a third kind of regulatory abnormality refers to abnormal
+rates with respect to a participant of the process that is being
+regulated. For example, a cytoplasmic flow rate can be increased or
+decreased not within a single {\em cytoplasmic streaming} process but
+rather the total flow rate, as a summation over all cytoplasmic
+streaming processes that occur within an organism (or a particular
+anatomical location), is increased or decreased.
 

### 2011-11-23 trunk/paper/drinking.tex

@@
   delayed} and {\em premature} to define these types of regulatory
-abnormality. Similarly, we use PATO's {\em offset ({\tt PATO:...})
+abnormality. Similarly, we use PATO's {\em offset} ({\tt PATO:...})
   quality and its children to characterize regulatory abnormalities in

### 2011-11-23 paper/drinking.tex

@@
 {\em decreased}. For example, an {\em Increased rate of cytoplasmic
-  streaming} is defined as...
+  streaming} can be defined as an increased amount of inputs or an
+increased amount of outputs of a {\em cytoplasmic streaming} process.
 
@@
 \begin{table*}
+  \centering
   \begin{tabular}{l|l|l|l}
-    & Increased blood flow rate & Normal blood flow rate & Decreased
-    blood flow rate \\
-    Increased frequency &increased&increased&?\\
-    Normal frequency &increased&normal&decreased\\
-    Decreased frequency &?&decreased&decreased
+    & Increased cytoplasmic flow rate & Normal cytoplasmic flow rate & Decreased
+    cytoplasmic flow rate \\
+    \hline
+    Increased frequency &increased total flow rate &increased total
+    flow rate &?\\
+    Normal frequency &increased total flow rate &normal total flow
+    rate &decreased total flow rate\\
+    Decreased frequency &?&decreased total flow rate &decreased total
+    flow rate\\
+    \hline
   \end{tabular}
-  \caption{\label{...}Total blood flow rate...}
+  \caption{\label{tbl:flow}Total cytoplasmic flow rate...}
 \end{table*}
@@
 rates with respect to a participant of the process that is being
-regulated. For example, a cytoplasmic flow rate can
+regulated. For example, a cytoplasmic flow rate can be increased or
+decreased not within a single {\em cytoplasmic streaming} process but
+rather the total cytoplasmic flow rate, as a summation over all
+cytoplasmic streaming processes that occur within an organism (or a
+particular anatomical location), is increased or decreased. While a
+flow rate of a single {\em cytoplasmic streaming} process is a quality
+of that process, an increased {\em total} cytoplasmic flow rate is a
+quality of the regulation of {\em cytoplasmic streaming}. In
+particular, it is possible for an organism to have a normal --- or
+even a decreased --- cytoplasmic flow rate in each individual
+cytoplasmic streaming process while at the same time having an
+increased total cytoplasmic flow rate due to a large increase in the
+frequency of occurrence of cytoplasmic streaming processes. Similarly,
+the frequency of occurrence of cytoplasmic streaming may be normal or
+decreased while the total cytoplasmic flow rate is increased due to an
+increased cytoplasmic flow rate in each individual cytoplasmic
+streaming process. Table \ref{tbl:flow} illustrates the dependencies
+between rates of individual processes, their frequency of occurrence
+and the total rate of these processes. We include total rates as
+regulatory abnormalities in the CPO since these are the attributes of
+processes that are often measured or observed, while the rates of
+individual processes are inferred following a schema such as Table
+\ref{tbl:flow}.
 

### 2011-11-23 paper/drinking.tex
added GO blurb, including regulation stuff

@@
 \subsection{Gene Ontology}
+
+The Gene Ontology provides a set of ontologies for molecular and
+cellular biology, originally designed to support structured
+annotations for genes and gene products in all species with respect to
+molecular function (MF), biological process (BP) and cellular
+component (CC). MF and BP both describe processes, but at different
+spatiotemporal scales; in particular, BP includes processes that
+unfold within cells and within tissues and organs of multicellular
+organisms. Gene product annotations identify participants in the
+processes.
+
+Over time, GO development has increasingly emphasized a normalized
+approach that includes supplementing existing human-readable text
+description with formally specified explicit definitions for GO
+classes. The formalization of GO is readily apparent in its
+representation of biological regulation.
+
+Regulatory processes may regulate other processes, at either the MF or
+BP scale, or biological qualities. GO accordingly includes three broad
+categories of regulation terms, regulation of molecular function,
+regulation of biological process, and regulation of biological
+quality. The first two are explicitly defined entirely with respect to
+other GO terms, whereas the third comprises classes in which the
+regulated qualities are specified by terms from PATO (see below) or
+anatomy ontologies.
+
+All GO regulation terms use one of three relations, regulates,
+negatively_regulates and positively_regulates, to link regulation
+terms to process or quality terms. The regulates relations are defined
+in terms of qualities: a regulatory process causes a change in
+magnitude to some quality, which in turn has an effect on the
+frequency, rate or duration of some other type of process. Effects
+that results in increases and decreases use positively_regulates and
+negatively_regulates respectively. \cite{} (cite the paper on
+cross-products in GO, PMID:20152934) The existing ontology structure
+would also support the addition of subclasses to distinguish, for
+example, regulation of the rate of a process from regulation of its
+duration or time of onset.
+
 quick intro...

### 2011-11-25 paper/drinking.tex

@@
 process. For example, a runner may participate in a {\em Triathlon}
-process as a {\em Runner (role)}, while a particular {\em goal line}
-can participate in the same process in a different role.
+process as a {\em Runner (role)}, while a {\em Referee} can
+participate in the same process but in a different role.
 
@@
 All GO regulation terms use one of three relations, regulates,
-negatively_regulates and positively_regulates, to link regulation
+negatively\_regulates and positively\_regulates, to link regulation
 terms to process or quality terms. The regulates relations are defined
@@
 frequency, rate or duration of some other type of process. Effects
-that results in increases and decreases use positively_regulates and
-negatively_regulates respectively. \cite{} (cite the paper on
+that results in increases and decreases use positively\_regulates and
+negatively\_regulates respectively. \cite{} (cite the paper on
 cross-products in GO, PMID:20152934) The existing ontology structure
@@
 \subsection{PATO and the EQ model}
+The Phenotype And Trait Ontology (PATO) was envisaged and designed to
+provide a platform for allowing the integration of quantitative and
+qualitative phenotype related information across different levels of
+granularity of domain knowledge granularity and species
+\cite{Gkoutos2005}.  PATO forms an ontology of phenotype qualities
+that form basic entities that we can perceive and/or measure such as
+colors, sizes, rates etc. One of it's classification axis lies on the
+basis of the basic type of entity these qualities inhere to and
+distinguishes between qualities of physical objects and qualities of
+processes.
+
+PATO allows for the description of affected entities by combining
+various ontologies that describe the entities that have been affected,
+such as the various anatomical ontologies, GO\cite {Ashburner2000b},
+the Cell Type Ontology\cite {Bard2005a}, (SO)\cite {Eilbeck2005} etc
+with the various qualities it provides for defining how these entities
+were affected.  PATO can be used for annotation either directly in a
+so called post-composed (post-coordinated) manner or for providing
+formal (logical) definitions (equivalence axioms) to ontologies
+containing a set of precomposed (pre-coordinated) phenotype terms. For
+instance, to describe the decrease in the length of the sexual cycle
+of female animals, we can combine the PATO term \textit{decreased
+  duration} (PATO:0000499) with the Gene Ontology term \textit{estrous
+  cycle} (GO:0042698), whilst if such a term existed in a precomposed
+ontology (for the MP:0009007 term from the Mammalian Phenotype) it
+could be used to provide an equivalence statement between that class
+and the above PATO-based description.
 

### 2011-11-25 paper/drinking.tex

@@
 ontology as well as for verification of data consistency and complex
-queries \cite{Hoehndorf2011incon, Hoehndorf2011models}.
+queries \cite{Hoehndorf2011incon, Hoehndorf2011models}. Highly
+efficient automated reasoners are available to process OWL ontologies
+\cite{Sirin2004, Tsarkov2006, Motik2009a}. OWL profiles were developed
+to support even large ontologies by further reducing the expressivity
+of OWL in order to enable polynomial-time inferences. In particular
+the OWL EL profile was found to provide the expressivity required for
+most biomedical ontologies \cite{el4, elvira}, and highly optimized
+OWL EL reasoners are available or under development to support
+reasoning over very large ontologies \cite{el4, cbreasoner}.
 
-Layered approach: although OWL has high expressivity, some things
+A high expressivity is required to accurately specify complex axioms
+that constrain the domain under investigation, and languages with
+higher expressivity than OWL are often required in the biomedical
+domain to achieve this goal \cite{Hoehndorf2009sequences,
+  rnao}. On the other hand, automated reasoning over large ontologies
+and associated datasets benefits from languages with a low complexity
+of inferences in which complex axioms cannot be formulated. Therefore,
+we use a layered approach: we specify the meaning of terms
+
+although OWL has high expressivity, some things
 cannot be expressed in OWL; some things should not be expressed in OWL
@@
 
-All GO regulation terms use one of three relations, regulates,
-negatively\_regulates and positively\_regulates, to link regulation
-terms to process or quality terms. The regulates relations are defined
-in terms of qualities: a regulatory process causes a change in
-magnitude to some quality, which in turn has an effect on the
-frequency, rate or duration of some other type of process. Effects
-that results in increases and decreases use positively\_regulates and
-negatively\_regulates respectively. \cite{} (cite the paper on
-cross-products in GO, PMID:20152934) The existing ontology structure
-would also support the addition of subclasses to distinguish, for
-example, regulation of the rate of a process from regulation of its
-duration or time of onset.
-
-quick intro...
-
-mention regulation relations...
-
-regulation of X processes
+All GO regulation terms use one of three relations, {\bf regulates},
+{\bf negatively\_regulates} and {\bf positively\_regulates}, to link
+regulation terms to process or quality terms. The {\bf regulates}
+relations are defined in terms of qualities: a regulatory process
+causes a change in magnitude to some quality, which in turn has an
+effect on the frequency, rate or duration of some other type of
+process. Effects that results in increases and decreases use {\bf
+  positively\_regulates} and {\bf negatively\_regulates} respectively
+\cite{Mungall2010go}. The existing ontology structure would also
+support the addition of subclasses to distinguish, for example,
+regulation of the rate of a process from regulation of its duration or
+time of onset.
 
@@
 of female animals, we can combine the PATO term \textit{decreased
-  duration} (PATO:0000499) with the Gene Ontology term \textit{estrous
-  cycle} (GO:0042698), whilst if such a term existed in a precomposed
-ontology (for the MP:0009007 term from the Mammalian Phenotype) it
-could be used to provide an equivalence statement between that class
-and the above PATO-based description.
+  duration} ({\tt PATO:0000499}) with the Gene Ontology term
+\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
+existed in a pre-composed ontology (for the {\tt MP:0009007} term from
+the Mammalian Phenotype) it could be used to provide an equivalence
+statement between that class and the above PATO-based description.
 

### 2011-11-25 paper/drinking.tex

@@
 higher expressivity than OWL are often required in the biomedical
-domain to achieve this goal \cite{Hoehndorf2009sequences,
-  rnao}. On the other hand, automated reasoning over large ontologies
-and associated datasets benefits from languages with a low complexity
-of inferences in which complex axioms cannot be formulated. Therefore,
-we use a layered approach: we specify the meaning of terms
-
-although OWL has high expressivity, some things
-cannot be expressed in OWL; some things should not be expressed in OWL
-for specific applications due to tractability reasons
-
-Most applications rely on graph structure of ontology...
+domain to achieve this goal \cite{Hoehndorf2009sequences, rnao}. On
+the other hand, automated reasoning over large ontologies and
+associated datasets benefits from languages with a low complexity of
+inferences in which complex axioms cannot be formulated. Therefore, a
+possible solution is to use a layered approach: to specify the meaning
+of terms using an expressive language, and derive the axioms that
+must obtain in a weaker language using deductive inferences.
 
@@
 process. For example, a runner may participate in a {\em Triathlon}
-process as a {\em Runner (role)}, while a {\em Referee} can
-participate in the same process but in a different role.
+process as a {\em Runner (role)}, while another person can participate
+in the same process but in a different role (such as {\em Referee}).
 
@@
 \subsection{Gene Ontology}
-
 The Gene Ontology provides a set of ontologies for molecular and

### 2011-11-25 trunk/paper/drinking.tex

@@
 ontology as well as for verification of data consistency and complex
-queries \cite{Hoehndorf2011incon, Hoehndorf2011models}.
+queries \cite{Hoehndorf2011incon, Hoehndorf2011models}. Highly
+efficient automated reasoners are available to process OWL ontologies
+\cite{Sirin2004, Tsarkov2006, Motik2009a}. OWL profiles were developed
+to support even large ontologies by further reducing the expressivity
+of OWL in order to enable polynomial-time inferences. In particular
+the OWL EL profile was found to provide the expressivity required for
+most biomedical ontologies \cite{el4, elvira}, and highly optimized
+OWL EL reasoners are available or under development to support
+reasoning over very large ontologies \cite{el4, cbreasoner}.
 
-Layered approach: although OWL has high expressivity, some things
-cannot be expressed in OWL; some things should not be expressed in OWL
-for specific applications due to tractability reasons
-
-Most applications rely on graph structure of ontology...
+A high expressivity is required to accurately specify complex axioms
+that constrain the domain under investigation, and languages with
+higher expressivity than OWL are often required in the biomedical
+domain to achieve this goal \cite{Hoehndorf2009sequences, rnao}. On
+the other hand, automated reasoning over large ontologies and
+associated datasets benefits from languages with a low complexity of
+inferences in which complex axioms cannot be formulated. Therefore, a
+possible solution is to use a layered approach: to specify the meaning
+of terms using an expressive language, and derive the axioms that
+must obtain in a weaker language using deductive inferences.
 
@@
 process. For example, a runner may participate in a {\em Triathlon}
-process as a {\em Runner (role)}, while a particular {\em goal line}
-can participate in the same process in a different role.
+process as a {\em Runner (role)}, while another person can participate
+in the same process but in a different role (such as {\em Referee}).
 
@@
 \subsection{Gene Ontology}
-quick intro...
+The Gene Ontology provides a set of ontologies for molecular and
+cellular biology, originally designed to support structured
+annotations for genes and gene products in all species with respect to
+molecular function (MF), biological process (BP) and cellular
+component (CC). MF and BP both describe processes, but at different
+spatiotemporal scales; in particular, BP includes processes that
+unfold within cells and within tissues and organs of multicellular
+organisms. Gene product annotations identify participants in the
+processes.
 
-mention regulation relations...
+Over time, GO development has increasingly emphasized a normalized
+approach that includes supplementing existing human-readable text
+description with formally specified explicit definitions for GO
+classes. The formalization of GO is readily apparent in its
+representation of biological regulation.
 
-regulation of X processes
+Regulatory processes may regulate other processes, at either the MF or
+BP scale, or biological qualities. GO accordingly includes three broad
+categories of regulation terms, regulation of molecular function,
+regulation of biological process, and regulation of biological
+quality. The first two are explicitly defined entirely with respect to
+other GO terms, whereas the third comprises classes in which the
+regulated qualities are specified by terms from PATO (see below) or
+anatomy ontologies.
+
+All GO regulation terms use one of three relations, {\bf regulates},
+{\bf negatively\_regulates} and {\bf positively\_regulates}, to link
+regulation terms to process or quality terms. The {\bf regulates}
+relations are defined in terms of qualities: a regulatory process
+causes a change in magnitude to some quality, which in turn has an
+effect on the frequency, rate or duration of some other type of
+process. Effects that results in increases and decreases use {\bf
+  positively\_regulates} and {\bf negatively\_regulates} respectively
+\cite{Mungall2010go}. The existing ontology structure would also
+support the addition of subclasses to distinguish, for example,
+regulation of the rate of a process from regulation of its duration or
+time of onset.
 
 \subsection{PATO and the EQ model}
+The Phenotype And Trait Ontology (PATO) was envisaged and designed to
+provide a platform for allowing the integration of quantitative and
+qualitative phenotype related information across different levels of
+granularity of domain knowledge granularity and species
+\cite{Gkoutos2005}.  PATO forms an ontology of phenotype qualities
+that form basic entities that we can perceive and/or measure such as
+colors, sizes, rates etc. One of it's classification axis lies on the
+basis of the basic type of entity these qualities inhere to and
+distinguishes between qualities of physical objects and qualities of
+processes.
+
+PATO allows for the description of affected entities by combining
+various ontologies that describe the entities that have been affected,
+such as the various anatomical ontologies, GO \cite {Ashburner2000b},
+the Cell Type Ontology \cite {Bard2005a}, SO \cite {Eilbeck2005} etc
+with the various qualities it provides for defining how these entities
+were affected.  PATO can be used for annotation either directly in a
+so called post-composed (post-coordinated) manner or for providing
+formal (logical) definitions (equivalence axioms) to ontologies
+containing a set of precomposed (pre-coordinated) phenotype terms. For
+instance, to describe the decrease in the length of the sexual cycle
+of female animals, we can combine the PATO term \textit{decreased
+  duration} ({\tt PATO:0000499}) with the Gene Ontology term
+\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
+existed in a pre-composed ontology (for the {\tt MP:0009007} term from
+the Mammalian Phenotype) it could be used to provide an equivalence
+statement between that class and the above PATO-based description.
 
@@
 \subsection{Cellular Phenotype Ontology}
-
 A domain that would greatly benefit from the development of a CPO is
@@
 decreased not within a single {\em cytoplasmic streaming} process but
-rather the total flow rate, as a summation over all cytoplasmic
-streaming processes that occur within an organism (or a particular
-anatomical location), is increased or decreased.
+rather the total cytoplasmic flow rate, as a summation over all
+cytoplasmic streaming processes that occur within an organism (or a
+particular anatomical location), is increased or decreased. While a
+flow rate of a single {\em cytoplasmic streaming} process is a quality
+of that process, an increased {\em total} cytoplasmic flow rate is a
+quality of the regulation of {\em cytoplasmic streaming}. In
+particular, it is possible for an organism to have a normal --- or
+even a decreased --- cytoplasmic flow rate in each individual
+cytoplasmic streaming process while at the same time having an
+increased total cytoplasmic flow rate due to a large increase in the
+frequency of occurrence of cytoplasmic streaming processes. Similarly,
+the frequency of occurrence of cytoplasmic streaming may be normal or
+decreased while the total cytoplasmic flow rate is increased due to an
+increased cytoplasmic flow rate in each individual cytoplasmic
+streaming process. Table \ref{tbl:flow} illustrates the dependencies
+between rates of individual processes, their frequency of occurrence
+and the total rate of these processes. We include total rates as
+regulatory abnormalities in the CPO since these are the attributes of
+processes that are often measured or observed, while the rates of
+individual processes are inferred following a schema such as Table
+\ref{tbl:flow}.
 
@@
 given time period is increased in comparison to a reference process.''
+
+When using the GO and PATO ontologies from 30 Nov 2011, CPO contains
+XX classes...
+Classifies in...
 

### 2011-11-25 paper/drinking.tex

@@
 
-\title{Process-based phenotypes: an application to the Cellular
-  Phenotype Ontology}
+\title{Process-based phenotypes: an application to the ontology of
+  cell physiology}
 
@@
 various ontologies that describe the entities that have been affected,
-such as the various anatomical ontologies, GO\cite {Ashburner2000b},
-the Cell Type Ontology\cite {Bard2005a}, (SO)\cite {Eilbeck2005} etc
+such as the various anatomical ontologies, GO \cite {Ashburner2000b},
+the Cell Type Ontology \cite {Bard2005a}, SO \cite {Eilbeck2005} etc
 with the various qualities it provides for defining how these entities
@@
 
+When using the GO and PATO ontologies from 30 Nov 2011, CPO contains
+XX classes...
+Classifies in...
+
 \section{Discussion}

### 2011-11-25 paper/drinking.tex
added FYPO blurb

@@
 \subsection{Processes, their participants and qualities}
+\begin{figure*}
+  \centering
+  \includegraphics[width=.9\textwidth]{onto.pdf}
+  \caption{Overview over ...\label{fig:onto}}
+\end{figure*}
+
 Processes are entities that are temporally extended: processes have
@@
 in order to generate an OWL representation of the CPO. The software
-requires three input files: a version of the GO on which to based the
+requires three input files: a version of the GO on which to base the
 generated CPO, a version of PATO that is used to define abnormal
@@
 
+The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
+to support annotation of phenotypes in {\em Schizosaccharomyces
+  pombe}, consists of pre-composed terms describing normal or abnormal
+cellular phenotypes. Over 80\% of FYPO definitions reference
+descendants of GO BP 'cellular process' as the entity; a further 11\%
+reference GO CC terms. All FYPO explicit definitions reference
+qualities in PATO, including {\em normal}, {\em abnormal}, and several
+process qualities including {\em increased duration}, {\em decreased
+  occurrence}, etc. FYPO will thus fit neatly under the CPO umbrella,
+and stands to benefit from the automated synchronization between CPO
+and GO. {\em S. pombe} annotations to FYPO terms will provide a rich
+body of highly specific, well-supported data to be integrated with
+data from other species.
+
 \subsection{Future research: acts and intentional states}
@@
 \section{Acknowledgements}
+MAH is supported by Wellcome Trust grant WT090548MA.
 

### 2011-11-25 trunk/paper/drinking.tex

@@
 \subsection{Processes, their participants and qualities}
-Processes are entities that are temporally extended: processes have
-temporal parts and {\em unfold} through time. Examples of classes that
-may have processes as instances include {\em Drinking}, {\em
-  Triathlon} or {\em Apoptosis}.
+\begin{figure}
+  \centering
+  \includegraphics[width=.45\textwidth]{onto.pdf}
+  \caption{Overview over ...\label{fig:onto}}
+\end{figure}
+
+Most biomedical ontologies share common distinctions between different
+kinds of entities: physical objects, qualities, functions and
+processes. A physical object is an entity that is present as a whole
+at a time point, i.e., an entity that has no temporal parts. A quality
+is an attribute or feature of an entity. Physical objects together
+with their qualities give rise to functions, which are capabilities or
+potentials of physical objects. These functions can then be realized
+by processes, which are temporally extended entities. Examples of
+classes that may have processes as instances include {\em Drinking},
+{\em Triathlon} or {\em Apoptosis}.  Figure \ref{fig:onto} illustrates
+these basic categories of being.
 

### 2011-11-25 paper/drinking.tex

@@
 \subsection{Processes, their participants and qualities}
-\begin{figure*}
+\begin{figure}
   \centering
-  \includegraphics[width=.9\textwidth]{onto.pdf}
+  \includegraphics[width=.45\textwidth]{onto.pdf}
   \caption{Overview over ...\label{fig:onto}}
-\end{figure*}
+\end{figure}
 
-Processes are entities that are temporally extended: processes have
-temporal parts and {\em unfold} through time. Examples of classes that
-may have processes as instances include {\em Drinking}, {\em
-  Triathlon} or {\em Apoptosis}.
+Most biomedical ontologies share common distinctions between different
+kinds of entities: physical objects, qualities, functions and
+processes. A physical object is an entity that is present as a whole
+at a time point, i.e., an entity that has no temporal parts. A quality
+is an attribute or feature of an entity. Physical objects together
+with their qualities give rise to functions, which are capabilities or
+potentials of physical objects. These functions can then be realized
+by processes, which are temporally extended entities. Examples of
+classes that may have processes as instances include {\em Drinking},
+{\em Triathlon} or {\em Apoptosis}.  Figure \ref{fig:onto} illustrates
+these basic categories of being.
 
@@
 \subsection{Attributes of processes}
-An attribute that can be directly linked to a process is its {\em
-  duration}. A duration is an attribute that characterizes the
-temporal extent of a process and is similar to {\em Length}, {\em
-  Area} and {\em Volume} for one-, two- and three-dimensional
-entities. Furthermore, processes may be attributed a {\em temporal
-  location} with respect to a particular coordinate system.
+We develop a model of process attributes that is applicable for
+representations of physiology and related phenotypes. In principle, we
+distinguish between three different kinds of process attributes:
+process attributes that are linked directly to processes, attributes
+that arise due to process parts and attributes of processes, and
+attributes arising from attributes of process participants and other
+process attributes.
+
+Attribute that can be directly linked to a process are the result of
+the temporal extension of a process and include process {\em duration}
+and {\em temporal location}. A duration is an attribute that
+characterizes the temporal extent of a process and is similar to {\em
+  Length}, {\em Area} and {\em Volume} for one-, two- and
+three-dimensional entities. A {\em Temporal location} positions the
+time interval during which the process occurs with respect to a
+reference coordinate system.
 
@@
 \subsection{Cellular Phenotype Ontology}
-A domain that would greatly benefit from the development of a CPO is
-Systems Microscopy, which aims to understand complex and dynamic
-cellular systems by combining automated fluorescence microscopy, cell
-microarray platforms, quantitative image analysis and data mining
-(Lock \& Stromblad, Exp. Cell Res. 2010).
-
-If we consider some of the studies, which have been published in this
-field in the last few years (Neumann et al, Nature 464: 722-727;
-Schmitz et al, Nature Cell Biology 12: 886-893; Fuchs et al, Mol Sys
-Biology 6: 370), the need for a CPO becomes evident.
-
-In all three studies here considered live-cell imaging assays and RNAi
-knockdown were used to generate phenotypic profiles that quantify the
-cellular response to a given siRNA thus allowing identification of
-hundreds of genes involved in diverse biological functions including
-cell division, migration and survival.
-
-In each study, several phenotypes were detected and described by the
-authors without the use of ontologies (see table), making the
-integration between datasets extremely difficult. For example, it is
-evident that cell division phenotypes were observed in the three
-datasets (e.g. 'Mitotic delay/arrest' and 'Prolonged mitotic exit' or
-'Methaphase delay' and 'Methaphase cells') but the overlap between
-such phenotypes is unclear.
-
-Data integration is also complicated by the lack of standardization at
-the level of data production and processing; all these issues are
-currently being address by the different groups involved in the
-Systems Microscopy Network of Excellence
-(http://www.systemsmicroscopy.eu/) and the first step towards data
-integration will be achieved by developing a CPO.
-
-This ontology will be used to integrate phenotypes' definitions across
-existing datasets and will then become an integrated part of the data
-processing pipeline and used to annotate the data as it gets generated
-(Conrad et al, Nature Methods 8: 246-249).
-
 \begin{figure*}[h]
@@
 \subsection{Use of CPO}
-
 The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
@@
 
-\subsection{Future research: acts and intentional states}
-activities/actions, intentional (goal-oriented) acts
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+A domain that would greatly benefit from the development of a CPO is
+Systems Microscopy, which aims to understand complex and dynamic
+cellular systems by combining automated fluorescence microscopy, cell
+microarray platforms, quantitative image analysis and data mining
+(Lock \& Stromblad, Exp. Cell Res. 2010).
 
-feeling thirsty (internal state) as driving force behind drinking
-behavior
+If we consider some of the studies, which have been published in this
+field in the last few years (Neumann et al, Nature 464: 722-727;
+Schmitz et al, Nature Cell Biology 12: 886-893; Fuchs et al, Mol Sys
+Biology 6: 370), the need for a CPO becomes evident.
 
-non-measurable, inference only
+In all three studies here considered live-cell imaging assays and RNAi
+knockdown were used to generate phenotypic profiles that quantify the
+cellular response to a given siRNA thus allowing identification of
+hundreds of genes involved in diverse biological functions including
+cell division, migration and survival.
+
+In each study, several phenotypes were detected and described by the
+authors without the use of ontologies (see table), making the
+integration between datasets extremely difficult. For example, it is
+evident that cell division phenotypes were observed in the three
+datasets (e.g. 'Mitotic delay/arrest' and 'Prolonged mitotic exit' or
+'Methaphase delay' and 'Methaphase cells') but the overlap between
+such phenotypes is unclear.
+
+Data integration is also complicated by the lack of standardization at
+the level of data production and processing; all these issues are
+currently being address by the different groups involved in the
+Systems Microscopy Network of Excellence
+(http://www.systemsmicroscopy.eu/) and the first step towards data
+integration will be achieved by developing a CPO.
+
+This ontology will be used to integrate phenotypes' definitions across
+existing datasets and will then become an integrated part of the data
+processing pipeline and used to annotate the data as it gets generated
+(Conrad et al, Nature Methods 8: 246-249).
+
+% \subsection{Future research: acts and intentional states}
+% activities/actions, intentional (goal-oriented) acts
+
+% feeling thirsty (internal state) as driving force behind drinking
+% behavior
+
+% non-measurable, inference only
 

### 2011-11-25 trunk/paper/drinking.tex

@@
 
-\subsection{Processes, their participants and qualities}
+\subsection{Processes and their participants}
 \begin{figure}
@@
 \subsection{Attributes of processes}
-An attribute that can be directly linked to a process is its {\em
-  duration}. A duration is an attribute that characterizes the
-temporal extent of a process and is similar to {\em Length}, {\em
-  Area} and {\em Volume} for one-, two- and three-dimensional
-entities. Furthermore, processes may be attributed a {\em temporal
-  location} with respect to a particular coordinate system.
+We develop a model of process attributes that is applicable for
+representations of physiology and related phenotypes. In principle, we
+distinguish between three different kinds of process attributes:
+the first are process attributes that arise directly from processes
+and include {\em duration} and {\em temporal location}; the second are
+attributes that arise from processes and their parts and include {\em
+  frequency}; and the third are attributes that arise from processes
+and qualities of their participants, and include {\em flow rates}.
+
+Attribute that can be directly linked to a process are the result of
+the temporal extension of a process and include process {\em duration}
+and {\em temporal location}. A duration is an attribute that
+characterizes the temporal extent of a process and is similar to {\em
+  Length}, {\em Area} and {\em Volume} for one-, two- and
+three-dimensional entities. A {\em Temporal location} positions the
+time interval during which the process occurs with respect to a
+reference coordinate system.
 
@@
 \section{Discussion}
-\subsection{Implementation as pre-composed ontology}
-Pre-composed or software support for pattern?
-
 \subsection{Use of CPO}
-
 The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
@@
 
-\subsection{Future research: acts and intentional states}
-activities/actions, intentional (goal-oriented) acts
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+A domain that would greatly benefit from the development of a CPO is
+Systems Microscopy, which aims to understand complex and dynamic
+cellular systems by combining automated fluorescence microscopy, cell
+microarray platforms, quantitative image analysis and data mining
+(Lock \& Stromblad, Exp. Cell Res. 2010).
 
-feeling thirsty (internal state) as driving force behind drinking
-behavior
+If we consider some of the studies, which have been published in this
+field in the last few years (Neumann et al, Nature 464: 722-727;
+Schmitz et al, Nature Cell Biology 12: 886-893; Fuchs et al, Mol Sys
+Biology 6: 370), the need for a CPO becomes evident.
 
-non-measurable, inference only
+In all three studies here considered live-cell imaging assays and RNAi
+knockdown were used to generate phenotypic profiles that quantify the
+cellular response to a given siRNA thus allowing identification of
+hundreds of genes involved in diverse biological functions including
+cell division, migration and survival.
+
+In each study, several phenotypes were detected and described by the
+authors without the use of ontologies (see table), making the
+integration between datasets extremely difficult. For example, it is
+evident that cell division phenotypes were observed in the three
+datasets (e.g. 'Mitotic delay/arrest' and 'Prolonged mitotic exit' or
+'Methaphase delay' and 'Methaphase cells') but the overlap between
+such phenotypes is unclear.
+
+Data integration is also complicated by the lack of standardization at
+the level of data production and processing; all these issues are
+currently being address by the different groups involved in the
+Systems Microscopy Network of Excellence
+(http://www.systemsmicroscopy.eu/) and the first step towards data
+integration will be achieved by developing a CPO.
+
+This ontology will be used to integrate phenotypes' definitions across
+existing datasets and will then become an integrated part of the data
+processing pipeline and used to annotate the data as it gets generated
+(Conrad et al, Nature Methods 8: 246-249).
+
+\subsection{Pre- vs. post-composed ontology}
+Pre-composed or software support for pattern?
 

### 2011-11-25 trunk/paper/drinking.tex

@@
 
-Attribute that can be directly linked to a process are the result of
-the temporal extension of a process and include process {\em duration}
-and {\em temporal location}. A duration is an attribute that
-characterizes the temporal extent of a process and is similar to {\em
-  Length}, {\em Area} and {\em Volume} for one-, two- and
-three-dimensional entities. A {\em Temporal location} positions the
-time interval during which the process occurs with respect to a
-reference coordinate system.
+Attribute that can be directly linked to a process arise from
+processes' temporal extension. For example, a duration is an attribute
+that characterizes the temporal extent of a process and is similar to
+{\em Length}, {\em Area} and {\em Volume} for one-, two- and
+three-dimensional physical objects. A {\em Temporal location}
+positions the time interval at which the process occurs with respect
+to a reference coordinate system.
 
 However, the majority of attributes that characterize processes are
-not based on genuine process attributes alone, but rather relate
-attributes of process participants with the duration of a process. In
-particular, a {\em rate} typically refers to an attribute of some
-entity {\em with respect to an attribute of another entity}, and in
-the context of processes, rates often refer to attributes of a process
-participant with respect to the duration of the process. For example,
-a {\em mass flow rate} refers to the {\em Mass} of a process
+not based on these types of process attributes alone, but rather
+relate attributes of process participants with the duration of a
+process. In particular, a {\em rate} typically refers to an attribute
+of some entity {\em with respect to an attribute of another entity},
+and in the context of processes, rates often refer to attributes of a
+process participant with respect to the duration of the process. For
+example, a {\em mass flow rate} refers to the {\em Mass} of a process
 participant with respect to the duration of the process, i.e., how
@@
 
-Not all rates of a process depend on attributes of the process
-participants. In particular, a {\em frequency of occurrence} or {\em
-  event rate} refers to the number of occurrences of a type of process
-during the reference process. For example, a {\em rate of heart
+However, not all rates of a process depend on attributes of the
+process participants. In particular, a {\em frequency of occurrence}
+or {\em event rate} refers to the number of occurrences of a type of
+process during a reference process. For example, a {\em rate of heart
   beating} refers to the number of {\em Heart beating} processes that
-occur within a reference process (e.g., a process containing the heart
-with a duration of one minute). Further attributes that depend on
-types of processes with regard to a reference process are {\em
-  distribution patterns}, i.e., how the occurrences of processes of a
-particular type are distributed within a reference process. For
+occur within a reference process (e.g., a process in which the heart
+participates with a duration of one minute). Further attributes that
+depend on types of processes with regard to a reference process are
+{\em distribution patterns}, i.e., how the occurrences of processes of
+a particular type are distributed within a reference process. For
 example, the heart may beat {\em rhythmically} or {\em arrhythmically}
@@
 processes. For example, the rate of heart beating may change ({\em
-  increase} or {\em decrease}) throughout the course of the reference
+  increase} or {\em decrease}) throughout the course of a reference
 process. A simple analysis of {\em increasing} ({\em decreasing})
 rates would be that the rate of a heart beating within the first half
-of a process if {\em lower} ({\em higher}) than in the second half of
-the process. Mathematically, this process sub-division can be iterated
-until we end up with infinitesimally small time intervals.
+of a process is {\em lower} ({\em higher}) than in the second half of
+the process. To make such an assertion, we divide a process into two
+temporal parts. Mathematically, this process of sub-division can be
+iterated until processes occur within infinitesimally small time
+intervals.
 
@@
 all temporal parts of $p$ are also instances of $C$. Examples include
-{\em continuous movements} or {\em regulation} processes. On the other
+{\em continuous movements} or {\em mass flow} processes. On the other
 hand, {\em discrete processes} can be subdivided into stages of
@@
 
-% In line with the GO, we call processes that cause
+We may further attribute a {\em frequency} or {\em rate} to an object
+instead of a process. For example, a heart that beats {\em now} with a
+frequency of 80 beats per minute, or a car that is moving at a speed
+of 180 kilometres per hour {\em at a particular point in time} (e.g.,
+as observed with a speed camera) can be considered attributes of the
+objects (the heart or the car), not attributes of the processes in
+which the objects participate. However, these are {\em different}
+kinds of attributes. Rates, when considered as attributes of objects,
+may be explicitly defined using rates of processes. For example, the
+heart beating frequency of a particular heart $h$ at a time point $t$
+is the frequency of a reference heart beating process in which $h$
+participates. Such a reference process is necessary in order to obtain
+a value for a frequency even when no {\em heart beating} process is
+occurring. However, the frequency is only an attribute of the heart in
+virtue of such a reference process in which {\em heart beating} is
+actually occurring.  This reference process can be uniquely determined
+for continuous processes (such that the frequency or rate of an object
+at a time $t$ is the frequency or rate of the infinitesimally small
+process around $t$), and is ambiguous for discrete processes (in which
+case the reference process must be explicitly stated).
 
-% [Regulation of X vs. X: regulation modulates the number/amount of
-% occurrence...]
-
-We may say that an object has a {\em frequency} or {\em rate} as if
-these were attributes of objects at time points. For example, that my
-heart {\em now} beats with a frequency of 80 beats per minute, or that
-a car is moving with a speed of 280 kilometres per minute {\em at a
-  particular point in time} (e.g., as observed with a speed camera)
-may be considered attributes of the objects, not attributes of
-processes. In our analysis, rates as attributes of objects can be
-explicitly defined using rates of processes. For example, the heart
-beating frequency of a particular heart $h$ at a time point $t$ is the
-frequency of heart beating processes in which $h$ participates at an
-infinitesimally small process around $t$ and within a reference
-process. The reference process is necessary in order to obtain a value
-for a frequency or rate even if the process in question is not
-occurring. For example, a rate of cardiac muscle contraction can be
-determined even when no cardiac muscle contraction is occurring in
-virtue of a reference process with a rate of cardiac muscle
-contraction.
-
-The reference process for an object at a time point $t$ is uniquely
-determined for continuous processes (the infinitesimally small process
-around $t$), and ambiguous for discrete processes.
-
-A similar construction as for continuous processes can be made for
-discrete processes. A rate of blood flow is immediately an attribute
-of a single heart beat process. However, during a period of time, the
-cumulative rate of blood flow can be observed as the sum of the blood
-flow rates of individual heart beating processes.
+% A similar construction as for continuous processes can be made for
+% discrete processes. A rate of blood flow is immediately an attribute
+% of a single heart beat process. However, during a period of time, the
+% cumulative rate of blood flow can be observed as the sum of the blood
+% flow rates of individual heart beating processes.
 
@@
 
-Complex relations between Based on these definition patterns,
-interdependencies between process attributes can be observed.
+% Complex relations between Based on these definition patterns,
+% interdependencies between process attributes can be observed.
 

### 2011-11-26 trunk/paper/drinking.tex

@@
 \section{Introduction}
+Phenotype studies on all scales and levels of granularity are now an
+invaluable tool for functional genomics research. Phenotypes of
+targeted mutations in animal models are now systematically recorded to
+reveal the role of individual genes within a biological system. These
+phenotype studies now play a key role in translational research and
+are being used to reveal candidate genes for orphan diseases and to
+identify chemicals that may have effects on these diseases \cite{}.
 
-problem of phenotypes...
+The large volume and diversity of phenotypes within different species
+and across multiple scales and levels of granularity necessitates the
+application of flexible strategies for managing and integrating data
+so that it becomes amenable to automated comparative analyses. To
+integrate biomedical data across heterogeneous information systems,
+biomedical ontologies were being developed \cite{Smith2007}. An
+ontology is an explicit specification of a conceptualization of a
+domain and can be used to make the meaning of terms in a vocabulary
+explicit \cite{Gruber1995, Guarino1998}. They play a crucial role in
+the annotation of biomedical data and the integration of model
+organism databases \cite{go2010, Bada2004, goble}.
 
-why reasoning, why formal, success,...
+Ontologies increasingly rely on the use of Semantic Web technologies
+\cite{Berners-Lee2001}. The Semantic Web provides a stack of protocols
+and languages to include explicit semantics in websites. In
+particular, the Web Ontology Language (OWL) \cite{Grau2008} has been
+designed to express and share ontologies within the Semantic Web. OWL
+is a language based on description logics (a group of formal languages
+based on first-order predicate logic). Automated reasoners have been
+developed within the Semantic Web to perform complex operations on
+ontologies formulated in OWL. In particular, automated reasoners can
+verify an ontology's consistency and use deductive inference to
+perform powerful queries over ontologies. To benefit from automated
+reasoning and the rapidly increasing number of software tools that are
+being developed within the Semantic Web, most biomedical ontologies
+are now available in OWL or can be converted into an OWL-based
+representation \cite{Horrocks2007, Hoehndorf2010patterns}.
 
-solved: structures, qualities
+In the domain of phenotypes, multiple ontologies have been
+developed. In particular, ontologies to characterize mammalian
+\cite{Smith2004}, human \cite{Robinson2008}, yeast \cite{ypo} and worm
+\cite{wpo} phenotypes are now available, with several more ontologies
+being under development.
+
+structures, qualities
 
@@
 physiology...
-measurements...
+measurements and inferences...
 
@@
 
+Figure \ref{fig:patterns} illustrates some examples of
+non-comparative and comparative process attributes.
+
 % A similar construction as for continuous processes can be made for
@@
 
-CPO contains both structural abnormalities and physiological
-abnormalities.
+While our considerations about process attributes are only the
+beginnings of a full-fledged theory, we have derived several phenotype
+formalization patterns and a high-level taxonomic structure of
+process-based phenotypes. To evaluate our approach, we created the
+Cellular Phenotype Ontology (CPO) by automatically applying our
+patterns to the GO.
 
+% CPO contains both structural abnormalities and physiological
+% abnormalities of cells.
 Phenotypes in the CPO are either based on structural abnormalities or

### 2011-11-26 trunk/paper/drinking.tex

@@
 \cite{Smith2004}, human \cite{Robinson2008}, yeast \cite{ypo} and worm
-\cite{wpo} phenotypes are now available, with several more ontologies
-being under development.
+\cite{wpo} phenotypes are now available, while several more phenotype
+ontologies are under development. To benefit from automated reasoning,
+integrate phenotypes across species and reuse the content of anatomy
+and process ontologies, classes in phenotype ontologies were defined
+using the framework of the Phenotypic Attribute and Trait (PATO)
+ontology \cite{Gkoutos2005}. According to the PATO framework, a
+phenotype can be decomposed, using an Entity-Quality model, into an
+affected entity and a quality that characterizes {\em how} the entity
+is affected \cite{Gkoutos2005}. Such decompositions have been created
+for several widely used phenotype ontologies \cite{Mungall2010,
+  Gkoutos2009b, obml2011h1}, and are being applied together with
+methods for reusing knowledge contained in anatomy ontologies
+\cite{Mungall2010, Hoehndorf2010phene}.
 
-structures, qualities
+While the PATO framework is now successfully being applied to the
+integration of phenotypes in which
 

### 2011-11-26 trunk/paper/drinking.tex

@@
 
-\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Gabriella
-  Rustici$^2$ and Midori A. Harris$^3$ and Georgios V. Gkoutos$^{1}$}
+\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Midori
+  A. Harris$^3$ and ... and Gabriella Rustici$^2$ and Georgios
+  V. Gkoutos$^{1}$}
 
@@
 
-While the PATO framework is now successfully being applied to the
-integration of phenotypes in which
+While the PATO framework is now successfully being applied to
+semantically integrate phenotypes across species, the diversity and
+complexity of phenotypes in which biological processes and functions
+are impaired continues to limit the interoperability between phenotype
+ontologies. Biological processes and functions are linked to the
+physiology which the components of biological systems exhibit and
 
@@
 potentials of physical objects. These functions can then be realized
-by processes, which are temporally extended entities. Examples of
-classes that may have processes as instances include {\em Drinking},
-{\em Triathlon} or {\em Apoptosis}.  Figure \ref{fig:onto} illustrates
-these basic categories of being.
+by processes, which are temporally extended entities
+\cite{Burek2006}. Examples of classes that may have processes as
+instances include {\em Drinking}, {\em Triathlon} or {\em Apoptosis}.
+Figure \ref{fig:onto} illustrates these basic categories of being.
 
@@
 processes are entities that are present at time points (i.e., they
-have no temporal parts) and may undergo changes during the process.
-We can further distinguish different {\em roles} of participation in a
-process. For example, a runner may participate in a {\em Triathlon}
-process as a {\em Runner (role)}, while another person can participate
-in the same process but in a different role (such as {\em Referee}).
+have no temporal parts) and may undergo changes during the process
+\cite{Herre2006, Herre2010}.  We can further distinguish different
+{\em roles} of participation in a process \cite{Loebe2007}. For
+example, a runner may participate in a {\em Triathlon} process as a
+{\em Runner (role)}, while another person can participate in the same
+process but in a different role (such as {\em Referee}).
 

### 2011-11-26 trunk/paper/drinking.tex

@@
 \author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Midori
-  A. Harris$^3$ and ... and Gabriella Rustici$^2$ and Georgios
-  V. Gkoutos$^{1}$}
+  A. Harris$^3$ and Heinrich Herre(?) and Gabriella Rustici$^2$ and
+  Georgios V. Gkoutos$^{1}$}
 
@@
 are impaired continues to limit the interoperability between phenotype
-ontologies. Biological processes and functions are linked to the
-physiology which the components of biological systems exhibit and
+ontologies. Major challenges for representing and integrating
+process-based phenotypes include establishing the link to the
+components of biological systems that have the capabilities to exhibit
+such a behaviour, and that attributes of processes are often measured
+{\em indirectly} and inferred from other attributes.
 
-open: process phenotypes very complex...  
-physiology...
-measurements and inferences...
+To illustrate these challenges, consider physiological processes of
+the heart. One of the heart's functions is {\em Heart beating}, i.e.,
+a capability that is realized through processes of the type {\em Heart
+  beating}. {\em Blood} is a participant of {\em Heart beating}
+processes.  An abnormal phenotype of an organism could be that the
+{\em rate of heart beating} is increased with the intended meaning
+that the number of {\em Heart beating} processes in a given time
+interval is higher than normal. An attribute such as {\em increased
+  rate of heart beating}
 
-\subsection{Problem statement 1}
 Heart rate increased: frequency of CMC increased fluid flow increased
@@
 
-\subsection{Problem statement 2}
-increased apoptosis
-
-\subsection{Observations}
+%%%%%%%%%%%%%%%%%%
 observation 1: there is a fundamental distinction between attribution

### 2011-11-27 trunk/paper/drinking.tex

@@
 processes.  An abnormal phenotype of an organism could be that the
-{\em rate of heart beating} is increased with the intended meaning
-that the number of {\em Heart beating} processes in a given time
-interval is higher than normal. An attribute such as {\em increased
-  rate of heart beating}
+{\em rate of heart beating} is increased. The intended meaning of such
+a description is that the number of {\em Heart beating} processes in a
+given time interval is higher than normal. Another important attribute
+of heart physiology is the fluid flow rate through the heart. For an
+abnormal phenotype such as {\em increased rate of fluid flow through
+  the heart}, the intended meaning could either be that the amount of
+fluid that is moved through the heart within a single heart beating
+process is increased or that the amount of fluid that is moved through
+the heart within a period of time interval is increased.
 
-Heart rate increased: frequency of CMC increased fluid flow increased
-volume per contraction increased ...
+Based on these examples, we can make several observations about
+process attributes. First, for a process like {\em Heart beating}, we
+can distinguish between single occurrences and processes in which {\em
+  Heart beating} occurs multiple times. Only the latter kind of
+process may have a {\em rate of heart beating}, while {\em Heart
+  beating} processes do not have such an attribute. Second, we can
+distinguish between abnormal fluid flow rates in {\em Heart beating}
+processes and rate of fluid flow through the heart within a given
+duration. Both may have entirely different underlying causes and it is
+therefore important to distinguish between them. Finally, we may be
+able to infer some phenotypes from others, thereby limiting the number
+of phenotypes that must be experimentally observed. For example, when
+the fluid flow rate in single heart beating processes is increased and
+the {\em rate of heart beating} is increased within a process $P$,
+then the rate of fluid flow through the heart will be increased for
+$P$.
 
-what is the relation to the heart?
-
-%%%%%%%%%%%%%%%%%%
-observation 1: there is a fundamental distinction between attribution
-of qualities for a fixed period (e.g., 24h) and single {\em act}s of
-drinking.
-
-observation 2: each of the different possible ``increased'' drinking
-behavior may be the result of different underlying causes and due to
-this, distinguishing between them may be important.
-
-observation 3: depending on the experiment performed, only some of the
-actual qualities can be measured.
-
-observation 4: some qualities can be inferred given others. For
-example,...
+Here, we present the foundations for an ontology of physiological
+phenotypes.
 

### 2011-11-27 paper/drinking.tex

@@
 
-\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Gabriella
-  Rustici$^2$ and Midori A. Harris$^3$ and Georgios V. Gkoutos$^{1}$}
+\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Midori
+  A. Harris$^3$ and Heinrich Herre(?) and Gabriella Rustici$^2$ and
+  Georgios V. Gkoutos$^{1}$}
 
@@
 \section{Introduction}
+Phenotype studies on all scales and levels of granularity are now an
+invaluable tool for functional genomics research. Phenotypes of
+targeted mutations in animal models are now systematically recorded to
+reveal the role of individual genes within a biological system. These
+phenotype studies now play a key role in translational research and
+are being used to reveal candidate genes for orphan diseases and to
+identify chemicals that may have effects on these diseases \cite{}.
 
-problem of phenotypes...
+The large volume and diversity of phenotypes within different species
+and across multiple scales and levels of granularity necessitates the
+application of flexible strategies for managing and integrating data
+so that it becomes amenable to automated comparative analyses. To
+integrate biomedical data across heterogeneous information systems,
+biomedical ontologies were being developed \cite{Smith2007}. An
+ontology is an explicit specification of a conceptualization of a
+domain and can be used to make the meaning of terms in a vocabulary
+explicit \cite{Gruber1995, Guarino1998}. They play a crucial role in
+the annotation of biomedical data and the integration of model
+organism databases \cite{go2010, Bada2004, goble}.
 
-why reasoning, why formal, success,...
+Ontologies increasingly rely on the use of Semantic Web technologies
+\cite{Berners-Lee2001}. The Semantic Web provides a stack of protocols
+and languages to include explicit semantics in websites. In
+particular, the Web Ontology Language (OWL) \cite{Grau2008} has been
+designed to express and share ontologies within the Semantic Web. OWL
+is a language based on description logics (a group of formal languages
+based on first-order predicate logic). Automated reasoners have been
+developed within the Semantic Web to perform complex operations on
+ontologies formulated in OWL. In particular, automated reasoners can
+verify an ontology's consistency and use deductive inference to
+perform powerful queries over ontologies. To benefit from automated
+reasoning and the rapidly increasing number of software tools that are
+being developed within the Semantic Web, most biomedical ontologies
+are now available in OWL or can be converted into an OWL-based
+representation \cite{Horrocks2007, Hoehndorf2010patterns}.
 
-solved: structures, qualities
+In the domain of phenotypes, multiple ontologies have been
+developed. In particular, ontologies to characterize mammalian
+\cite{Smith2004}, human \cite{Robinson2008}, yeast \cite{ypo} and worm
+\cite{wpo} phenotypes are now available, while several more phenotype
+ontologies are under development. To benefit from automated reasoning,
+integrate phenotypes across species and reuse the content of anatomy
+and process ontologies, classes in phenotype ontologies were defined
+using the framework of the Phenotypic Attribute and Trait (PATO)
+ontology \cite{Gkoutos2005}. According to the PATO framework, a
+phenotype can be decomposed, using an Entity-Quality model, into an
+affected entity and a quality that characterizes {\em how} the entity
+is affected \cite{Gkoutos2005}. Such decompositions have been created
+for several widely used phenotype ontologies \cite{Mungall2010,
+  Gkoutos2009b, obml2011h1}, and are being applied together with
+methods for reusing knowledge contained in anatomy ontologies
+\cite{Mungall2010, Hoehndorf2010phene}.
 
-open: process phenotypes very complex...  
-physiology...
-measurements...
+While the PATO framework is now successfully being applied to
+semantically integrate phenotypes across species, the diversity and
+complexity of phenotypes in which biological processes and functions
+are impaired continues to limit the interoperability between phenotype
+ontologies. Major challenges for representing and integrating
+process-based phenotypes include establishing the link to the
+components of biological systems that have the capabilities to exhibit
+such a behaviour, and that attributes of processes are often measured
+{\em indirectly} and inferred from other attributes.
 
-\subsection{Problem statement 1}
-Heart rate increased: frequency of CMC increased fluid flow increased
-volume per contraction increased ...
+To illustrate these challenges, consider physiological processes of
+the heart. One of the heart's functions is {\em Heart beating}, i.e.,
+a capability that is realized through processes of the type {\em Heart
+  beating}. {\em Blood} is a participant of {\em Heart beating}
+processes.  An abnormal phenotype of an organism could be that the
+{\em rate of heart beating} is increased. The intended meaning of such
+a description is that the number of {\em Heart beating} processes in a
+given time interval is higher than normal. Another important attribute
+of heart physiology is the fluid flow rate through the heart. For an
+abnormal phenotype such as {\em increased rate of fluid flow through
+  the heart}, the intended meaning could either be that the amount of
+fluid that is moved through the heart within a single heart beating
+process is increased or that the amount of fluid that is moved through
+the heart within a period of time interval is increased.
 
-what is the relation to the heart?
+Based on these examples, we can make several observations about
+process attributes. First, for a process like {\em Heart beating}, we
+can distinguish between single occurrences and processes in which {\em
+  Heart beating} occurs multiple times. Only the latter kind of
+process may have a {\em rate of heart beating}, while {\em Heart
+  beating} processes do not have such an attribute. Second, we can
+distinguish between abnormal fluid flow rates in {\em Heart beating}
+processes and rate of fluid flow through the heart within a given
+duration. Both may have entirely different underlying causes and it is
+therefore important to distinguish between them. Finally, we may be
+able to infer some phenotypes from others, thereby limiting the number
+of phenotypes that must be experimentally observed. For example, when
+the fluid flow rate in single heart beating processes is increased and
+the {\em rate of heart beating} is increased within a process $P$,
+then the rate of fluid flow through the heart will be increased for
+$P$.
 
-\subsection{Problem statement 2}
-increased apoptosis
-
-\subsection{Observations}
-observation 1: there is a fundamental distinction between attribution
-of qualities for a fixed period (e.g., 24h) and single {\em act}s of
-drinking.
-
-observation 2: each of the different possible ``increased'' drinking
-behavior may be the result of different underlying causes and due to
-this, distinguishing between them may be important.
-
-observation 3: depending on the experiment performed, only some of the
-actual qualities can be measured.
-
-observation 4: some qualities can be inferred given others. For
-example,...
+Here, we present the foundations for an ontology of process
+phenotypes. We present the outline of a theory in which several kinds
+of process attributes can be distinguished so that normal and abnormal
+physiology of biological systems can be formally characterized. We
+apply this theory to cellular processes and create the Cell Phenotype
+Ontology (CPO). The CPO is linked to reference ontologies for
+qualities, biological processes, functions and cell components, and
+its prime application is the unification of phenotypes on the cellular
+level across different species as well as for annotation of cellular
+phenotypes in domains in which no such ontology exists.
 
@@
 distinguish between three different kinds of process attributes:
-process attributes that are linked directly to processes, attributes
-that arise due to process parts and attributes of processes, and
-attributes arising from attributes of process participants and other
-process attributes.
+the first are process attributes that arise directly from processes
+and include {\em duration} and {\em temporal location}; the second are
+attributes that arise from processes and their parts and include {\em
+  frequency}; and the third are attributes that arise from processes
+and qualities of their participants, and include {\em flow rates}.
 
-Attribute that can be directly linked to a process are the result of
-the temporal extension of a process and include process {\em duration}
-and {\em temporal location}. A duration is an attribute that
-characterizes the temporal extent of a process and is similar to {\em
-  Length}, {\em Area} and {\em Volume} for one-, two- and
-three-dimensional entities. A {\em Temporal location} positions the
-time interval during which the process occurs with respect to a
-reference coordinate system.
+Attribute that can be directly linked to a process arise from
+processes' temporal extension. For example, a duration is an attribute
+that characterizes the temporal extent of a process and is similar to
+{\em Length}, {\em Area} and {\em Volume} for one-, two- and
+three-dimensional physical objects. A {\em Temporal location}
+positions the time interval at which the process occurs with respect
+to a reference coordinate system.
 
 However, the majority of attributes that characterize processes are
-not based on genuine process attributes alone, but rather relate
-attributes of process participants with the duration of a process. In
-particular, a {\em rate} typically refers to an attribute of some
-entity {\em with respect to an attribute of another entity}, and in
-the context of processes, rates often refer to attributes of a process
-participant with respect to the duration of the process. For example,
-a {\em mass flow rate} refers to the {\em Mass} of a process
+not based on these types of process attributes alone, but rather
+relate attributes of process participants with the duration of a
+process. In particular, a {\em rate} typically refers to an attribute
+of some entity {\em with respect to an attribute of another entity},
+and in the context of processes, rates often refer to attributes of a
+process participant with respect to the duration of the process. For
+example, a {\em mass flow rate} refers to the {\em Mass} of a process
 participant with respect to the duration of the process, i.e., how
@@
 
-% In line with the GO, we call processes that cause
+We may further attribute a {\em frequency} or {\em rate} to an object
+instead of a process. For example, a heart that beats {\em now} with a
+frequency of 80 beats per minute, or a car that is moving at a speed
+of 180 kilometres per hour {\em at a particular point in time} (e.g.,
+as observed with a speed camera) can be considered attributes of the
+objects (the heart or the car), not attributes of the processes in
+which the objects participate. However, these are {\em different}
+kinds of attributes. Rates, when considered as attributes of objects,
+may be explicitly defined using rates of processes. For example, the
+heart beating frequency of a particular heart $h$ at a time point $t$
+is the frequency of a reference heart beating process in which $h$
+participates. Such a reference process is necessary in order to obtain
+a value for a frequency even when no {\em heart beating} process is
+occurring. However, the frequency is only an attribute of the heart in
+virtue of such a reference process in which {\em heart beating} is
+actually occurring.  This reference process can be uniquely determined
+for continuous processes (such that the frequency or rate of an object
+at a time $t$ is the frequency or rate of the infinitesimally small
+process around $t$), and is ambiguous for discrete processes (in which
+case the reference process must be explicitly stated).
 
-% [Regulation of X vs. X: regulation modulates the number/amount of
-% occurrence...]
+Figure \ref{fig:patterns} illustrates some examples of
+non-comparative and comparative process attributes.
 
-We may say that an object has a {\em frequency} or {\em rate} as if
-these were attributes of objects at time points. For example, that my
-heart {\em now} beats with a frequency of 80 beats per minute, or that
-a car is moving with a speed of 280 kilometres per minute {\em at a
-  particular point in time} (e.g., as observed with a speed camera)
-may be considered attributes of the objects, not attributes of
-processes. In our analysis, rates as attributes of objects can be
-explicitly defined using rates of processes. For example, the heart
-beating frequency of a particular heart $h$ at a time point $t$ is the
-frequency of heart beating processes in which $h$ participates at an
-infinitesimally small process around $t$ and within a reference
-process. The reference process is necessary in order to obtain a value
-for a frequency or rate even if the process in question is not
-occurring. For example, a rate of cardiac muscle contraction can be
-determined even when no cardiac muscle contraction is occurring in
-virtue of a reference process with a rate of cardiac muscle
-contraction.
-
-The reference process for an object at a time point $t$ is uniquely
-determined for continuous processes (the infinitesimally small process
-around $t$), and ambiguous for discrete processes.
-
-A similar construction as for continuous processes can be made for
-discrete processes. A rate of blood flow is immediately an attribute
-of a single heart beat process. However, during a period of time, the
-cumulative rate of blood flow can be observed as the sum of the blood
-flow rates of individual heart beating processes.
+% A similar construction as for continuous processes can be made for
+% discrete processes. A rate of blood flow is immediately an attribute
+% of a single heart beat process. However, during a period of time, the
+% cumulative rate of blood flow can be observed as the sum of the blood
+% flow rates of individual heart beating processes.
 
@@
 \section{Discussion}
-\subsection{Implementation as pre-composed ontology}
-Pre-composed or software support for pattern?
-
 \subsection{Use of CPO}
@@
 
-% \subsection{Future research: acts and intentional states}
-% activities/actions, intentional (goal-oriented) acts
-
-% feeling thirsty (internal state) as driving force behind drinking
-% behavior
-
-% non-measurable, inference only
+\subsection{Pre- vs. post-composed ontology}
+Pre-composed or software support for pattern?
 

### 2011-11-27 paper/drinking.tex

@@
 level across different species as well as for annotation of cellular
-phenotypes in domains in which no such ontology exists.
+phenotypes in domains in which no such ontology exists...
 
@@
 the primary constituents of the facts that go to make up reality''
-\cite{}.
+\cite{Barwise1989}.
 
@@
 
-\subsection{Cellular Phenotype Ontology}
+\subsection{Cell Phenotype Ontology}
 \begin{figure*}[h]
@@
 
-\subsection{Cell Phenotype Ontology: implementation}
+\subsection{Implementation}
 We were faced with two choices for implementing the CPO: we could
@@
 \subsection{Pre- vs. post-composed ontology}
-Pre-composed or software support for pattern?
+We implemented the CPO using only a few patterns for formulating
+phenotypes involving processes.
 
-\subsection{Formalization}
+\subsection{Future work}
 Formal, axiomatic system for process qualities, maybe based on PSL,

### 2011-11-28 paper/drinking.tex
added captions

@@
 \author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Midori
-  A. Harris$^3$ and Heinrich Herre(?) and Gabriella Rustici$^2$ and
+  A. Harris$^2$ and Heinrich Herre$^3$ and Gabriella Rustici$^4$ and
   Georgios V. Gkoutos$^{1}$}
@@
   Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
-  $^{2}$EBI, Wellcome Trust Genome Campus, Hinxton, Cambridge,
-  Cambridge CB10 1SD, UK\\
-  $^{3}$Department of Biochemistry; University of Cambridge, 80 Tennis
-  Court Road, Cambridge CB2 1GA, UK}
+  $^{2}$Department of Biochemistry; University of Cambridge, 80 Tennis
+  Court Road, Cambridge CB2 1GA, UK\\
+  $^{3}$Institute for Medical Informatics, Statistics and
+  Epidemiology, University of Leipzig, Haertelstrasse 16-18, 04107
+  Leipzig, Germany\\
+  $^{4}$EBI, Wellcome Trust Genome Campus, Hinxton, Cambridge,
+  Cambridge CB10 1SD, UK
+}
 
@@
   \includegraphics[width=.45\textwidth]{onto.pdf}
-  \caption{Overview over ...\label{fig:onto}}
+  \caption{Overview over basic top-level categories. Physical objects
+    are entities that are wholly present at time points. Qualities are
+    attributes of entities. Functions are capabilities that arise from
+    physical objects with their qualities, and processes are
+    temporally extended entities that may be realizations of a
+    function.\label{fig:onto}}
 \end{figure}
@@
   \includegraphics[width=.5\textwidth]{processpatterns.pdf}
-  \caption{Overview over ...\label{fig:patterns}}
+
+  \caption{\label{fig:patterns}Six examples of processes with
+    non-comparative and comparative process attributes.  We assume
+    that the processes labelled $a$, $b$, $c$ and $d$ are all
+    instances of the class of processes $P$.  On the left side, three
+    regulation (of $P$) processes are illustrated which exhibit
+    non-comparative attributes. The first process has an attribute of
+    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
+    temporally equidistantly distributed. The second example shows an
+    {\em arrhythmic} occurrence of $P$, and the third examples shows
+    an {\em increasing frequency} (of $P$). A regulation process with
+    an increasing frequency (of $P$) is a process in which the
+    frequency of occurrences of $P$ is lower in the first half of the
+    process than in the second...}
 \end{figure}

### 2011-11-28 paper/drinking.tex
added captions

@@
     frequency of occurrences of $P$ is lower in the first half of the
-    process than in the second...}
+    process than in the second half. The right side of the figure
+    illustrates comparative phenotypic descriptions of processes. On
+    the upper right, the {\em normal} reference is shown. The second
+    example illustrates a {\em late onset} of $P$, i.e., the attribute
+    that $P$ processes begin later than {\em normal}. Finally, the
+    lower right illustrates a {\em decreased frequency} (of $P$),
+    since fewer processes of the type $P$ occur within the reference
+    process than {\em normal}.}
 \end{figure}

### 2011-11-28 paper/drinking.tex

@@
 \subsection{Pre- vs. post-composed ontology}
-We implemented the CPO using only a few patterns for formulating
-phenotypes involving processes.
+We implemented the CPO using a pattern-based approach to formulating
+phenotypes involving processes. The patterns we identify use
+pre-existing ontologies, in particular the PATO ontology and the
+classification of cellular processes as well as cellular components in
+the GO. The result is a large ontology in which classes for phenotypes
+are {\em pre-composed}: they are named and defined within an OWL
+ontology. However, the large size of the resulting ontology may impair
+its utility for data annotation and integration, and software tools
+may not always support such very large ontologies. The alternative to
+pre-composing all possible phenotype classes using the patterns we
+describe is to dynamically generate appropriately defined classes at
+the time at which they are being used. To achieve this goal, software
+must be developed to support ontology users in applying these patterns
+and generate the appropriate class description when required.
 
-\subsection{Future work}
-Formal, axiomatic system for process qualities, maybe based on PSL,
-FOIS paper 2010
+% \subsection{Future work}
+% Formal, axiomatic system for process qualities, maybe based on PSL,
+% FOIS paper 2010
 
-one example:...
+% one example:...
 
-\subsection{Conclusion}
+% \subsection{Conclusion}
 

### 2011-11-28 paper/drinking.tex

@@
   \end{tabular}
-  \caption{\label{tbl:flow}Total cytoplasmic flow rate...}
+  \caption{\label{tbl:flow}Interdependency for the attribute {\em
+      Total cytoplasmic flow rate}. Depending both on whether the
+    cytoplasmic flow rate in individual {\em cytoplasmic streaming}
+    processes is increased or decreased and whether the frequency of
+    occurrence of {\em cytoplasmic streaming} is increased or
+    decreased, the total cytoplasmic flow rate can be increased or
+    decreased.}
 \end{table*}

### 2011-11-28 trunk/paper/drinking.tex

@@
 are being used to reveal candidate genes for orphan diseases and to
-identify chemicals that may have effects on these diseases \cite{}.
+identify chemicals that may have effects on these diseases
+\cite{Schofield2011}.
 
@@
 
-Here, we present the foundations for an ontology of physiological
-phenotypes.
+Here, we present the foundations for an ontology of process
+phenotypes. We present the outline of a theory in which several kinds
+of process attributes can be distinguished so that normal and abnormal
+physiology of biological systems can be formally characterized. We
+apply this theory to cellular processes and create the Cell Phenotype
+Ontology (CPO). The CPO is linked to reference ontologies for
+qualities, biological processes, functions and cell components, and
+its prime application is the unification of phenotypes on the cellular
+level across different species as well as for annotation of cellular
+phenotypes in domains in which no such ontology exists...
 
@@
 such as the various anatomical ontologies, GO \cite {Ashburner2000b},
-the Cell Type Ontology \cite {Bard2005a}, SO \cite {Eilbeck2005} etc
+the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
 with the various qualities it provides for defining how these entities
@@
   \includegraphics[width=.5\textwidth]{processpatterns.pdf}
-  \caption{Overview over ...\label{fig:patterns}}
+
+  \caption{\label{fig:patterns}Six examples of processes with
+    non-comparative and comparative process attributes.  We assume
+    that the processes labelled $a$, $b$, $c$ and $d$ are all
+    instances of the class of processes $P$.  On the left side, three
+    regulation (of $P$) processes are illustrated which exhibit
+    non-comparative attributes. The first process has an attribute of
+    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
+    temporally equidistantly distributed. The second example shows an
+    {\em arrhythmic} occurrence of $P$, and the third examples shows
+    an {\em increasing frequency} (of $P$). A regulation process with
+    an increasing frequency (of $P$) is a process in which the
+    frequency of occurrences of $P$ is lower in the first half of the
+    process than in the second half. The right side of the figure
+    illustrates comparative phenotypic descriptions of processes. On
+    the upper right, the {\em normal} reference is shown. The second
+    example illustrates a {\em late onset} of $P$, i.e., the attribute
+    that $P$ processes begin later than {\em normal}. Finally, the
+    lower right illustrates a {\em decreased frequency} (of $P$),
+    since fewer processes of the type $P$ occur within the reference
+    process than {\em normal}.}
 \end{figure}
@@
 \subsection{Pre- vs. post-composed ontology}
-Pre-composed or software support for pattern?
+We implemented the CPO using a pattern-based approach to formulating
+phenotypes involving processes. The patterns we identify use
+pre-existing ontologies, in particular the PATO ontology and the
+classification of cellular processes as well as cellular components in
+the GO. The result is a large ontology in which classes for phenotypes
+are {\em pre-composed}: they are named and defined within an OWL
+ontology. However, the large size of the resulting ontology may impair
+its utility for data annotation and integration, and software tools
+may not always support such very large ontologies. The alternative to
+pre-composing all possible phenotype classes using the patterns we
+describe is to dynamically generate appropriately defined classes at
+the time at which they are being used. To achieve this goal, software
+must be developed to support ontology users in applying these patterns
+and generate the appropriate class description when required.
 
-\subsection{Formalization}
-Formal, axiomatic system for process qualities, maybe based on PSL,
-FOIS paper 2010
+% \subsection{Future work}
+% Formal, axiomatic system for process qualities, maybe based on PSL,
+% FOIS paper 2010
 
-one example:...
+% one example:...
 
-\subsection{Conclusion}
+% \subsection{Conclusion}
 

### 2011-11-30 paper/drinking.tex

@@
   Guarino1998}.  Ontologies commonly distinguish between {\em classes}
-and {\em individuals} within a domain of knowledge. A class is an
-entity that can have {\em instances}, while individuals are entities
-that cannot be instantiated. Examples of individuals include the
-Eiffel tower or the 2009 Ironman Triathlon in Hawaii, while examples
-of classes include {\em Tower} or {\em Triathlon}. The Eiffel tower
-can be an instance of the class {\em Tower}, and the 2009 Ironman
-Triathlon an instance of {\em Triathlon}.  The meaning of classes is
-specified by stating what must be true of their instances.
+(also called {\em concepts}, {\em categories} or {\em universals}) and
+{\em individuals} within a domain of knowledge. A class is an entity
+that can have {\em instances}, while individuals are entities that
+cannot be instantiated \cite{Herre2006}. Examples of individuals
+include the Eiffel tower or the 2009 Ironman Triathlon in Hawaii,
+while examples of classes include {\em Tower} or {\em Triathlon}. The
+Eiffel tower can be an instance of the class {\em Tower}, and the 2009
+Ironman Triathlon an instance of {\em Triathlon}.  The meaning of
+classes is specified by stating what must be true of their instances.
 
 In addition to classes and individuals, ontologies often include {\em
-  relations}. Relations are the ``the glue that holds things together,
-the primary constituents of the facts that go to make up reality''
-\cite{Barwise1989}.
+  relations}. Relations hold between entities, they are the ``the glue
+that holds things together, the primary constituents of the facts that
+go to make up reality'' \cite{Barwise1989}.
 
@@
 Eventually, a set of {\em primitive terms} remains that are not
-further defined. Using only the primitive terms, we can construct
-complex sentences. Based on the intended meaning of the primitive
-terms, we consider some of these sentences true and some of them false
-in our domain. We select some of the true sentences as {\em axioms}
-which provide the core of our ontology. Ideally, the axioms are chosen
-so that all true sentences in the domain we intend to represent follow
-by means of logical deduction from the axioms. More commonly, however,
+further defined. Following the axiomatic method \cite{Hilbert1918},
+using only the primitive terms, we can construct complex
+sentences. Based on the intended meaning of the primitive terms, we
+consider some of these sentences true and some of them false in our
+domain. We select some of the true sentences as {\em axioms} which
+provide the core of our ontology. Ideally, the axioms are chosen so
+that all true sentences in the domain we intend to represent follow by
+means of logical deduction from the axioms. More commonly, however,
 only {\em some aspects} of the intended meaning are formally
@@
 hidden knowledge and thereby performing powerful queries.  An ontology
-is inconsistent if {\em every} statement can be inferred from the
-ontology. In particular, an ontology is inconsistent if a statement
+is formally inconsistent if there is a statement $\phi$ such that
 $\phi$ and its negation $\neg \phi$ can be inferred from the
-ontology's axioms. Classes in an ontology may be unsatisfiable: a
-class $C$ is unsatisfiable, if it is not possible that the class has
+ontology's axioms.  If an ontology is formally inconsistent, {\em
+  every} statement can be inferred from the ontology.
+
+We assume that the classes of an ontology are
+satisfiable, i.e. that they have instances
+
+Classes in an ontology may be unsatisfiable: a class $C$ is
+unsatisfiable, if it is not possible that the class has
 instances. Unsatisfiable classes are commonly the result of a

### 2011-11-30 paper/drinking.tex

@@
 
-We assume that the classes of an ontology are
-satisfiable, i.e. that they have instances
-
-Classes in an ontology may be unsatisfiable: a class $C$ is
-unsatisfiable, if it is not possible that the class has
-instances. Unsatisfiable classes are commonly the result of a
-contradictory class definition. That a class is unsatisfiable is often
-a special kind of unintended consequence that can be drawn from an
-ontology.
+Automated reasoning can further determine whether classes in an
+ontology are unsatisfiable: a class $C$ is unsatisfiable, if it is
+impossible for the class to have any instances. Unsatisfiable classes
+in an ontology are commonly the result of a contradictory class
+definition. % That a class is unsatisfiable is often a special kind of
+% unintended consequence that can be drawn from an ontology.
 
@@
 
-Processes commonly have participants\footnote{It may be argued that
-  processes necessarily have participants or that processes may have
-  no participants. However, these discussions do not affect our aims
-  here and take no position in this regard.}. Participants of
-processes are entities that are present at time points (i.e., they
-have no temporal parts) and may undergo changes during the process
-\cite{Herre2006, Herre2010}.  We can further distinguish different
-{\em roles} of participation in a process \cite{Loebe2007}. For
-example, a runner may participate in a {\em Triathlon} process as a
-{\em Runner (role)}, while another person can participate in the same
-process but in a different role (such as {\em Referee}).
+Processes commonly have physical objects as participants. Physical
+objects are entities that are present as a whole at time points (i.e.,
+they have no temporal parts) and may persist through time, i.e., they
+may undergo changes during the process \cite{Herre2006, Herre2010}.
+We can further distinguish different {\em roles} of participation in a
+process \cite{Loebe2007}. For example, a runner may participate in a
+{\em Triathlon} process as a {\em Runner (role)}, while another person
+can participate in the same process but in a different role (such as
+{\em Referee}).
 
-Amongst the participants of some processes, we will distinguish
-between {\em inputs} and {\em outputs} of processes. Chemical
-reactions, for example, are processes with definite inputs (the {\em
-  reactants}) and outputs ({\em the products}).
+% Amongst the participants of some processes, we will distinguish
+% between {\em inputs} and {\em outputs} of processes. Chemical
+% reactions, for example, are processes with definite inputs (the {\em
+%   reactants}) and outputs ({\em the products}).
 
@@
 representations of physiology and related phenotypes. In principle, we
-distinguish between three different kinds of process attributes:
-the first are process attributes that arise directly from processes
-and include {\em duration} and {\em temporal location}; the second are
-attributes that arise from processes and their parts and include {\em
-  frequency}; and the third are attributes that arise from processes
-and qualities of their participants, and include {\em flow rates}.
+distinguish between three different kinds of process attributes: the
+first are process attributes that arise directly from processes and
+include {\em duration} and {\em temporal location}; the second are
+attributes that arise from processes and their temporal parts and
+include {\em frequency} and {\em onset}; and the third are attributes
+that arise from processes and qualities of their participants, and
+include {\em flow rates}.
 
-Attribute that can be directly linked to a process arise from
+Attributes that can be directly linked to a process arise from
 processes' temporal extension. For example, a duration is an attribute
@@
 example, the heart may beat {\em rhythmically} or {\em arrhythmically}
-within a period of time.
+within a period of time (see Figure \ref{fig:patterns}).
 
@@
 
-While some processes can be subdivided indefinitely without changing
-the type of process, others cannot. A class $C$ of {\em continuous
-  processes} is a class that has processes $p$ as instance such that
-all temporal parts of $p$ are also instances of $C$. Examples include
-{\em continuous movements} or {\em mass flow} processes. On the other
-hand, {\em discrete processes} can be subdivided into stages of
-activity and stages of inactivity: a class $C$ of {\em discrete
-  processes} is a class with processes $p$ as instance and there is a
-part of $p$ that is not an instance of $C$. An example is {\em heart
-  beating} which has periods of activity (a single heart beat) and
-inactivity.
+While some processes can be subdivided indefinitely while retaining
+certain kinds of attributes, others cannot.
+% A class $C$ of {\em continuous
+%   processes} is a class that has processes $p$ as instance such that
+% all temporal parts of $p$ are also instances of $C$. 
+Examples of processes that can be divided include {\em continuous
+  movements} or {\em mass flow} processes, for which all parts have a
+{\em speed} or {\em flow rate} attribute. On the other hand, some
+processes can be subdivided into stages of activity and stages of
+inactivity and cannot arbitrarily be divided. For example, a process
+of {\em heart beating} has periods of activity (a single heart beat)
+and inactivity. Consequently, not all parts of the process have a {\em
+  heart rate} attribute.
 
@@
 actually occurring.  This reference process can be uniquely determined
-for continuous processes (such that the frequency or rate of an object
-at a time $t$ is the frequency or rate of the infinitesimally small
-process around $t$), and is ambiguous for discrete processes (in which
-case the reference process must be explicitly stated).
+for processes such as {\em continuous movement}, where the rate of an
+object at a time $t$ is the rate of the infinitesimally small process
+that occurs around $t$. The reference process is ambiguous for
+processes such as {\em heart beating}, and the reference process must
+be explicitly stated.
 
-Figure \ref{fig:patterns} illustrates some examples of
-non-comparative and comparative process attributes.
+% Figure \ref{fig:patterns} illustrates some examples of
+% non-comparative and comparative process attributes.
 

### 2011-11-30 trunk/paper/drinking.tex

@@
 hidden knowledge and thereby performing powerful queries.  An ontology
-is inconsistent if {\em every} statement can be inferred from the
-ontology. In particular, an ontology is inconsistent if a statement
+is formally inconsistent if there is a statement $\phi$ such that
 $\phi$ and its negation $\neg \phi$ can be inferred from the
-ontology's axioms. Classes in an ontology may be unsatisfiable: a
-class $C$ is unsatisfiable, if it is not possible that the class has
-instances. Unsatisfiable classes are commonly the result of a
-contradictory class definition. That a class is unsatisfiable is often
-a special kind of unintended consequence that can be drawn from an
-ontology.
+ontology's axioms.  If an ontology is formally inconsistent, {\em
+  every} statement can be inferred from the ontology.
+
+Automated reasoning can further determine whether classes in an
+ontology are unsatisfiable: a class $C$ is unsatisfiable, if it is
+impossible for the class to have any instances. Unsatisfiable classes
+in an ontology are commonly the result of a contradictory class
+definition. % That a class is unsatisfiable is often a special kind of
+% unintended consequence that can be drawn from an ontology.
 
@@
 functioning. For example, abnormalities in cell division resulting in
-{\em Aneuploidy} refer to abnormalities of functioning processes,
-while an {\em increased rate of cell division} refers to an
+{\em Aneuploidy} refer to abnormalities of {\em cell division}
+processes, while an {\em increased rate of cell division} refers to an
 abnormality in the pattern of occurrence of multiple cell division
-processes. In the CPO, abnormalities in the pattern of occurrence of
-$X$ are represented as abnormalities in the {\em regulation of $X$}.
+processes. In the CPO we follow the GO and represent abnormalities in
+the pattern of occurrence of $X$ as abnormalities of {\em regulation
+  of $X$} processes. In particular, an {\em increased rate of cell
+  division} is not an attribute of {\em cell division} processes, but
+rather an attribute of the {\em regulation of cell division}.
 

### 2011-11-30 paper/drinking.tex

@@
 
-When using the GO and PATO ontologies from 30 Nov 2011, CPO contains
-XX classes...
+When using the GO and PATO ontologies from 29 Nov 2011, CPO contains
+ classes...
 Classifies in...

### 2011-12-01 trunk/paper/drinking.tex

@@
 \begin{abstract}
-  Phenotypes important for...
 
-  Include behavioral phenotypes, based on behavior ontology
-
-  We identify a problem in representing phenotypes such as drinking
-  behavior that...
-
-  provide solution...
+  \url{http://cell-phenotype.googlecode.com}.
 \end{abstract}
@@
 level across different species as well as for annotation of cellular
-phenotypes in domains in which no such ontology exists...
+phenotypes in domains in which no such ontology exists.
 
@@
 \subsection{Cell Phenotype Ontology}
-\begin{figure*}[h]
+\begin{figure*}
   \centering
@@
   height=.33\textheight]{overview.pdf}
-  \caption{Overview over ...\label{fig:overview}}
+  \caption{Overview over the taxonomic structure of CPO. The structure
+    is based on a cellular component class $X$ and the cellular
+    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
 \end{figure*}
@@
 $C$ \cite{Hoehndorf2010phene}. For example, an absence of caveolae
-({\tt MP:...}) would be a subclass of {\em Abnormal morphology of
+({\tt MP:0004150}) would be a subclass of {\em Abnormal morphology of
   plasma membrane} in virtue of caveolae necessarily being part of the
@@
 
+\begin{table*}
+  \centering
+  \begin{tabular}{l|l|l|l}
+    & Increased cytoplasmic flow rate & Normal cytoplasmic flow rate & Decreased
+    cytoplasmic flow rate \\
+    \hline
+    Increased frequency &increased total flow rate &increased total
+    flow rate &?\\
+    Normal frequency &increased total flow rate &normal total flow
+    rate &decreased total flow rate\\
+    Decreased frequency &?&decreased total flow rate &decreased total
+    flow rate\\
+    \hline
+  \end{tabular}
+  \caption{\label{tbl:flow}Interdependency for the attribute {\em
+      Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
+      rate} is an attribute of {\em Regulation of cytoplasmic
+      streaming} processes, while {\em Cytoplasmic flow rate} is an
+    attribute of individual {\em cytoplasmic streaming}
+    processes. Depending both on whether the cytoplasmic flow rate in
+    individual {\em cytoplasmic streaming} processes is increased or
+    decreased and whether the frequency of occurrence of {\em
+      cytoplasmic streaming} is increased or decreased, the total
+    cytoplasmic flow rate can be increased or decreased.}
+\end{table*}
+
 A first kind of abnormality of regulatory processes are {\em abnormal
@@
   apoptosis} in which apoptosis is induced later than normal.  We use
-the PATO quality {\em onset} ({\tt PATO:...}) and its children {\em
-  delayed} and {\em premature} to define these types of regulatory
-abnormality. Similarly, we use PATO's {\em offset} ({\tt PATO:...})
-  quality and its children to characterize regulatory abnormalities in
-  which a process ends prematurely or too late.
-
-\begin{table*}
-  \centering
-  \begin{tabular}{l|l|l|l}
-    & Increased cytoplasmic flow rate & Normal cytoplasmic flow rate & Decreased
-    cytoplasmic flow rate \\
-    \hline
-    Increased frequency &increased total flow rate &increased total
-    flow rate &?\\
-    Normal frequency &increased total flow rate &normal total flow
-    rate &decreased total flow rate\\
-    Decreased frequency &?&decreased total flow rate &decreased total
-    flow rate\\
-    \hline
-  \end{tabular}
-  \caption{\label{tbl:flow}Interdependency for the attribute {\em
-      Total cytoplasmic flow rate}. Depending both on whether the
-    cytoplasmic flow rate in individual {\em cytoplasmic streaming}
-    processes is increased or decreased and whether the frequency of
-    occurrence of {\em cytoplasmic streaming} is increased or
-    decreased, the total cytoplasmic flow rate can be increased or
-    decreased.}
-\end{table*}
+the PATO quality {\em onset} ({\tt PATO:0002325}) and its children
+{\em delayed} and {\em premature} to define these types of regulatory
+abnormality. Similarly, we use PATO's {\em offset} ({\tt
+  PATO:0002324}) quality and its children to characterize regulatory
+abnormalities in which a process ends prematurely or too late.
 
@@
 
+\begin{table*}
+  \centering
+  \begin{tabular}{|p{7cm}|l|p{4cm}|}
+    Neumann et al.\cite{Neumann2010}&Schmitz et
+    al.\cite{Schmitz2010}&Fuchs et al.\cite{Fuchs2010}\\
+    \hline
+    binuclear & normal mitotic exit & actin fiver cells\\
+    cell death & prolonged mitotic exit & big cells \\
+    cell migration & & bright and large cells phenotype \\
+    condensation followed by decondensation without completion of
+    mitosis & & bright nuclei\\
+    condensation without mitosis/collapse of nucleus & & cells with protrusions\\
+    dynamic changes & & elongated cells\\
+    failure in decondensation & & elongated cells with protrusions\\
+    grape& & high actin ratio cells \\
+    increased proliferation& & lamellipodia + high actin ratio cells\\
+    large& & lamellipodia cells\\
+    large nucleus& & large cells\\
+    metaphase alignment problems/including no metaphase& & large nuclei\\
+    metaphase delay/arrest& & low eccentricity cells\\
+    migration (distance)& & metaphase cells\\
+    migration (speed)& & proliferating cells\\
+    mitotic delay/arrest& & small cells\\
+    nuclei stay close together& & small cells with an enrichment of mitotic cells\\
+    polylobed& &\\
+    pulsating nuclei& &\\
+    segregation problems/chromatin bridges/lagging chromosomes/multiple DNA masses&&\\
+    small nucleus&&\\
+    strange nuclear shape&&\\
+    \hline
+  \end{tabular}
+  \caption{\label{tbl:studies}The table summarizes cellular phenotype
+    terms used in three recent systems microscopy studies.}
+\end{table*}
+
 We use the labels of GO classes to automatically generate class labels
@@
 occurrences of {\em Apoptosis} is {\em Increased frequency of
-  occurrences of Apoptosis}, and its textual definition states: ``An
-increased frequency of occurrences of Apoptosis is a phenotype of
-entities in which the number of occurrences of Apoptosis within a
-given time period is increased in comparison to a reference process.''
+  occurrences of Apoptosis}, and its textual definition states that an
+increased frequency of occurrences of {\em Apoptosis} is a phenotype
+of {\em Regulation of apoptosis} in which the number of occurrences of
+{\em Apoptosis} within a given time period is increased in comparison
+to a reference process that is considered {\em normal}.
 
 When using the GO and PATO ontologies from 30 Nov 2011, CPO contains
-XX classes...
-Classifies in...
+125,466 classes of which 79,236 are explicitly defined.  The ELK
+reasoner \cite{Kazakov2011} is able to perform a classification of the
+ontology in under 10 seconds. We make the ontology and the source code
+that is used to generate it freely available on
+\url{http://cell-phenotype.googlecode.com}.
 
 \section{Discussion}
-\subsection{Use of CPO}
+
+\subsection{Applications of the CPO}
 The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
@@
 cellular phenotypes. Over 80\% of FYPO definitions reference
-descendants of GO BP 'cellular process' as the entity; a further 11\%
-reference GO CC terms. All FYPO explicit definitions reference
+descendants of GO-BP's {\em Cellular process} as the entity; a further
+11\% reference GO-CC terms. All FYPO explicit definitions reference
 qualities in PATO, including {\em normal}, {\em abnormal}, and several
-process qualities including {\em increased duration}, {\em decreased
-  occurrence}, etc. FYPO will thus fit neatly under the CPO umbrella,
-and stands to benefit from the automated synchronization between CPO
-and GO. {\em S. pombe} annotations to FYPO terms will provide a rich
-body of highly specific, well-supported data to be integrated with
-data from other species.
+process qualities including {\em increased duration} and {\em
+  decreased occurrence}. FYPO will thus fit neatly under the CPO
+umbrella, and stands to benefit from the automated synchronization
+between CPO and GO, as well as the integration of cellular phenotypes
+across species that the CPO can provide. {\em Schizosaccharomyces
+  pombe} annotations to FYPO terms will provide a rich body of highly
+specific, well-supported data to be integrated with data from other
+species.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-A domain that would greatly benefit from the development of a CPO is
-Systems Microscopy, which aims to understand complex and dynamic
+
+A further domain that will greatly benefit from the CPO is {\em
+  systems microscopy}, which aims to understand complex and dynamic
 cellular systems by combining automated fluorescence microscopy, cell
 microarray platforms, quantitative image analysis and data mining
-(Lock \& Stromblad, Exp. Cell Res. 2010).
-
-If we consider some of the studies, which have been published in this
-field in the last few years (Neumann et al, Nature 464: 722-727;
-Schmitz et al, Nature Cell Biology 12: 886-893; Fuchs et al, Mol Sys
-Biology 6: 370), the need for a CPO becomes evident.
-
-In all three studies here considered live-cell imaging assays and RNAi
-knockdown were used to generate phenotypic profiles that quantify the
-cellular response to a given siRNA thus allowing identification of
-hundreds of genes involved in diverse biological functions including
-cell division, migration and survival.
-
+\cite{Lock2010}.  If we consider some of the studies, which have been
+published in this field in the last few years \cite{Neumann2010,
+  Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
+three studies illustrated in Table \ref{tbl:studies}, live-cell
+imaging assays and RNAi knockdown were used to generate phenotypic
+profiles that quantify the cellular response to a given siRNA thus
+allowing identification of hundreds of genes involved in diverse
+biological functions including cell division, migration and survival.
 In each study, several phenotypes were detected and described by the
-authors without the use of ontologies (see table), making the
-integration between datasets extremely difficult. For example, it is
-evident that cell division phenotypes were observed in the three
-datasets (e.g. 'Mitotic delay/arrest' and 'Prolonged mitotic exit' or
-'Methaphase delay' and 'Methaphase cells') but the overlap between
-such phenotypes is unclear.
+authors without the use of ontologies (see Table \ref{tbl:studies},
+making the integration between datasets extremely difficult. For
+example, it is evident that cell division phenotypes were observed in
+all three datasets (e.g. {\em Mitotic delay/arrest} and {\em Prolonged
+  mitotic exit} or {\em Methaphase delay} and {\em Methaphase cells})
+but the overlap between such phenotypes is unclear.
 
@@
 (http://www.systemsmicroscopy.eu/) and the first step towards data
-integration will be achieved by developing a CPO.
+integration can be achieved by further developing CPO.
 
@@
 processing pipeline and used to annotate the data as it gets generated
-(Conrad et al, Nature Methods 8: 246-249).
+\cite{Conrad2011}.
 
-\subsection{Pre- vs. post-composed ontology}
+\subsection{Future research}
 We implemented the CPO using a pattern-based approach to formulating
-phenotypes involving processes. The patterns we identify use
+phenotypes involving processes. The patterns we identify are based on
 pre-existing ontologies, in particular the PATO ontology and the
 classification of cellular processes as well as cellular components in
-the GO. The result is a large ontology in which classes for phenotypes
-are {\em pre-composed}: they are named and defined within an OWL
-ontology. However, the large size of the resulting ontology may impair
-its utility for data annotation and integration, and software tools
-may not always support such very large ontologies. The alternative to
-pre-composing all possible phenotype classes using the patterns we
-describe is to dynamically generate appropriately defined classes at
-the time at which they are being used. To achieve this goal, software
-must be developed to support ontology users in applying these patterns
-and generate the appropriate class description when required.
+the GO. The result of our method is a large ontology in which classes
+for phenotypes are {\em pre-composed}: they are named and defined
+within an OWL ontology. However, the large size of the resulting
+ontology may impair its utility for data annotation and integration,
+and software tools may not always support such very large
+ontologies. The alternative to pre-composing all possible phenotype
+classes using the patterns we describe is to dynamically generate
+appropriately defined classes at the time at which they are being
+used. To achieve this goal, software must be developed to support
+ontology users in applying these patterns and generate the appropriate
+class description when required.
 
-% \subsection{Future work}
-% Formal, axiomatic system for process qualities, maybe based on PSL,
-% FOIS paper 2010
-
-% one example:...
-
-% \subsection{Conclusion}
+A further important task is to develop the theory we outlined and
+applied for the CPO. In particular, a precise formal characterization
+of this theory in terms of axioms will further improve the clarity of
+phenotypic descriptions of processes and enable its integration in
+well-developed formal ontologies of processes \cite{Herre2006,
+  Gruninger2010}.
 
 \section{Acknowledgements}
-MAH is supported by Wellcome Trust grant WT090548MA.
+Funding for RH is provided by the European Commission's 7th
+Framework Programme, RICORDO project, grant number 248502. MAH is
+supported by Wellcome Trust, grant WT090548MA. HH is supported by the
+Institute for Medical Informatics, Statistics and Epidemiology,
+University of Leipzig. GVG is funded by BBSRC grant BBG0043581.
 

### 2011-12-01 paper/drinking.tex

@@
 \begin{abstract}
-  Phenotypes important for...
+  The systematic observation of phenotypes has become a crucial tool
+  of functional genomics, and several large international projects are
+  currently underway to identify and characterize the phenotypes that
+  are associated with genotypes in several species. To integrate
+  phenotype descriptions within and across species, phenotype
+  ontologies have been developed.
 
-  Include behavioral phenotypes, based on behavior ontology
+  In particular phenotypes on a cellular scale
 
-  We identify a problem in representing phenotypes such as drinking
-  behavior that...
-
-  provide solution...
+  \url{http://cell-phenotype.googlecode.com}.
 \end{abstract}
@@
 integrate biomedical data across heterogeneous information systems,
-biomedical ontologies were being developed \cite{Smith2007}. An
+biomedical ontologies are being developed \cite{Smith2007}. An
 ontology is an explicit specification of a conceptualization of a
@@
 methodology used to semantically categorise phenodeviance.  The
-biomedical research community has invested considerable amount of
-effort and resources in the development and establishment of
-ontologies that are becoming increasingly successful as information
-management and integration tools in many disparate scientific fields
-allowing interoperability and semantic information processing between
-diverse biomedical resources and domains.
+biomedical research community has invested considerable effort and
+resources in the development and establishment of ontologies that are
+becoming increasingly successful as information management and
+integration tools in many disparate scientific fields allowing
+interoperability and semantic information processing between diverse
+biomedical resources and domains.
 
@@
   reasoning}. Automated reasoning allows users of an ontology to carry
-our key activities: verifying the ontology's consistency, inferring
+out key activities: verifying the ontology's consistency, inferring
 hidden knowledge and thereby performing powerful queries.  An ontology
@@
 \subsection{PATO and the EQ model}
-The Phenotype And Trait Ontology (PATO) was envisaged and designed to
-provide a platform for allowing the integration of quantitative and
-qualitative phenotype related information across different levels of
-granularity of domain knowledge granularity and species
-\cite{Gkoutos2005}.  PATO forms an ontology of phenotype qualities
-that form basic entities that we can perceive and/or measure such as
-colors, sizes, rates etc. One of it's classification axis lies on the
-basis of the basic type of entity these qualities inhere to and
-distinguishes between qualities of physical objects and qualities of
-processes.
+PATO was envisaged and designed to provide a platform for allowing the
+integration of quantitative and qualitative phenotype related
+information across different domains, levels of granularity and
+species \cite{Gkoutos2005}.  PATO is an ontology of phenotype
+qualities that form basic entities that we can perceive and/or measure
+such as colors, sizes, rates etc. One of its classification axes is
+based on the basic type of entity to which a qualities belongs, and
+PATO distinguishes between qualities of physical objects and qualities
+of processes.
 
@@
 \textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
-existed in a pre-composed ontology (for the {\tt MP:0009007} term from
-the Mammalian Phenotype) it could be used to provide an equivalence
-statement between that class and the above PATO-based description.
+existed in a pre-composed ontology (for example the {\tt MP:0009007}
+term from the Mammalian Phenotype) it could be used to provide an
+equivalence statement between that class and the above PATO-based
+description.
 
@@
 occurrences of {\em Apoptosis} is {\em Increased frequency of
-  occurrences of Apoptosis}, and its textual definition states: ``An
-increased frequency of occurrences of Apoptosis is a phenotype of
-entities in which the number of occurrences of Apoptosis within a
-given time period is increased in comparison to a reference process.''
+  occurrences of Apoptosis}, and its textual definition states that an
+increased frequency of occurrences of {\em Apoptosis} is a phenotype
+of {\em Regulation of apoptosis} in which the number of occurrences of
+{\em Apoptosis} within a given time period is increased in comparison
+to a reference process that is considered {\em normal}.
 
-When using the GO and PATO ontologies from 29 Nov 2011, CPO contains
- classes...
-Classifies in...
+As of November 2011, CPO contains 125,466 classes of which 79,236 are
+explicitly defined.  The ELK reasoner \cite{Kazakov2011} is able to
+perform a classification of the ontology in under 10 seconds. We make
+the ontology and the source code that is used to generate it freely
+available on \url{http://cell-phenotype.googlecode.com}.
 
 \section{Discussion}
-\subsection{Use of CPO}
+
+\subsection{Applications of the CPO}
 The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
@@
 cellular phenotypes. Over 80\% of FYPO definitions reference
-descendants of GO BP 'cellular process' as the entity; a further 11\%
-reference GO CC terms. All FYPO explicit definitions reference
+descendants of GO-BP's {\em Cellular process} as the entity; a further
+11\% reference GO-CC terms. All FYPO explicit definitions reference
 qualities in PATO, including {\em normal}, {\em abnormal}, and several
-process qualities including {\em increased duration}, {\em decreased
-  occurrence}, etc. FYPO will thus fit neatly under the CPO umbrella,
-and stands to benefit from the automated synchronization between CPO
-and GO. {\em S. pombe} annotations to FYPO terms will provide a rich
-body of highly specific, well-supported data to be integrated with
-data from other species.
+process qualities including {\em increased duration} and {\em
+  decreased occurrence}. FYPO will thus fit neatly under the CPO
+umbrella, and stands to benefit from the automated synchronization
+between CPO and GO, as well as the integration of cellular phenotypes
+across species that the CPO can provide. {\em Schizosaccharomyces
+  pombe} annotations to FYPO terms will provide a rich body of highly
+specific, well-supported data to be integrated with data from other
+species.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-A domain that would greatly benefit from the development of a CPO is
-Systems Microscopy, which aims to understand complex and dynamic
+
+A further domain that will greatly benefit from the CPO is {\em
+  systems microscopy}, which aims to understand complex and dynamic
 cellular systems by combining automated fluorescence microscopy, cell
 microarray platforms, quantitative image analysis and data mining
-(Lock \& Stromblad, Exp. Cell Res. 2010).
-
-If we consider some of the studies, which have been published in this
-field in the last few years (Neumann et al, Nature 464: 722-727;
-Schmitz et al, Nature Cell Biology 12: 886-893; Fuchs et al, Mol Sys
-Biology 6: 370), the need for a CPO becomes evident.
-
-In all three studies here considered live-cell imaging assays and RNAi
-knockdown were used to generate phenotypic profiles that quantify the
-cellular response to a given siRNA thus allowing identification of
-hundreds of genes involved in diverse biological functions including
-cell division, migration and survival.
-
+\cite{Lock2010}.  If we consider some of the studies, which have been
+published in this field in the last few years \cite{Neumann2010,
+  Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
+three studies illustrated in Table \ref{tbl:studies}, live-cell
+imaging assays and RNAi knockdown were used to generate phenotypic
+profiles that quantify the cellular response to a given siRNA thus
+allowing identification of hundreds of genes involved in diverse
+biological functions including cell division, migration and survival.
 In each study, several phenotypes were detected and described by the
-authors without the use of ontologies (see table), making the
-integration between datasets extremely difficult. For example, it is
-evident that cell division phenotypes were observed in the three
-datasets (e.g. 'Mitotic delay/arrest' and 'Prolonged mitotic exit' or
-'Methaphase delay' and 'Methaphase cells') but the overlap between
-such phenotypes is unclear.
+authors without the use of ontologies (see Table \ref{tbl:studies}),
+making the integration between datasets extremely difficult. For
+example, it is evident that cell division phenotypes were observed in
+all three datasets (e.g. {\em Mitotic delay/arrest} and {\em Prolonged
+  mitotic exit} or {\em Methaphase delay} and {\em Methaphase cells})
+but the overlap between such phenotypes is unclear.
 
@@
 processing pipeline and used to annotate the data as it gets generated
-(Conrad et al, Nature Methods 8: 246-249).
+\cite{Conrad2011}.
 
-\subsection{Pre- vs. post-composed ontology}
+\subsection{Future research}
 We implemented the CPO using a pattern-based approach to formulating
-phenotypes involving processes. The patterns we identify use
+phenotypes involving processes. The patterns we identify are based on
 pre-existing ontologies, in particular the PATO ontology and the
 classification of cellular processes as well as cellular components in
-the GO. The result is a large ontology in which classes for phenotypes
-are {\em pre-composed}: they are named and defined within an OWL
-ontology. However, the large size of the resulting ontology may impair
-its utility for data annotation and integration, and software tools
-may not always support such very large ontologies. The alternative to
-pre-composing all possible phenotype classes using the patterns we
-describe is to dynamically generate appropriately defined classes at
-the time at which they are being used. To achieve this goal, software
-must be developed to support ontology users in applying these patterns
-and generate the appropriate class description when required.
+the GO. The result of our method is a large ontology in which classes
+for phenotypes are {\em pre-composed}: they are named and defined
+within an OWL ontology. However, the large size of the resulting
+ontology may impair its utility for data annotation and integration,
+and software tools may not always support such very large
+ontologies. The alternative to pre-composing all possible phenotype
+classes using the patterns we describe is to dynamically generate
+appropriately defined classes at the time at which they are being
+used. To achieve this goal, software must be developed to support
+ontology users in applying these patterns and generate the appropriate
+class description when required.
 
-% \subsection{Future work}
-% Formal, axiomatic system for process qualities, maybe based on PSL,
-% FOIS paper 2010
-
-% one example:...
-
-% \subsection{Conclusion}
+A further important task is to develop the theory we outlined and
+applied for the CPO. In particular, a precise formal characterization
+of this theory in terms of axioms will further improve the clarity of
+phenotypic descriptions of processes and enable its integration in
+well-developed formal ontologies of processes \cite{Herre2006,
+  Gruninger2010}.
 
 \section{Acknowledgements}
-MAH is supported by Wellcome Trust grant WT090548MA.
+Funding for RH is provided by the European Commission's 7th
+Framework Programme, RICORDO project, grant number 248502. MAH is
+supported by Wellcome Trust, grant WT090548MA. HH is supported by the
+Institute for Medical Informatics, Statistics and Epidemiology,
+University of Leipzig. GVG is funded by NIH grant number R01
+HG004838-02.
 

### 2011-12-01 trunk/paper/drinking.tex

@@
 
-\title{Process-based phenotypes: an application to the ontology of
-  cell physiology}
+\title{Semantic integration of physiology phenotypes: an application
+  to the cellular phenotype ontology}
 
@@
   $^{4}$EBI, Wellcome Trust Genome Campus, Hinxton, Cambridge,
-  Cambridge CB10 1SD, UK
-}
+  Cambridge CB10 1SD, UK}
 
@@
 \begin{abstract}
-
+  The systematic observation of phenotypes has become a crucial tool
+  of functional genomics, and several large international projects are
+  currently underway to identify and characterize the phenotypes that
+  are associated with genotypes in several species. To integrate
+  phenotype descriptions within and across species, phenotype
+  ontologies have been developed. Applying ontologies to unify
+  phenotype descriptions in the domain of physiology has been a
+  particular challenge due to the high complexity of the underlying
+  domain.
+  Here, we present the outline of a theory for an ontology of
+  physiology-related phenotypes. We provide a formal description of
+  process attributes and relate them to the attributes of their parts
+  and participants. We apply our theory to create the Cellular
+  Phenotype Ontology (CPO). The CPO is an ontology of morphological
+  and physiological phenotypic abnormalities of cells, cell components
+  and cellular processes. Its prime application is the unification of
+  cellular phenotype descriptions across species by providing terms
+  and uniform definition patterns. The CPO can further be used for the
+  annotation of observed abnormalities in domains, such as systems
+  microscopy, in which cellular abnormalities are observed and for
+  which no phenotype ontology has been created.  The CPO and
+  associated source code is available on
   \url{http://cell-phenotype.googlecode.com}.
@@
 
-When using the GO and PATO ontologies from 30 Nov 2011, CPO contains
-125,466 classes of which 79,236 are explicitly defined.  The ELK
-reasoner \cite{Kazakov2011} is able to perform a classification of the
-ontology in under 10 seconds. We make the ontology and the source code
-that is used to generate it freely available on
-\url{http://cell-phenotype.googlecode.com}.
+As of November 2011, CPO contains 125,466 classes of which 79,236 are
+explicitly defined.  The ELK reasoner \cite{Kazakov2011} is able to
+perform a classification of the ontology in under 10 seconds. We make
+the ontology and the source code that is used to generate it freely
+available on \url{http://cell-phenotype.googlecode.com}.
 
@@
 In each study, several phenotypes were detected and described by the
-authors without the use of ontologies (see Table \ref{tbl:studies},
+authors without the use of ontologies (see Table \ref{tbl:studies}),
 making the integration between datasets extremely difficult. For
@@
 \section{Acknowledgements}
-Funding for RH is provided by the European Commission's 7th
-Framework Programme, RICORDO project, grant number 248502. MAH is
-supported by Wellcome Trust, grant WT090548MA. HH is supported by the
-Institute for Medical Informatics, Statistics and Epidemiology,
-University of Leipzig. GVG is funded by BBSRC grant BBG0043581.
+Funding for RH is provided by the European Commission's 7th Framework
+Programme, RICORDO project, grant number 248502. MAH is supported by
+Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
+Medical Informatics, Statistics and Epidemiology, University of
+Leipzig.  Funding for GR is provided by the European Union's Seventh
+Framework Programme (FP7/2007-2013) under grant agreement n°258068;
+EU-FP7-Systems Microscopy NoE. GVG is funded by NIH grant number R01
+HG004838-02.
 

### 2011-12-01 trunk/paper/drinking.tex

@@
   particular challenge due to the high complexity of the underlying
-  domain.
-  Here, we present the outline of a theory for an ontology of
+  domain.  Here, we present the outline of a theory for an ontology of
   physiology-related phenotypes. We provide a formal description of
@@
   microscopy, in which cellular abnormalities are observed and for
-  which no phenotype ontology has been created.  The CPO and
-  associated source code is available on
+  which no phenotype ontology has been created.  The CPO and the
+  source code we generated to create the CPO are available on
   \url{http://cell-phenotype.googlecode.com}.
@@
 Leipzig.  Funding for GR is provided by the European Union's Seventh
-Framework Programme (FP7/2007-2013) under grant agreement n°258068;
-EU-FP7-Systems Microscopy NoE. GVG is funded by NIH grant number R01
-HG004838-02.
+Framework Programme (FP7/2007-2013) under grant agreement number
+258068, EU-FP7-Systems Microscopy NoE. GVG is funded by NIH grant
+number R01 HG004838-02.
 

### 2011-12-01 trunk/paper/drinking.tex

@@
   Leipzig, Germany\\
-  $^{4}$EBI, Wellcome Trust Genome Campus, Hinxton, Cambridge,
+  $^{4}$European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridge,
   Cambridge CB10 1SD, UK}
@@
 Framework Programme (FP7/2007-2013) under grant agreement number
-258068, EU-FP7-Systems Microscopy NoE. GVG is funded by NIH grant
-number R01 HG004838-02.
+258068, EU-FP7-Systems Microscopy NoE. 
+%GVG is funded by NIH grant number R01 HG004838-02.
 

### 2011-12-01 paper/drinking.tex

@@
 
-\title{Process-based phenotypes: an application to the ontology of
-  cell physiology}
+\title{Semantic integration of physiology phenotypes: an application
+  to the cellular phenotype ontology}
 
-\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Midori
+\author{Robert Hoehndorf$^1$ and Midori
   A. Harris$^2$ and Heinrich Herre$^3$ and Gabriella Rustici$^4$ and
@@
   Leipzig, Germany\\
-  $^{4}$EBI, Wellcome Trust Genome Campus, Hinxton, Cambridge,
-  Cambridge CB10 1SD, UK
-}
+  $^{4}$European Bioinformatics Institute, Wellcome Trust Genome
+  Campus, Hinxton, Cambridge, Cambridge CB10 1SD, UK}
 
@@
   phenotype descriptions within and across species, phenotype
-  ontologies have been developed.
-
-  In particular phenotypes on a cellular scale
-
+  ontologies have been developed. Applying ontologies to unify
+  phenotype descriptions in the domain of physiology has been a
+  particular challenge due to the high complexity of the underlying
+  domain.  Here, we present the outline of a theory for an ontology of
+  physiology-related phenotypes. We provide a formal description of
+  process attributes and relate them to the attributes of their parts
+  and participants. We apply our theory to create the Cellular
+  Phenotype Ontology (CPO). The CPO is an ontology of morphological
+  and physiological phenotypic abnormalities of cells, cell components
+  and cellular processes. Its prime application is the unification of
+  cellular phenotype descriptions across species by providing terms
+  and uniform definition patterns. The CPO can further be used for the
+  annotation of observed abnormalities in domains, such as systems
+  microscopy, in which cellular abnormalities are observed and for
+  which no phenotype ontology has been created.  The CPO and the
+  source code we generated to create the CPO are available on
   \url{http://cell-phenotype.googlecode.com}.
@@
 \section{Acknowledgements}
-Funding for RH is provided by the European Commission's 7th
-Framework Programme, RICORDO project, grant number 248502. MAH is
-supported by Wellcome Trust, grant WT090548MA. HH is supported by the
-Institute for Medical Informatics, Statistics and Epidemiology,
-University of Leipzig. GVG is funded by NIH grant number R01
-HG004838-02.
+Funding for RH is provided by the European Commission's 7th Framework
+Programme, RICORDO project, grant number 248502. MAH is supported by
+Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
+Medical Informatics, Statistics and Epidemiology, University of
+Leipzig.  Funding for GR is provided by the European Union's Seventh
+Framework Programme (FP7/2007-2013) under grant agreement number
+258068, EU-FP7-Systems Microscopy NoE. 
+%GVG is funded by NIH grant number R01 HG004838-02.
 

### 2012-01-19 trunk/paper/drinking-bioinformatics.tex

--- 2011-12-01 paper/drinking.tex
@@
 \documentclass{bioinfo}
-\usepackage{setspace}
-
 \copyrightyear{2005}
@@
 \usepackage{multirow}
-\usepackage{pifont}
-% Load packages
-\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
-\usepackage{url}  % Formatting web addresses  
-\usepackage{ifthen}  % Conditional 
-\usepackage[utf8]{inputenc} %unicode support
-\urlstyle{rm}
-\usepackage{natbib}
-\renewcommand{\cite}{\citep}
-\usepackage{amsmath}
-\usepackage{amssymb}
-
 \usepackage{pifont}
@@
 
-\title{Semantic integration of physiology phenotypes: an application
-  to the cellular phenotype ontology}
+\title[Cell Phenotype Ontology]{Semantic integration of physiology
+  phenotypes: an application to the cellular phenotype ontology}
 
-\author{Robert Hoehndorf$^1$ and Midori
-  A. Harris$^2$ and Heinrich Herre$^3$ and Gabriella Rustici$^4$ and
+\author[Hoehndorf \textit{et~al}]{Robert Hoehndorf$^{1}$\footnote{to
+    whom correspondence should be addressed}, Midori
+  A. Harris$^2$, Heinrich Herre$^3$, Gabriella Rustici$^4$ and
   Georgios V. Gkoutos$^{1}$}
@@
 
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+
+\editor{Associate Editor: XXXXXXX}
+
 \maketitle
@@
 \begin{abstract}
-  The systematic observation of phenotypes has become a crucial tool
-  of functional genomics, and several large international projects are
+\section{Motivation:}
+The systematic observation of phenotypes has become a crucial tool of
+functional genomics, and several large international projects are
 currently underway to identify and characterize the phenotypes that
 are associated with genotypes in several species. To integrate
-  phenotype descriptions within and across species, phenotype
-  ontologies have been developed. Applying ontologies to unify
-  phenotype descriptions in the domain of physiology has been a
-  particular challenge due to the high complexity of the underlying
-  domain.  Here, we present the outline of a theory for an ontology of
+phenotype descriptions within and across species, phenotype ontologies
+have been developed. Applying ontologies to unify phenotype
+descriptions in the domain of physiology has been a particular
+challenge due to the high complexity of the underlying domain.
+
+\section{Results:}
+Here, we present the outline of a theory for an ontology of
 physiology-related phenotypes. We provide a formal description of
 process attributes and relate them to the attributes of their parts
-  and participants. We apply our theory to create the Cellular
-  Phenotype Ontology (CPO). The CPO is an ontology of morphological
-  and physiological phenotypic abnormalities of cells, cell components
-  and cellular processes. Its prime application is the unification of
-  cellular phenotype descriptions across species by providing terms
-  and uniform definition patterns. The CPO can further be used for the
+and participants. We apply our theory to create the Cellular Phenotype
+Ontology (CPO). The CPO is an ontology of morphological and
+physiological phenotypic abnormalities of cells, cell components and
+cellular processes. Its prime application is the unification of
+cellular phenotype descriptions across species by providing terms and
+uniform definition patterns. The CPO can further be used for the
 annotation of observed abnormalities in domains, such as systems
-  microscopy, in which cellular abnormalities are observed and for
-  which no phenotype ontology has been created.  The CPO and the
-  source code we generated to create the CPO are available on
-  \url{http://cell-phenotype.googlecode.com}.
+microscopy, in which cellular abnormalities are observed and for which
+no phenotype ontology has been created.
+
+\section{Availability and implementation:} 
+The CPO and the source code we generated to create the CPO are freely
+available on \url{http://cell-phenotype.googlecode.com}.
+
+\section{Contact:} \href{rh497@cam.ac.uk}{rh497@cam.ac.uk}
 \end{abstract}
-
-{\bf Keywords:} cell phenotype, process, ontology, Semantic Web
 
@@
 
-\section{Materials and Methods}
+\begin{methods}
+\section{System and methods}
 \subsection{Formal ontology}
@@
 \subsection{Processes and their participants}
-\begin{figure}
-  \centering
-  \includegraphics[width=.45\textwidth]{onto.pdf}
-  \caption{Overview over basic top-level categories. Physical objects
-    are entities that are wholly present at time points. Qualities are
-    attributes of entities. Functions are capabilities that arise from
-    physical objects with their qualities, and processes are
-    temporally extended entities that may be realizations of a
-    function.\label{fig:onto}}
-\end{figure}
-
 Most biomedical ontologies share common distinctions between different
@@
 description.
+
+\end{methods}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=.5\textwidth]{processpatterns.pdf}
-
-  \caption{\label{fig:patterns}Six examples of processes with
-    non-comparative and comparative process attributes.  We assume
-    that the processes labelled $a$, $b$, $c$ and $d$ are all
-    instances of the class of processes $P$.  On the left side, three
-    regulation (of $P$) processes are illustrated which exhibit
-    non-comparative attributes. The first process has an attribute of
-    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
-    temporally equidistantly distributed. The second example shows an
-    {\em arrhythmic} occurrence of $P$, and the third examples shows
-    an {\em increasing frequency} (of $P$). A regulation process with
-    an increasing frequency (of $P$) is a process in which the
-    frequency of occurrences of $P$ is lower in the first half of the
-    process than in the second half. The right side of the figure
-    illustrates comparative phenotypic descriptions of processes. On
-    the upper right, the {\em normal} reference is shown. The second
-    example illustrates a {\em late onset} of $P$, i.e., the attribute
-    that $P$ processes begin later than {\em normal}. Finally, the
-    lower right illustrates a {\em decreased frequency} (of $P$),
-    since fewer processes of the type $P$ occur within the reference
-    process than {\em normal}.}
-\end{figure}
-
 % Complex relations between Based on these definition patterns,
@@
 \subsection{Cell Phenotype Ontology}
-\begin{figure*}
-  \centering
-  \includegraphics[width=\textwidth,
-  height=.33\textheight]{overview.pdf}
-  \caption{Overview over the taxonomic structure of CPO. The structure
-    is based on a cellular component class $X$ and the cellular
-    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
-\end{figure*}
-
 While our considerations about process attributes are only the
@@
   \centering
-  \begin{tabular}{l|l|l|l}
-    & Increased cytoplasmic flow rate & Normal cytoplasmic flow rate & Decreased
-    cytoplasmic flow rate \\
+  \begin{tabular}{p{2.8cm}|p{4cm}|p{4cm}|p{4cm}}
+    & Increased flow rate & Normal flow rate & Decreased
+    flow rate \\
     \hline
@@
 
-\begin{table*}
-  \centering
-  \begin{tabular}{|p{7cm}|l|p{4cm}|}
-    Neumann et al.\cite{Neumann2010}&Schmitz et
-    al.\cite{Schmitz2010}&Fuchs et al.\cite{Fuchs2010}\\
-    \hline
-    binuclear & normal mitotic exit & actin fiver cells\\
-    cell death & prolonged mitotic exit & big cells \\
-    cell migration & & bright and large cells phenotype \\
-    condensation followed by decondensation without completion of
-    mitosis & & bright nuclei\\
-    condensation without mitosis/collapse of nucleus & & cells with protrusions\\
-    dynamic changes & & elongated cells\\
-    failure in decondensation & & elongated cells with protrusions\\
-    grape& & high actin ratio cells \\
-    increased proliferation& & lamellipodia + high actin ratio cells\\
-    large& & lamellipodia cells\\
-    large nucleus& & large cells\\
-    metaphase alignment problems/including no metaphase& & large nuclei\\
-    metaphase delay/arrest& & low eccentricity cells\\
-    migration (distance)& & metaphase cells\\
-    migration (speed)& & proliferating cells\\
-    mitotic delay/arrest& & small cells\\
-    nuclei stay close together& & small cells with an enrichment of mitotic cells\\
-    polylobed& &\\
-    pulsating nuclei& &\\
-    segregation problems/chromatin bridges/lagging chromosomes/multiple DNA masses&&\\
-    small nucleus&&\\
-    strange nuclear shape&&\\
-    \hline
-  \end{tabular}
-  \caption{\label{tbl:studies}The table summarizes cellular phenotype
-    terms used in three recent systems microscopy studies.}
-\end{table*}
+% \begin{table*}
+%   \centering
+%   \begin{tabular}{|p{7cm}|l|p{4cm}|}
+%     Neumann et al.\cite{Neumann2010}&Schmitz et
+%     al.\cite{Schmitz2010}&Fuchs et al.\cite{Fuchs2010}\\
+%     \hline
+%     binuclear & normal mitotic exit & actin fiver cells\\
+%     cell death & prolonged mitotic exit & big cells \\
+%     cell migration & & bright and large cells phenotype \\
+%     condensation followed by decondensation without completion of
+%     mitosis & & bright nuclei\\
+%     condensation without mitosis/collapse of nucleus & & cells with protrusions\\
+%     dynamic changes & & elongated cells\\
+%     failure in decondensation & & elongated cells with protrusions\\
+%     grape& & high actin ratio cells \\
+%     increased proliferation& & lamellipodia + high actin ratio cells\\
+%     large& & lamellipodia cells\\
+%     large nucleus& & large cells\\
+%     metaphase alignment problems/including no metaphase& & large nuclei\\
+%     metaphase delay/arrest& & low eccentricity cells\\
+%     migration (distance)& & metaphase cells\\
+%     migration (speed)& & proliferating cells\\
+%     mitotic delay/arrest& & small cells\\
+%     nuclei stay close together& & small cells with an enrichment of mitotic cells\\
+%     polylobed& &\\
+%     pulsating nuclei& &\\
+%     segregation problems/chromatin bridges/lagging chromosomes/multiple DNA masses&&\\
+%     small nucleus&&\\
+%     strange nuclear shape&&\\
+%     \hline
+%   \end{tabular}
+%   \caption{\label{tbl:studies}The table summarizes cellular phenotype
+%     terms used in three recent systems microscopy studies.}
+% \end{table*}
 
@@
 \section{Discussion}
-
 \subsection{Applications of the CPO}
@@
 
-\section{Acknowledgements}
+\begin{figure}
+  \centering
+  \includegraphics[width=1\textwidth]{onto.pdf}
+  \caption{Overview over basic top-level categories. Physical objects
+    are entities that are wholly present at time points. Qualities are
+    attributes of entities. Functions are capabilities that arise from
+    physical objects with their qualities, and processes are
+    temporally extended entities that may be realizations of a
+    function.\label{fig:onto}}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1\textwidth]{processpatterns.pdf}
+
+  \caption{\label{fig:patterns}Six examples of processes with
+    non-comparative and comparative process attributes.  We assume
+    that the processes labelled $a$, $b$, $c$ and $d$ are all
+    instances of the class of processes $P$.  On the left side, three
+    regulation (of $P$) processes are illustrated which exhibit
+    non-comparative attributes. The first process has an attribute of
+    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
+    temporally equidistantly distributed. The second example shows an
+    {\em arrhythmic} occurrence of $P$, and the third examples shows
+    an {\em increasing frequency} (of $P$). A regulation process with
+    an increasing frequency (of $P$) is a process in which the
+    frequency of occurrences of $P$ is lower in the first half of the
+    process than in the second half. The right side of the figure
+    illustrates comparative phenotypic descriptions of processes. On
+    the upper right, the {\em normal} reference is shown. The second
+    example illustrates a {\em late onset} of $P$, i.e., the attribute
+    that $P$ processes begin later than {\em normal}. Finally, the
+    lower right illustrates a {\em decreased frequency} (of $P$),
+    since fewer processes of the type $P$ occur within the reference
+    process than {\em normal}.}
+\end{figure}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
+  \caption{Overview over the taxonomic structure of CPO. The structure
+    is based on a cellular component class $X$ and the cellular
+    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
+\end{figure*}
+
+\section{Acknowledgement}
 Funding for RH is provided by the European Commission's 7th Framework
@@
 258068, EU-FP7-Systems Microscopy NoE.
-%GVG is funded by NIH grant number R01 HG004838-02.
 
-% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section*{Authors contributions}
-
-% % %%%%%%%%%%%%%%%%%%%%%%%%%%%
-% The authors declare that they have no competing interests.
-
-% % %%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section*{Acknowledgements}
-% \ifthenelse{\boolean{publ}}{\small}{}
-
-% The research work in its first unrevised form was presented at the
-% SMBM 2010, Hinxton, Cambridge, U.K.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The Bibliography %%
-%%                                                         %%
-%% Bmc_article.bst will be used to %%
-%% create a .BBL file for submission, which includes %%
-%% XML structured for BMC.  %%
-%%                                                         %%
-%%                                                         %%
-%% Note that the displayed Bibliography will not %%
-%% necessarily be rendered by Latex exactly as specified %%
-%% in the online Instructions for Authors.  %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\bibliographystyle{plain} % Style BST file
+\bibliographystyle{plain}
 \bibliography{lc}

### 2012-01-19 trunk/paper/drinking-sr.tex

--- 2012-01-19 trunk/paper/drinking-bioinformatics.tex
@@
-\documentclass{bioinfo}
-\copyrightyear{2005}
-\pubyear{2005}
+\documentclass[11pt]{article}
+%\usepackage{times}
+\usepackage{setspace}
+\usepackage{url}
+\usepackage[normalem]{ulem}
+\usepackage{multirow}
+%\copyrightyear{2005}
+%\pubyear{2005}
+\usepackage{graphicx}
 \usepackage{url}
@@
 \usepackage{pifont}
-%\newcommand{\tick}{\ding{52}}
-\newcommand{\tick}{\checkmark}
-\newcommand{\tickNo}{\hspace{1pt}\ding{55}}
+% Load packages
+\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
+\usepackage{url}  % Formatting web addresses  
+\usepackage{ifthen}  % Conditional 
+\usepackage[utf8]{inputenc} %unicode support
+\urlstyle{rm}
+\usepackage[sort&compress]{natbib}
+\renewcommand{\cite}{\citep}
+\usepackage{amsmath}
+\usepackage{amssymb}
 
 \begin{document}
-\firstpage{1}
 
-\title[Cell Phenotype Ontology]{Semantic integration of physiology
+\title{Semantic integration of physiology
   phenotypes: an application to the cellular phenotype ontology}
 
-\author[Hoehndorf \textit{et~al}]{Robert Hoehndorf$^{1}$\footnote{to
-    whom correspondence should be addressed}, Midori
-  A. Harris$^2$, Heinrich Herre$^3$, Gabriella Rustici$^4$ and
+\author{Robert Hoehndorf$^{1,*}$, Midori
+  A. Harris$^2$, Heinrich Herre$^3$, Gabriella Rustici$^4$ and\\
   Georgios V. Gkoutos$^{1}$}
 
-\address{$^{1}$Department of Genetics, University of Cambridge,
+\date{$^{1}$Department of Genetics, University of Cambridge,
   Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
@@
 
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+% \history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
 
-\editor{Associate Editor: XXXXXXX}
+% \editor{Associate Editor: XXXXXXX}
 
@@
 \begin{abstract}
-\section{Motivation:}
 The systematic observation of phenotypes has become a crucial tool of
@@
 challenge due to the high complexity of the underlying domain.
-
-\section{Results:}
 Here, we present the outline of a theory for an ontology of
@@
 no phenotype ontology has been created.  
-
-\section{Availability and implementation:} 
 The CPO and the source code we generated to create the CPO are freely
 available on \url{http://cell-phenotype.googlecode.com}.
-
-\section{Contact:} \href{rh497@cam.ac.uk}{rh497@cam.ac.uk}
 \end{abstract}
 
-\section{Introduction}
+%\section{Introduction}
 Phenotype studies on all scales and levels of granularity are now an
@@
 phenotypes in domains in which no such ontology exists.
-
-\begin{methods}
-\section{System and methods}
-\subsection{Formal ontology}
-The ontology as an approach to semantic standardisation was proposed
-more than a decade ago and since then has become the dominant
-methodology used to semantically categorise phenodeviance.  The
-biomedical research community has invested considerable effort and
-resources in the development and establishment of ontologies that are
-becoming increasingly successful as information management and
-integration tools in many disparate scientific fields allowing
-interoperability and semantic information processing between diverse
-biomedical resources and domains.
-
-In computer science, an ontology is a specification of a
-conceptualization of a domain of knowledge \cite{Gruber1995,
-  Guarino1998}.  Ontologies commonly distinguish between {\em classes}
-(also called {\em concepts}, {\em categories} or {\em universals}) and
-{\em individuals} within a domain of knowledge. A class is an entity
-that can have {\em instances}, while individuals are entities that
-cannot be instantiated \cite{Herre2006}. Examples of individuals
-include the Eiffel tower or the 2009 Ironman Triathlon in Hawaii,
-while examples of classes include {\em Tower} or {\em Triathlon}. The
-Eiffel tower can be an instance of the class {\em Tower}, and the 2009
-Ironman Triathlon an instance of {\em Triathlon}.  The meaning of
-classes is specified by stating what must be true of their instances.
-
-In addition to classes and individuals, ontologies often include {\em
-  relations}. Relations hold between entities, they are the ``the glue
-that holds things together, the primary constituents of the facts that
-go to make up reality'' \cite{Barwise1989}.
-
-In {\em formal} ontologies, the specification of classes and relations
-follows the axiomatic-deductive method. Given a set of terms that are
-used within a domain and whose meaning we wish to specify, we begin by
-providing {\em explicit definitions} for some terms, potentially
-introducing new terms. An explicit definition of a term $t$ is a
-statement that can replace every occurrence of $t$ in any sentence.
-
-Eventually, a set of {\em primitive terms} remains that are not
-further defined. Following the axiomatic method \cite{Hilbert1918},
-using only the primitive terms, we can construct complex
-sentences. Based on the intended meaning of the primitive terms, we
-consider some of these sentences true and some of them false in our
-domain. We select some of the true sentences as {\em axioms} which
-provide the core of our ontology. Ideally, the axioms are chosen so
-that all true sentences in the domain we intend to represent follow by
-means of logical deduction from the axioms. More commonly, however,
-only {\em some aspects} of the intended meaning are formally
-represented while other aspects are omitted either due to limitations
-in language expressivity or due to their irrelevance to the problem
-for which an ontology is developed.
-
-Based on the axioms and definitions, we can use deduction to infer
-statements that logically follow from the axioms.  The process of
-automatically deducing sentences from axioms is called {\em automated
-  reasoning}. Automated reasoning allows users of an ontology to carry
-out key activities: verifying the ontology's consistency, inferring
-hidden knowledge and thereby performing powerful queries.  An ontology
-is formally inconsistent if there is a statement $\phi$ such that
-$\phi$ and its negation $\neg \phi$ can be inferred from the
-ontology's axioms.  If an ontology is formally inconsistent, {\em
-  every} statement can be inferred from the ontology.
-
-Automated reasoning can further determine whether classes in an
-ontology are unsatisfiable: a class $C$ is unsatisfiable, if it is
-impossible for the class to have any instances. Unsatisfiable classes
-in an ontology are commonly the result of a contradictory class
-definition. % That a class is unsatisfiable is often a special kind of
-% unintended consequence that can be drawn from an ontology.
-
-Automated reasoning in the Web Ontology Language (OWL) can be employed
-to automatically compute the generalization hierarchy underlying an
-ontology as well as for verification of data consistency and complex
-queries \cite{Hoehndorf2011incon, Hoehndorf2011models}. Highly
-efficient automated reasoners are available to process OWL ontologies
-\cite{Sirin2004, Tsarkov2006, Motik2009a}. OWL profiles were developed
-to support even large ontologies by further reducing the expressivity
-of OWL in order to enable polynomial-time inferences. In particular
-the OWL EL profile was found to provide the expressivity required for
-most biomedical ontologies \cite{el4, elvira}, and highly optimized
-OWL EL reasoners are available or under development to support
-reasoning over very large ontologies \cite{el4, cbreasoner}.
-
-A high expressivity is required to accurately specify complex axioms
-that constrain the domain under investigation, and languages with
-higher expressivity than OWL are often required in the biomedical
-domain to achieve this goal \cite{Hoehndorf2009sequences, rnao}. On
-the other hand, automated reasoning over large ontologies and
-associated datasets benefits from languages with a low complexity of
-inferences in which complex axioms cannot be formulated. Therefore, a
-possible solution is to use a layered approach: to specify the meaning
-of terms using an expressive language, and derive the axioms that
-must obtain in a weaker language using deductive inferences.
-
-\subsection{Processes and their participants}
-Most biomedical ontologies share common distinctions between different
-kinds of entities: physical objects, qualities, functions and
-processes. A physical object is an entity that is present as a whole
-at a time point, i.e., an entity that has no temporal parts. A quality
-is an attribute or feature of an entity. Physical objects together
-with their qualities give rise to functions, which are capabilities or
-potentials of physical objects. These functions can then be realized
-by processes, which are temporally extended entities
-\cite{Burek2006}. Examples of classes that may have processes as
-instances include {\em Drinking}, {\em Triathlon} or {\em Apoptosis}.
-Figure \ref{fig:onto} illustrates these basic categories of being.
-
-Processes commonly have physical objects as participants. Physical
-objects are entities that are present as a whole at time points (i.e.,
-they have no temporal parts) and may persist through time, i.e., they
-may undergo changes during the process \cite{Herre2006, Herre2010}.
-We can further distinguish different {\em roles} of participation in a
-process \cite{Loebe2007}. For example, a runner may participate in a
-{\em Triathlon} process as a {\em Runner (role)}, while another person
-can participate in the same process but in a different role (such as
-{\em Referee}).
-
-% Amongst the participants of some processes, we will distinguish
-% between {\em inputs} and {\em outputs} of processes. Chemical
-% reactions, for example, are processes with definite inputs (the {\em
-%   reactants}) and outputs ({\em the products}).
-
-\subsection{Gene Ontology}
-The Gene Ontology provides a set of ontologies for molecular and
-cellular biology, originally designed to support structured
-annotations for genes and gene products in all species with respect to
-molecular function (MF), biological process (BP) and cellular
-component (CC). MF and BP both describe processes, but at different
-spatiotemporal scales; in particular, BP includes processes that
-unfold within cells and within tissues and organs of multicellular
-organisms. Gene product annotations identify participants in the
-processes.
-
-Over time, GO development has increasingly emphasized a normalized
-approach that includes supplementing existing human-readable text
-description with formally specified explicit definitions for GO
-classes. The formalization of GO is readily apparent in its
-representation of biological regulation.
-
-Regulatory processes may regulate other processes, at either the MF or
-BP scale, or biological qualities. GO accordingly includes three broad
-categories of regulation terms, regulation of molecular function,
-regulation of biological process, and regulation of biological
-quality. The first two are explicitly defined entirely with respect to
-other GO terms, whereas the third comprises classes in which the
-regulated qualities are specified by terms from PATO (see below) or
-anatomy ontologies.
-
-All GO regulation terms use one of three relations, {\bf regulates},
-{\bf negatively\_regulates} and {\bf positively\_regulates}, to link
-regulation terms to process or quality terms. The {\bf regulates}
-relations are defined in terms of qualities: a regulatory process
-causes a change in magnitude to some quality, which in turn has an
-effect on the frequency, rate or duration of some other type of
-process. Effects that results in increases and decreases use {\bf
-  positively\_regulates} and {\bf negatively\_regulates} respectively
-\cite{Mungall2010go}. The existing ontology structure would also
-support the addition of subclasses to distinguish, for example,
-regulation of the rate of a process from regulation of its duration or
-time of onset.
-
-\subsection{PATO and the EQ model}
-PATO was envisaged and designed to provide a platform for allowing the
-integration of quantitative and qualitative phenotype related
-information across different domains, levels of granularity and
-species \cite{Gkoutos2005}.  PATO is an ontology of phenotype
-qualities that form basic entities that we can perceive and/or measure
-such as colors, sizes, rates etc. One of its classification axes is
-based on the basic type of entity to which a qualities belongs, and
-PATO distinguishes between qualities of physical objects and qualities
-of processes.
-
-PATO allows for the description of affected entities by combining
-various ontologies that describe the entities that have been affected,
-such as the various anatomical ontologies, GO \cite {Ashburner2000b},
-the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
-with the various qualities it provides for defining how these entities
-were affected.  PATO can be used for annotation either directly in a
-so called post-composed (post-coordinated) manner or for providing
-formal (logical) definitions (equivalence axioms) to ontologies
-containing a set of precomposed (pre-coordinated) phenotype terms. For
-instance, to describe the decrease in the length of the sexual cycle
-of female animals, we can combine the PATO term \textit{decreased
-  duration} ({\tt PATO:0000499}) with the Gene Ontology term
-\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
-existed in a pre-composed ontology (for example the {\tt MP:0009007}
-term from the Mammalian Phenotype) it could be used to provide an
-equivalence statement between that class and the above PATO-based
-description.
-
-\end{methods}
 
@@
 
-\begin{table*}
-  \centering
-  \begin{tabular}{p{2.8cm}|p{4cm}|p{4cm}|p{4cm}}
-    & Increased flow rate & Normal flow rate & Decreased
-    flow rate \\
-    \hline
-    Increased frequency &increased total flow rate &increased total
-    flow rate &?\\
-    Normal frequency &increased total flow rate &normal total flow
-    rate &decreased total flow rate\\
-    Decreased frequency &?&decreased total flow rate &decreased total
-    flow rate\\
-    \hline
-  \end{tabular}
-  \caption{\label{tbl:flow}Interdependency for the attribute {\em
-      Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
-      rate} is an attribute of {\em Regulation of cytoplasmic
-      streaming} processes, while {\em Cytoplasmic flow rate} is an
-    attribute of individual {\em cytoplasmic streaming}
-    processes. Depending both on whether the cytoplasmic flow rate in
-    individual {\em cytoplasmic streaming} processes is increased or
-    decreased and whether the frequency of occurrence of {\em
-      cytoplasmic streaming} is increased or decreased, the total
-    cytoplasmic flow rate can be increased or decreased.}
-\end{table*}
-
 A first kind of abnormality of regulatory processes are {\em abnormal
@@
 \section{Discussion}
-\subsection{Applications of the CPO}
+%\subsection{Applications of the CPO}
 The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 A further domain that will greatly benefit from the CPO is {\em
@@
   Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
-three studies illustrated in Table \ref{tbl:studies}, live-cell
-imaging assays and RNAi knockdown were used to generate phenotypic
-profiles that quantify the cellular response to a given siRNA thus
-allowing identification of hundreds of genes involved in diverse
-biological functions including cell division, migration and survival.
-In each study, several phenotypes were detected and described by the
-authors without the use of ontologies (see Table \ref{tbl:studies}),
-making the integration between datasets extremely difficult. For
-example, it is evident that cell division phenotypes were observed in
-all three datasets (e.g. {\em Mitotic delay/arrest} and {\em Prolonged
-  mitotic exit} or {\em Methaphase delay} and {\em Methaphase cells})
-but the overlap between such phenotypes is unclear.
+three studies, live-cell imaging assays and RNAi knockdown were used
+to generate phenotypic profiles that quantify the cellular response to
+a given siRNA thus allowing identification of hundreds of genes
+involved in diverse biological functions including cell division,
+migration and survival.  In each study, several phenotypes were
+detected and described by the authors without the use of ontologies or
+controlled vocabularies, making the integration between datasets
+extremely difficult. For example, it is evident that cell division
+phenotypes were observed in all three datasets and describes using
+terms such as {\em Mitotic delay/arrest}, {\em Prolonged mitotic
+  exit}, {\em Methaphase delay} and {\em Methaphase cells}). the
+overlap between such phenotypes is unclear.
 
@@
 Systems Microscopy Network of Excellence
-(http://www.systemsmicroscopy.eu/) and the first step towards data
-integration can be achieved by further developing CPO.
+(\url{http://www.systemsmicroscopy.eu/}) and the first step towards
+data integration can be achieved by further developing CPO.
 
@@
 
-\subsection{Future research}
+%\subsection{Future research}
 We implemented the CPO using a pattern-based approach to formulating
@@
 
-\begin{figure}
+\section{Methods}
+\subsection{Formal ontology}
+The ontology as an approach to semantic standardisation was proposed
+more than a decade ago and since then has become the dominant
+methodology used to semantically categorise phenodeviance.  The
+biomedical research community has invested considerable effort and
+resources in the development and establishment of ontologies that are
+becoming increasingly successful as information management and
+integration tools in many disparate scientific fields allowing
+interoperability and semantic information processing between diverse
+biomedical resources and domains.
+
+In computer science, an ontology is a specification of a
+conceptualization of a domain of knowledge \cite{Gruber1995,
+  Guarino1998}.  Ontologies commonly distinguish between {\em classes}
+(also called {\em concepts}, {\em categories} or {\em universals}) and
+{\em individuals} within a domain of knowledge. A class is an entity
+that can have {\em instances}, while individuals are entities that
+cannot be instantiated \cite{Herre2006}. Examples of individuals
+include the Eiffel tower or the 2009 Ironman Triathlon in Hawaii,
+while examples of classes include {\em Tower} or {\em Triathlon}. The
+Eiffel tower can be an instance of the class {\em Tower}, and the 2009
+Ironman Triathlon an instance of {\em Triathlon}.  The meaning of
+classes is specified by stating what must be true of their instances.
+
+In addition to classes and individuals, ontologies often include {\em
+  relations}. Relations hold between entities, they are the ``the glue
+that holds things together, the primary constituents of the facts that
+go to make up reality'' \cite{Barwise1989}.
+
+In {\em formal} ontologies, the specification of classes and relations
+follows the axiomatic-deductive method. Given a set of terms that are
+used within a domain and whose meaning we wish to specify, we begin by
+providing {\em explicit definitions} for some terms, potentially
+introducing new terms. An explicit definition of a term $t$ is a
+statement that can replace every occurrence of $t$ in any sentence.
+
+Eventually, a set of {\em primitive terms} remains that are not
+further defined. Following the axiomatic method \cite{Hilbert1918},
+using only the primitive terms, we can construct complex
+sentences. Based on the intended meaning of the primitive terms, we
+consider some of these sentences true and some of them false in our
+domain. We select some of the true sentences as {\em axioms} which
+provide the core of our ontology. Ideally, the axioms are chosen so
+that all true sentences in the domain we intend to represent follow by
+means of logical deduction from the axioms. More commonly, however,
+only {\em some aspects} of the intended meaning are formally
+represented while other aspects are omitted either due to limitations
+in language expressivity or due to their irrelevance to the problem
+for which an ontology is developed.
+
+Based on the axioms and definitions, we can use deduction to infer
+statements that logically follow from the axioms.  The process of
+automatically deducing sentences from axioms is called {\em automated
+  reasoning}. Automated reasoning allows users of an ontology to carry
+out key activities: verifying the ontology's consistency, inferring
+hidden knowledge and thereby performing powerful queries.  An ontology
+is formally inconsistent if there is a statement $\phi$ such that
+$\phi$ and its negation $\neg \phi$ can be inferred from the
+ontology's axioms.  If an ontology is formally inconsistent, {\em
+  every} statement can be inferred from the ontology.
+
+Automated reasoning can further determine whether classes in an
+ontology are unsatisfiable: a class $C$ is unsatisfiable, if it is
+impossible for the class to have any instances. Unsatisfiable classes
+in an ontology are commonly the result of a contradictory class
+definition. % That a class is unsatisfiable is often a special kind of
+% unintended consequence that can be drawn from an ontology.
+
+Automated reasoning in the Web Ontology Language (OWL) can be employed
+to automatically compute the generalization hierarchy underlying an
+ontology as well as for verification of data consistency and complex
+queries \cite{Hoehndorf2011incon, Hoehndorf2011models}. Highly
+efficient automated reasoners are available to process OWL ontologies
+\cite{Sirin2004, Tsarkov2006, Motik2009a}. OWL profiles were developed
+to support even large ontologies by further reducing the expressivity
+of OWL in order to enable polynomial-time inferences. In particular
+the OWL EL profile was found to provide the expressivity required for
+most biomedical ontologies \cite{el4, elvira}, and highly optimized
+OWL EL reasoners are available or under development to support
+reasoning over very large ontologies \cite{el4, cbreasoner}.
+
+A high expressivity is required to accurately specify complex axioms
+that constrain the domain under investigation, and languages with
+higher expressivity than OWL are often required in the biomedical
+domain to achieve this goal \cite{Hoehndorf2009sequences, rnao}. On
+the other hand, automated reasoning over large ontologies and
+associated datasets benefits from languages with a low complexity of
+inferences in which complex axioms cannot be formulated. Therefore, a
+possible solution is to use a layered approach: to specify the meaning
+of terms using an expressive language, and derive the axioms that
+must obtain in a weaker language using deductive inferences.
+
+\subsection{Processes and their participants}
+Most biomedical ontologies share common distinctions between different
+kinds of entities: physical objects, qualities, functions and
+processes. A physical object is an entity that is present as a whole
+at a time point, i.e., an entity that has no temporal parts. A quality
+is an attribute or feature of an entity. Physical objects together
+with their qualities give rise to functions, which are capabilities or
+potentials of physical objects. These functions can then be realized
+by processes, which are temporally extended entities
+\cite{Burek2006}. Examples of classes that may have processes as
+instances include {\em Drinking}, {\em Triathlon} or {\em Apoptosis}.
+Figure \ref{fig:onto} illustrates these basic categories of being.
+
+Processes commonly have physical objects as participants. Physical
+objects are entities that are present as a whole at time points (i.e.,
+they have no temporal parts) and may persist through time, i.e., they
+may undergo changes during the process \cite{Herre2006, Herre2010}.
+We can further distinguish different {\em roles} of participation in a
+process \cite{Loebe2007}. For example, a runner may participate in a
+{\em Triathlon} process as a {\em Runner (role)}, while another person
+can participate in the same process but in a different role (such as
+{\em Referee}).
+
+% Amongst the participants of some processes, we will distinguish
+% between {\em inputs} and {\em outputs} of processes. Chemical
+% reactions, for example, are processes with definite inputs (the {\em
+%   reactants}) and outputs ({\em the products}).
+
+\subsection{Gene Ontology}
+The Gene Ontology provides a set of ontologies for molecular and
+cellular biology, originally designed to support structured
+annotations for genes and gene products in all species with respect to
+molecular function (MF), biological process (BP) and cellular
+component (CC). MF and BP both describe processes, but at different
+spatiotemporal scales; in particular, BP includes processes that
+unfold within cells and within tissues and organs of multicellular
+organisms. Gene product annotations identify participants in the
+processes.
+
+Over time, GO development has increasingly emphasized a normalized
+approach that includes supplementing existing human-readable text
+description with formally specified explicit definitions for GO
+classes. The formalization of GO is readily apparent in its
+representation of biological regulation.
+
+Regulatory processes may regulate other processes, at either the MF or
+BP scale, or biological qualities. GO accordingly includes three broad
+categories of regulation terms, regulation of molecular function,
+regulation of biological process, and regulation of biological
+quality. The first two are explicitly defined entirely with respect to
+other GO terms, whereas the third comprises classes in which the
+regulated qualities are specified by terms from PATO (see below) or
+anatomy ontologies.
+
+All GO regulation terms use one of three relations, {\bf regulates},
+{\bf negatively\_regulates} and {\bf positively\_regulates}, to link
+regulation terms to process or quality terms. The {\bf regulates}
+relations are defined in terms of qualities: a regulatory process
+causes a change in magnitude to some quality, which in turn has an
+effect on the frequency, rate or duration of some other type of
+process. Effects that results in increases and decreases use {\bf
+  positively\_regulates} and {\bf negatively\_regulates} respectively
+\cite{Mungall2010go}. The existing ontology structure would also
+support the addition of subclasses to distinguish, for example,
+regulation of the rate of a process from regulation of its duration or
+time of onset.
+
+\subsection{PATO and the EQ model}
+PATO was envisaged and designed to provide a platform for allowing the
+integration of quantitative and qualitative phenotype related
+information across different domains, levels of granularity and
+species \cite{Gkoutos2005}.  PATO is an ontology of phenotype
+qualities that form basic entities that we can perceive and/or measure
+such as colors, sizes, rates etc. One of its classification axes is
+based on the basic type of entity to which a qualities belongs, and
+PATO distinguishes between qualities of physical objects and qualities
+of processes.
+
+PATO allows for the description of affected entities by combining
+various ontologies that describe the entities that have been affected,
+such as the various anatomical ontologies, GO \cite {Ashburner2000b},
+the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
+with the various qualities it provides for defining how these entities
+were affected.  PATO can be used for annotation either directly in a
+so called post-composed (post-coordinated) manner or for providing
+formal (logical) definitions (equivalence axioms) to ontologies
+containing a set of precomposed (pre-coordinated) phenotype terms. For
+instance, to describe the decrease in the length of the sexual cycle
+of female animals, we can combine the PATO term \textit{decreased
+  duration} ({\tt PATO:0000499}) with the Gene Ontology term
+\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
+existed in a pre-composed ontology (for example the {\tt MP:0009007}
+term from the Mammalian Phenotype) it could be used to provide an
+equivalence statement between that class and the above PATO-based
+description.
+
+\bibliographystyle{unsrt}
+\bibliography{lc}
+
+\clearpage
+\section*{Acknowledgement}
+Funding for RH is provided by the European Commission's 7th Framework
+Programme, RICORDO project, grant number 248502. MAH is supported by
+Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
+Medical Informatics, Statistics and Epidemiology, University of
+Leipzig.  Funding for GR is provided by the European Union's Seventh
+Framework Programme (FP7/2007-2013) under grant agreement number
+258068, EU-FP7-Systems Microscopy NoE.
+
+\section*{Author contributions}
+...
+
+\section*{Additional information}
+\paragraph*{Competing financial interests:} The author(s) declare no
+competing financial interests.
+
+\clearpage
+\section*{Figure legends}
+\begin{figure}[h]
   \centering
-  \includegraphics[width=1\textwidth]{onto.pdf}
+%  \includegraphics[width=1\textwidth]{onto.pdf}
   \caption{Overview over basic top-level categories. Physical objects
@@
 
-\begin{figure}
+\begin{figure}[h]
   \centering
-  \includegraphics[width=1\textwidth]{processpatterns.pdf}
+ % \includegraphics[width=1\textwidth]{processpatterns.pdf}
 
@@
 
-\begin{figure*}
+\begin{figure*}[h]
   \centering
-  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
+%  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
   \caption{Overview over the taxonomic structure of CPO. The structure
@@
 
-\section{Acknowledgement}
-Funding for RH is provided by the European Commission's 7th Framework
-Programme, RICORDO project, grant number 248502. MAH is supported by
-Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
-Medical Informatics, Statistics and Epidemiology, University of
-Leipzig.  Funding for GR is provided by the European Union's Seventh
-Framework Programme (FP7/2007-2013) under grant agreement number
-258068, EU-FP7-Systems Microscopy NoE.
-
-\bibliographystyle{plain}
-\bibliography{lc}
+\clearpage
+\section*{Tables}
+\begin{table*}[h]
+  \centering
+  \begin{tabular}{p{2.5cm}|p{3cm}|p{3cm}|p{3cm}}
+    & Increased cytoplasmic flow rate & Normal cytoplasmic flow rate &
+    Decreased cytoplasmic flow rate \\
+    \hline
+    Increased frequency &increased total flow rate &increased total
+    flow rate &?\\
+    Normal frequency &increased total flow rate &normal total flow
+    rate &decreased total flow rate\\
+    Decreased frequency &?&decreased total flow rate &decreased total
+    flow rate\\
+    \hline
+  \end{tabular}
+  \caption{\label{tbl:flow}Interdependency for the attribute {\em
+      Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
+      rate} is an attribute of {\em Regulation of cytoplasmic
+      streaming} processes, while {\em Cytoplasmic flow rate} is an
+    attribute of individual {\em cytoplasmic streaming}
+    processes. Depending both on whether the cytoplasmic flow rate in
+    individual {\em cytoplasmic streaming} processes is increased or
+    decreased and whether the frequency of occurrence of {\em
+      cytoplasmic streaming} is increased or decreased, the total
+    cytoplasmic flow rate can be increased or decreased.}
+\end{table*}
 

### 2012-01-19 paper/drinking-with-figs.tex

--- 2011-12-01 paper/drinking.tex
@@
-\documentclass{bioinfo}
+\documentclass[12pt]{article}
+\usepackage{times}
 \usepackage{setspace}
+\doublespacing
 
-\copyrightyear{2005}
-\pubyear{2005}
+%\copyrightyear{2005}
+%\pubyear{2005}
+\usepackage{graphicx}
 \usepackage{url}
@@
 \begin{document}
-\firstpage{1}
+%\firstpage{1}
 
@@
 \author{Robert Hoehndorf$^1$ and Midori
-  A. Harris$^2$ and Heinrich Herre$^3$ and Gabriella Rustici$^4$ and
+  A. Harris$^2$ and\\Heinrich Herre$^3$ and Gabriella Rustici$^4$ and
   Georgios V. Gkoutos$^{1}$}
 
-\address{$^{1}$Department of Genetics, University of Cambridge,
+\date{$^{1}$Department of Genetics, University of Cambridge,
   Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
@@
   \centering
-  \begin{tabular}{l|l|l|l}
+  \begin{tabular}{p{2cm}|p{4cm}|p{4cm}|p{4cm}}
     & Increased cytoplasmic flow rate & Normal cytoplasmic flow rate & Decreased
@@
 
-\bibliographystyle{plain} % Style BST file
+\bibliographystyle{natbib} % Style BST file
 \bibliography{lc}
 
+\clearpage
+
+\section*{Figure legends}
+\begin{figure}
+  \centering
+  \includegraphics[width=1\textwidth]{onto.pdf}
+  \caption{Overview over basic top-level categories. Physical objects
+    are entities that are wholly present at time points. Qualities are
+    attributes of entities. Functions are capabilities that arise from
+    physical objects with their qualities, and processes are
+    temporally extended entities that may be realizations of a
+    function.\label{fig:onto}}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1\textwidth]{processpatterns.pdf}
+
+  \caption{\label{fig:patterns}Six examples of processes with
+    non-comparative and comparative process attributes.  We assume
+    that the processes labelled $a$, $b$, $c$ and $d$ are all
+    instances of the class of processes $P$.  On the left side, three
+    regulation (of $P$) processes are illustrated which exhibit
+    non-comparative attributes. The first process has an attribute of
+    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
+    temporally equidistantly distributed. The second example shows an
+    {\em arrhythmic} occurrence of $P$, and the third examples shows
+    an {\em increasing frequency} (of $P$). A regulation process with
+    an increasing frequency (of $P$) is a process in which the
+    frequency of occurrences of $P$ is lower in the first half of the
+    process than in the second half. The right side of the figure
+    illustrates comparative phenotypic descriptions of processes. On
+    the upper right, the {\em normal} reference is shown. The second
+    example illustrates a {\em late onset} of $P$, i.e., the attribute
+    that $P$ processes begin later than {\em normal}. Finally, the
+    lower right illustrates a {\em decreased frequency} (of $P$),
+    since fewer processes of the type $P$ occur within the reference
+    process than {\em normal}.}
+\end{figure}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
+  \caption{Overview over the taxonomic structure of CPO. The structure
+    is based on a cellular component class $X$ and the cellular
+    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
+\end{figure*}
+
 \end{document}

### 2012-01-20 trunk/paper/drinking-bioinformatics.tex

@@
 
-\title[Cell Phenotype Ontology]{Semantic integration of physiology
-  phenotypes: an application to the cellular phenotype ontology}
+\title[Physiology phenotypes]{Semantic integration of physiology
+  phenotypes with an application to the Cellular Phenotype Ontology}
 
@@
 \section{Results:}
-Here, we present the outline of a theory for an ontology of
-physiology-related phenotypes. We provide a formal description of
-process attributes and relate them to the attributes of their parts
-and participants. We apply our theory to create the Cellular Phenotype
-Ontology (CPO). The CPO is an ontology of morphological and
-physiological phenotypic abnormalities of cells, cell components and
-cellular processes. Its prime application is the unification of
-cellular phenotype descriptions across species by providing terms and
-uniform definition patterns. The CPO can further be used for the
-annotation of observed abnormalities in domains, such as systems
-microscopy, in which cellular abnormalities are observed and for which
-no phenotype ontology has been created.
+Here, we present the outline of a theory and its implementation for an
+ontology of physiology-related phenotypes. We provide a formal
+description of process attributes and relate them to the attributes of
+their temporal parts and participants. We apply our theory to create
+the Cellular Phenotype Ontology (CPO). The CPO is an ontology of
+morphological and physiological phenotypic characteristics of cells,
+cell components and cellular processes. Its prime application is the
+unification of cellular phenotype descriptions across species by
+providing terms and uniform definition patterns. The CPO can further
+be used for the annotation of observed abnormalities in domains, such
+as systems microscopy, in which cellular abnormalities are observed
+and for which no phenotype ontology has been created.
 
@@
 integrate phenotypes across species and reuse the content of anatomy
-and process ontologies, classes in phenotype ontologies were defined
-using the framework of the Phenotypic Attribute and Trait (PATO)
-ontology \cite{Gkoutos2005}. According to the PATO framework, a
-phenotype can be decomposed, using an Entity-Quality model, into an
-affected entity and a quality that characterizes {\em how} the entity
-is affected \cite{Gkoutos2005}. Such decompositions have been created
-for several widely used phenotype ontologies \cite{Mungall2010,
-  Gkoutos2009b, obml2011h1}, and are being applied together with
-methods for reusing knowledge contained in anatomy ontologies
-\cite{Mungall2010, Hoehndorf2010phene}.
+and process ontologies, we have defined process classes using the
+framework of the Phenotypic Attribute and Trait (PATO) ontology
+\cite{Gkoutos2005}. According to the PATO framework, a phenotype can
+be decomposed, using an Entity-Quality model, into an affected entity
+and a quality that characterizes {\em how} the entity is affected
+\cite{Gkoutos2005}. Such decompositions have been created for several
+widely used phenotype ontologies \cite{Mungall2010, Gkoutos2009b,
+  obml2011h1}, and are being applied together with methods for reusing
+knowledge contained in anatomy ontologies \cite{Mungall2010,
+  Hoehndorf2010phene}.
 
@@
 are impaired continues to limit the interoperability between phenotype
-ontologies. Major challenges for representing and integrating
-process-based phenotypes include establishing the link to the
-components of biological systems that have the capabilities to exhibit
-such a behaviour, and that attributes of processes are often measured
-{\em indirectly} and inferred from other attributes.
+ontologies. Major challenges for representing and integrating process
+phenotypes include establishing the link to the components of
+biological systems that have the capabilities to exhibit such a
+behaviour, and that attributes of processes are often measured {\em
+  indirectly} and inferred from other attributes.
 
@@
 \section{System and methods}
-\subsection{Formal ontology}
-The ontology as an approach to semantic standardisation was proposed
-more than a decade ago and since then has become the dominant
-methodology used to semantically categorise phenodeviance.  The
-biomedical research community has invested considerable effort and
-resources in the development and establishment of ontologies that are
-becoming increasingly successful as information management and
-integration tools in many disparate scientific fields allowing
-interoperability and semantic information processing between diverse
-biomedical resources and domains.
+% \subsection{Formal ontology}
+% The ontology as an approach to semantic standardisation was proposed
+% more than a decade ago and since then has become the dominant
+% methodology used to semantically categorise phenodeviance.  The
+% biomedical research community has invested considerable effort and
+% resources in the development and establishment of ontologies that are
+% becoming increasingly successful as information management and
+% integration tools in many disparate scientific fields allowing
+% interoperability and semantic information processing between diverse
+% biomedical resources and domains.
 
-In computer science, an ontology is a specification of a
-conceptualization of a domain of knowledge \cite{Gruber1995,
-  Guarino1998}.  Ontologies commonly distinguish between {\em classes}
-(also called {\em concepts}, {\em categories} or {\em universals}) and
-{\em individuals} within a domain of knowledge. A class is an entity
-that can have {\em instances}, while individuals are entities that
-cannot be instantiated \cite{Herre2006}. Examples of individuals
-include the Eiffel tower or the 2009 Ironman Triathlon in Hawaii,
-while examples of classes include {\em Tower} or {\em Triathlon}. The
-Eiffel tower can be an instance of the class {\em Tower}, and the 2009
-Ironman Triathlon an instance of {\em Triathlon}.  The meaning of
-classes is specified by stating what must be true of their instances.
+% In computer science, an ontology is a specification of a
+% conceptualization of a domain of knowledge \cite{Gruber1995,
+%   Guarino1998}.  Ontologies commonly distinguish between {\em classes}
+% (also called {\em concepts}, {\em categories} or {\em universals}) and
+% {\em individuals} within a domain of knowledge. A class is an entity
+% that can have {\em instances}, while individuals are entities that
+% cannot be instantiated \cite{Herre2006}. Examples of individuals
+% include the Eiffel tower or the 2009 Ironman Triathlon in Hawaii,
+% while examples of classes include {\em Tower} or {\em Triathlon}. The
+% Eiffel tower can be an instance of the class {\em Tower}, and the 2009
+% Ironman Triathlon an instance of {\em Triathlon}.  The meaning of
+% classes is specified by stating what must be true of their instances.
 
-In addition to classes and individuals, ontologies often include {\em
-  relations}. Relations hold between entities, they are the ``the glue
-that holds things together, the primary constituents of the facts that
-go to make up reality'' \cite{Barwise1989}.
+% In addition to classes and individuals, ontologies often include {\em
+%   relations}. Relations hold between entities, they are the ``the glue
+% that holds things together, the primary constituents of the facts that
+% go to make up reality'' \cite{Barwise1989}.
 
-In {\em formal} ontologies, the specification of classes and relations
-follows the axiomatic-deductive method. Given a set of terms that are
-used within a domain and whose meaning we wish to specify, we begin by
-providing {\em explicit definitions} for some terms, potentially
-introducing new terms. An explicit definition of a term $t$ is a
-statement that can replace every occurrence of $t$ in any sentence.
+% In {\em formal} ontologies, the specification of classes and relations
+% follows the axiomatic-deductive method. Given a set of terms that are
+% used within a domain and whose meaning we wish to specify, we begin by
+% providing {\em explicit definitions} for some terms, potentially
+% introducing new terms. An explicit definition of a term $t$ is a
+% statement that can replace every occurrence of $t$ in any sentence.
 
-Eventually, a set of {\em primitive terms} remains that are not
-further defined. Following the axiomatic method \cite{Hilbert1918},
-using only the primitive terms, we can construct complex
-sentences. Based on the intended meaning of the primitive terms, we
-consider some of these sentences true and some of them false in our
-domain. We select some of the true sentences as {\em axioms} which
-provide the core of our ontology. Ideally, the axioms are chosen so
-that all true sentences in the domain we intend to represent follow by
-means of logical deduction from the axioms. More commonly, however,
-only {\em some aspects} of the intended meaning are formally
-represented while other aspects are omitted either due to limitations
-in language expressivity or due to their irrelevance to the problem
-for which an ontology is developed.
+% Eventually, a set of {\em primitive terms} remains that are not
+% further defined. Following the axiomatic method \cite{Hilbert1918},
+% using only the primitive terms, we can construct complex
+% sentences. Based on the intended meaning of the primitive terms, we
+% consider some of these sentences true and some of them false in our
+% domain. We select some of the true sentences as {\em axioms} which
+% provide the core of our ontology. Ideally, the axioms are chosen so
+% that all true sentences in the domain we intend to represent follow by
+% means of logical deduction from the axioms. More commonly, however,
+% only {\em some aspects} of the intended meaning are formally
+% represented while other aspects are omitted either due to limitations
+% in language expressivity or due to their irrelevance to the problem
+% for which an ontology is developed.
 
-Based on the axioms and definitions, we can use deduction to infer
-statements that logically follow from the axioms.  The process of
-automatically deducing sentences from axioms is called {\em automated
-  reasoning}. Automated reasoning allows users of an ontology to carry
-out key activities: verifying the ontology's consistency, inferring
-hidden knowledge and thereby performing powerful queries.  An ontology
-is formally inconsistent if there is a statement $\phi$ such that
-$\phi$ and its negation $\neg \phi$ can be inferred from the
-ontology's axioms.  If an ontology is formally inconsistent, {\em
-  every} statement can be inferred from the ontology.
+% Based on the axioms and definitions, we can use deduction to infer
+% statements that logically follow from the axioms.  The process of
+% automatically deducing sentences from axioms is called {\em automated
+%   reasoning}. Automated reasoning allows users of an ontology to carry
+% out key activities: verifying the ontology's consistency, inferring
+% hidden knowledge and thereby performing powerful queries.  An ontology
+% is formally inconsistent if there is a statement $\phi$ such that
+% $\phi$ and its negation $\neg \phi$ can be inferred from the
+% ontology's axioms.  If an ontology is formally inconsistent, {\em
+%   every} statement can be inferred from the ontology.
 
-Automated reasoning can further determine whether classes in an
-ontology are unsatisfiable: a class $C$ is unsatisfiable, if it is
-impossible for the class to have any instances. Unsatisfiable classes
-in an ontology are commonly the result of a contradictory class
-definition. % That a class is unsatisfiable is often a special kind of
-% unintended consequence that can be drawn from an ontology.
+% Automated reasoning can further determine whether classes in an
+% ontology are unsatisfiable: a class $C$ is unsatisfiable, if it is
+% impossible for the class to have any instances. Unsatisfiable classes
+% in an ontology are commonly the result of a contradictory class
+% definition. % That a class is unsatisfiable is often a special kind of
+% % unintended consequence that can be drawn from an ontology.
 
-Automated reasoning in the Web Ontology Language (OWL) can be employed
-to automatically compute the generalization hierarchy underlying an
-ontology as well as for verification of data consistency and complex
-queries \cite{Hoehndorf2011incon, Hoehndorf2011models}. Highly
-efficient automated reasoners are available to process OWL ontologies
-\cite{Sirin2004, Tsarkov2006, Motik2009a}. OWL profiles were developed
-to support even large ontologies by further reducing the expressivity
-of OWL in order to enable polynomial-time inferences. In particular
-the OWL EL profile was found to provide the expressivity required for
-most biomedical ontologies \cite{el4, elvira}, and highly optimized
-OWL EL reasoners are available or under development to support
-reasoning over very large ontologies \cite{el4, cbreasoner}.
+% Automated reasoning in the Web Ontology Language (OWL) can be employed
+% to automatically compute the generalization hierarchy underlying an
+% ontology as well as for verification of data consistency and complex
+% queries \cite{Hoehndorf2011incon, Hoehndorf2011models}. Highly
+% efficient automated reasoners are available to process OWL ontologies
+% \cite{Sirin2004, Tsarkov2006, Motik2009a}. OWL profiles were developed
+% to support even large ontologies by further reducing the expressivity
+% of OWL in order to enable polynomial-time inferences. In particular
+% the OWL EL profile was found to provide the expressivity required for
+% most biomedical ontologies \cite{el4, elvira}, and highly optimized
+% OWL EL reasoners are available or under development to support
+% reasoning over very large ontologies \cite{el4, cbreasoner}.
 
-A high expressivity is required to accurately specify complex axioms
-that constrain the domain under investigation, and languages with
-higher expressivity than OWL are often required in the biomedical
-domain to achieve this goal \cite{Hoehndorf2009sequences, rnao}. On
-the other hand, automated reasoning over large ontologies and
-associated datasets benefits from languages with a low complexity of
-inferences in which complex axioms cannot be formulated. Therefore, a
-possible solution is to use a layered approach: to specify the meaning
-of terms using an expressive language, and derive the axioms that
-must obtain in a weaker language using deductive inferences.
+% A high expressivity is required to accurately specify complex axioms
+% that constrain the domain under investigation, and languages with
+% higher expressivity than OWL are often required in the biomedical
+% domain to achieve this goal \cite{Hoehndorf2009sequences, rnao}. On
+% the other hand, automated reasoning over large ontologies and
+% associated datasets benefits from languages with a low complexity of
+% inferences in which complex axioms cannot be formulated. Therefore, a
+% possible solution is to use a layered approach: to specify the meaning
+% of terms using an expressive language, and derive the axioms that
+% must obtain in a weaker language using deductive inferences.
 
-\subsection{Processes and their participants}
-Most biomedical ontologies share common distinctions between different
-kinds of entities: physical objects, qualities, functions and
-processes. A physical object is an entity that is present as a whole
-at a time point, i.e., an entity that has no temporal parts. A quality
-is an attribute or feature of an entity. Physical objects together
-with their qualities give rise to functions, which are capabilities or
-potentials of physical objects. These functions can then be realized
-by processes, which are temporally extended entities
-\cite{Burek2006}. Examples of classes that may have processes as
-instances include {\em Drinking}, {\em Triathlon} or {\em Apoptosis}.
-Figure \ref{fig:onto} illustrates these basic categories of being.
+% \subsection{Processes and their participants}
+% Most biomedical ontologies share common distinctions between different
+% kinds of entities: physical objects, qualities, functions and
+% processes. A physical object is an entity that is present as a whole
+% at a time point, i.e., an entity that has no temporal parts. A quality
+% is an attribute or feature of an entity. Physical objects together
+% with their qualities give rise to functions, which are capabilities or
+% potentials of physical objects. These functions can then be realized
+% by processes, which are temporally extended entities
+% \cite{Burek2006}. Examples of classes that may have processes as
+% instances include {\em Drinking}, {\em Triathlon} or {\em Apoptosis}.
+% Figure \ref{fig:onto} illustrates these basic categories of being.
 
-Processes commonly have physical objects as participants. Physical
-objects are entities that are present as a whole at time points (i.e.,
-they have no temporal parts) and may persist through time, i.e., they
-may undergo changes during the process \cite{Herre2006, Herre2010}.
-We can further distinguish different {\em roles} of participation in a
-process \cite{Loebe2007}. For example, a runner may participate in a
-{\em Triathlon} process as a {\em Runner (role)}, while another person
-can participate in the same process but in a different role (such as
-{\em Referee}).
+% Processes commonly have physical objects as participants. Physical
+% objects are entities that are present as a whole at time points (i.e.,
+% they have no temporal parts) and may persist through time, i.e., they
+% may undergo changes during the process \cite{Herre2006, Herre2010}.
+% We can further distinguish different {\em roles} of participation in a
+% process \cite{Loebe2007}. For example, a runner may participate in a
+% {\em Triathlon} process as a {\em Runner (role)}, while another person
+% can participate in the same process but in a different role (such as
+% {\em Referee}).
 
-% Amongst the participants of some processes, we will distinguish
-% between {\em inputs} and {\em outputs} of processes. Chemical
-% reactions, for example, are processes with definite inputs (the {\em
-%   reactants}) and outputs ({\em the products}).
+% % Amongst the participants of some processes, we will distinguish
+% % between {\em inputs} and {\em outputs} of processes. Chemical
+% % reactions, for example, are processes with definite inputs (the {\em
+% %   reactants}) and outputs ({\em the products}).
 
@@
 unfold within cells and within tissues and organs of multicellular
-organisms. Gene product annotations identify participants in the
-processes.
+organisms. Gene product annotations can be interpreted as identifying
+participants in the processes.
 
@@
 description.
-
 \end{methods}
@@
 within a period of time (see Figure \ref{fig:patterns}).
+\begin{figure}
+  \centering
+  \includegraphics[width=.5\textwidth]{processpatterns.pdf}
+  \caption{\label{fig:patterns}Six examples of processes with
+    non-comparative and comparative process attributes.  We assume
+    that the processes labelled $a$, $b$, $c$ and $d$ are all
+    instances of the class of processes $P$.  On the left side, three
+    regulation (of $P$) processes are illustrated which exhibit
+    non-comparative attributes. The first process has an attribute of
+    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
+    temporally equidistantly distributed. The second example shows an
+    {\em arrhythmic} occurrence of $P$, and the third examples shows
+    an {\em increasing frequency} (of $P$). A regulation process with
+    an increasing frequency (of $P$) is a process in which the
+    frequency of occurrences of $P$ is lower in the first half of the
+    process than in the second half. The right side of the figure
+    illustrates comparative phenotypic descriptions of processes. On
+    the upper right, the {\em normal} reference is shown. The second
+    example illustrates a {\em late onset} of $P$, i.e., the attribute
+    that $P$ processes begin later than {\em normal}. Finally, the
+    lower right illustrates a {\em decreased frequency} (of $P$),
+    since fewer processes of the type $P$ occur within the reference
+    process than {\em normal}.}
+\end{figure}
 
@@
 single process abnormalities.
+\begin{figure*}
+  \centering
+%  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
+  \includegraphics[width=1.1\textwidth]{overview.pdf}
+  \caption{Overview over the taxonomic structure of CPO. The structure
+    is based on a cellular component class $X$ and the cellular
+    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
+\end{figure*}
 
@@
   Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
-three studies illustrated in Table \ref{tbl:studies}, live-cell
-imaging assays and RNAi knockdown were used to generate phenotypic
-profiles that quantify the cellular response to a given siRNA thus
-allowing identification of hundreds of genes involved in diverse
-biological functions including cell division, migration and survival.
-In each study, several phenotypes were detected and described by the
-authors without the use of ontologies (see Table \ref{tbl:studies}),
-making the integration between datasets extremely difficult. For
-example, it is evident that cell division phenotypes were observed in
-all three datasets (e.g. {\em Mitotic delay/arrest} and {\em Prolonged
-  mitotic exit} or {\em Methaphase delay} and {\em Methaphase cells})
-but the overlap between such phenotypes is unclear.
+three studies, live-cell imaging assays and RNAi knockdown were used
+to generate phenotypic profiles that quantify the cellular response to
+a given siRNA thus allowing identification of hundreds of genes
+involved in diverse biological functions including cell division,
+migration and survival.  In each study, several phenotypes were
+detected and described by the authors without the use of ontologies or
+controlled vocabularies, making the integration between datasets
+extremely difficult. For example, it is evident that cell division
+phenotypes were observed in all three datasets and referred to by
+terms such as {\em Mitotic delay/arrest}, {\em Prolonged mitotic
+  exit}, {\em Methaphase delay} and {\em Methaphase cells}). Without a
+controlled vocabulary of cellular phenotypes, the overlap between such
+phenotype descriptions is unclear.
 
@@
 \subsection{Future research}
+Our main contribution is an analysis of process phenotypes that are
+used across multiple domains and scales and which are crucial for
+understanding and representing physiology of living systems. The
+Ontology of Physics in Biology (OPB) \cite{opb} is an ontology that
+has recently been proposed to characterize physiological processes and
+the physical qualities of biological entities based on a theory of
+fluid dynamics. It is an important goal for future research to
+incorporate the OPB in phenotypic descriptions and make our theory of
+process phenotypes compatible with the physical descriptions of
+processes and their attributes as outlined by the OPB.
+
 We implemented the CPO using a pattern-based approach to formulating
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=1\textwidth]{onto.pdf}
-  \caption{Overview over basic top-level categories. Physical objects
-    are entities that are wholly present at time points. Qualities are
-    attributes of entities. Functions are capabilities that arise from
-    physical objects with their qualities, and processes are
-    temporally extended entities that may be realizations of a
-    function.\label{fig:onto}}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=1\textwidth]{processpatterns.pdf}
-
-  \caption{\label{fig:patterns}Six examples of processes with
-    non-comparative and comparative process attributes.  We assume
-    that the processes labelled $a$, $b$, $c$ and $d$ are all
-    instances of the class of processes $P$.  On the left side, three
-    regulation (of $P$) processes are illustrated which exhibit
-    non-comparative attributes. The first process has an attribute of
-    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
-    temporally equidistantly distributed. The second example shows an
-    {\em arrhythmic} occurrence of $P$, and the third examples shows
-    an {\em increasing frequency} (of $P$). A regulation process with
-    an increasing frequency (of $P$) is a process in which the
-    frequency of occurrences of $P$ is lower in the first half of the
-    process than in the second half. The right side of the figure
-    illustrates comparative phenotypic descriptions of processes. On
-    the upper right, the {\em normal} reference is shown. The second
-    example illustrates a {\em late onset} of $P$, i.e., the attribute
-    that $P$ processes begin later than {\em normal}. Finally, the
-    lower right illustrates a {\em decreased frequency} (of $P$),
-    since fewer processes of the type $P$ occur within the reference
-    process than {\em normal}.}
-\end{figure}
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
-  \caption{Overview over the taxonomic structure of CPO. The structure
-    is based on a cellular component class $X$ and the cellular
-    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
-\end{figure*}
-
 \section{Acknowledgement}
@@
 \bibliographystyle{plain}
-\bibliography{lc}
+\bibliography{/home/leechuck/Documents/papers/bibtex/lc}
 

### 2012-01-20 trunk/paper/drinking-sr.tex

@@
 \title{Semantic integration of physiology
-  phenotypes: an application to the cellular phenotype ontology}
+  phenotypes: application to the cellular phenotype ontology}
 
 \author{Robert Hoehndorf$^{1,*}$, Midori
-  A. Harris$^2$, Heinrich Herre$^3$, Gabriella Rustici$^4$ and\\
+  A. Harris$^2$,\\Heinrich Herre$^3$, Gabriella Rustici$^4$ and
   Georgios V. Gkoutos$^{1}$}
@@
 \bibliographystyle{unsrt}
-\bibliography{lc}
+\begin{thebibliography}{10}
+
+\bibitem{Schofield2011}
+Paul~N. Schofield, John~P. Sundberg, Robert Hoehndorf, and Georgios~V. Gkoutos.
+\newblock New approaches to the representation and analysis of phenotype
+  knowledge in human diseases and their animal models.
+\newblock {\em Briefings in Functional Genomics}, 10(5):258--265, 2011.
+
+\bibitem{Smith2007}
+Barry Smith, Michael Ashburner, Cornelius Rosse, Jonathan Bard, William Bug,
+  Werner Ceusters, Louis~J. Goldberg, Karen Eilbeck, Amelia Ireland,
+  Christopher~J. Mungall, Neocles Leontis, Philippe~R. Serra, Alan Ruttenberg,
+  Susanna~A. Sansone, Richard~H. Scheuermann, Nigam Shah, Patricia~L. Whetzel,
+  and Suzanna Lewis.
+\newblock The {OBO} {F}oundry: coordinated evolution of ontologies to support
+  biomedical data integration.
+\newblock {\em Nat Biotech}, 25(11):1251--1255, 2007.
+
+\bibitem{Gruber1995}
+Thomas~R. Gruber.
+\newblock Toward principles for the design of ontologies used for knowledge
+  sharing.
+\newblock {\em International Journal of Human-Computer Studies}, 43(5-6), 1995.
+
+\bibitem{Guarino1998}
+Nicola Guarino.
+\newblock Formal ontology and information systems.
+\newblock In Nicola Guarino, editor, {\em Proceedings of the 1st International
+  Conference on Formal Ontologies in Information Systems}, pages 3--15. IOS
+  Press, 1998.
+
+\bibitem{Bada2004}
+Michael Bada, Robert Stevens, Carole Goble, Yolanda Gil, Michael Ashburner,
+  Judith~A. Blake, Michael~J. Cherry, Midori Harris, and Suzanna Lewis.
+\newblock A short study on the success of the gene ontology.
+\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
+  Web}, 1(2):235--240, February 2004.
+
+\bibitem{go2010}
+{Gene Ontology Consortium}.
+\newblock The gene ontology in 2010: extensions and refinements.
+\newblock {\em Nucleic acids research}, 38(Database issue):D331--335, January
+  2010.
+
+\bibitem{goble}
+C.~Goble and R.~Stevens.
+\newblock State of the nation in data integration for bioinformatics.
+\newblock {\em Journal of Biomedical Informatics}, 41(5):687--693, 10 2008.
+
+\bibitem{Berners-Lee2001}
+T.~Berners-Lee, J.~Hendler, O.~Lassila, et~al.
+\newblock {The Semantic Web}.
+\newblock {\em Scientific American}, 284(5):28--37, 2001.
+
+\bibitem{Grau2008}
+B.~Grau, I.~Horrocks, B.~Motik, B.~Parsia, P.~Patelschneider, and U.~Sattler.
+\newblock {OWL} 2: The next step for {OWL}.
+\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
+  Web}, 6(4):309--322, November 2008.
+
+\bibitem{Hoehndorf2010patterns}
+Robert Hoehndorf, Anika Oellrich, Michel Dumontier, Janet Kelso, Dietrich
+  Rebholz-Schuhmann, and Heinrich Herre.
+\newblock Relations as patterns: Bridging the gap between {OBO} and {OWL}.
+\newblock {\em BMC Bioinformatics}, 11(1):441+, 2010.
+
+\bibitem{Horrocks2007}
+Ian Horrocks.
+\newblock {OBO} flat file format syntax and semantics and mapping to {OWL}
+  {W}eb {O}ntology {L}anguage.
+\newblock Technical report, University of Manchester, March 2007.
+\newblock \url{http://www.cs.man.ac.uk/~horrocks/obo/}.
+
+\bibitem{Smith2004}
+Cynthia~L. Smith, Carroll-Ann~W. Goldsmith, and Janan~T. Eppig.
+\newblock The mammalian phenotype ontology as a tool for annotating, analyzing
+  and comparing phenotypic information.
+\newblock {\em Genome Biology}, 6(1):R7, 2004.
+
+\bibitem{Robinson2008}
+P.~N. Robinson, S.~Koehler, S.~Bauer, D.~Seelow, D.~Horn, and S.~Mundlos.
+\newblock The human phenotype ontology: a tool for annotating and analyzing
+  human hereditary disease.
+\newblock {\em American journal of human genetics}, 83(5):610--615, 2008.
+
+\bibitem{ypo}
+Stacia~R. Engel, Rama Balakrishnan, Gail Binkley, Karen~R. Christie, Maria~C.
+  Costanzo, Selina~S. Dwight, Dianna~G. Fisk, Jodi~E. Hirschman, Benjamin~C.
+  Hitz, Eurie~L. Hong, Cynthia~J. Krieger, Michael~S. Livstone, Stuart~R.
+  Miyasato, Robert Nash, Rose Oughtred, Julie Park, Marek~S. Skrzypek, Shuai
+  Weng, Edith~D. Wong, Kara Dolinski, David Botstein, and J.~Michael Cherry.
+\newblock {Saccharomyces Genome Database provides mutant phenotype data}.
+\newblock {\em Nucleic Acids Research}, 38(suppl 1):D433--D436, 2010.
+
+\bibitem{wpo}
+Gary Schindelman, Jolene Fernandes, Carol Bastiani, Karen Yook, and Paul
+  Sternberg.
+\newblock Worm phenotype ontology: integrating phenotype data within and beyond
+  the c. elegans community.
+\newblock {\em BMC Bioinformatics}, 12(1):32, 2011.
+
+\bibitem{Gkoutos2005}
+Georgios~V. Gkoutos, Eain~C. Green, Ann-Marie~M. Mallon, John~M. Hancock, and
+  Duncan Davidson.
+\newblock {Using ontologies to describe mouse phenotypes.}
+\newblock {\em Genome biology}, 6(1), 2005.
+
+\bibitem{obml2011h1}
+Georgios~V. Gkoutos and Robert Hoehndorf.
+\newblock Ontology-based cross-species integration and analysis of
+  saccharomyces cerevisiae phenotypes.
+\newblock In {\em Proceedings of the 3rd Workshop for Ontologies in Biomedicine
+  and Life sciences (OBML)}, October 2011.
+
+\bibitem{Gkoutos2009b}
+Georgios~V. Gkoutos, Chris Mungall, Sandra Dolken, Michael Ashburner, Suzanna
+  Lewis, John Hancock, Paul Schofield, Sebastian Kohler, and Peter~N. Robinson.
+\newblock Entity/quality-based logical definitions for the human skeletal
+  phenome using {PATO}.
+\newblock {\em Annual International Conference of the IEEE Engineering in
+  Medicine and Biology Society.}, 1:7069--7072, 2009.
+
+\bibitem{Mungall2010}
+Christopher Mungall, Georgios Gkoutos, Cynthia Smith, Melissa Haendel, Suzanna
+  Lewis, and Michael Ashburner.
+\newblock Integrating phenotype ontologies across multiple species.
+\newblock {\em Genome Biology}, 11(1):R2+, 2010.
+
+\bibitem{Hoehndorf2010phene}
+Robert Hoehndorf, Anika Oellrich, and Dietrich Rebholz-Schuhmann.
+\newblock Interoperability between phenotype and anatomy ontologies.
+\newblock {\em Bioinformatics}, 26(24):3112 -- 3118, 10 2010.
+
+\bibitem{Mungall2010go}
+Christopher~J. Mungall, Michael Bada, Tanya~Z. Berardini, Jennifer Deegan,
+  Amelia Ireland, Midori~A. Harris, David~P. Hill, and Jane Lomax.
+\newblock Cross-product extensions of the gene ontology.
+\newblock {\em Journal of biomedical informatics}, February 2010.
+\newblock in press.
+
+\bibitem{Bada2007a}
+M.~Bada and L.~Hunter.
+\newblock Enrichment of obo ontologies.
+\newblock {\em Journal of Biomedical Informatics}, 40(3):300--315, June 2007.
+
+\bibitem{Ogren2004}
+P.~V. Ogren, K.~B. Cohen, G.~K. Acquaah-Mensah, J.~Eberlein, and L.~Hunter.
+\newblock The compositional structure of gene ontology terms.
+\newblock {\em Pac Symp Biocomput}, pages 214--225, 2004.
+
+\bibitem{Horridge2007}
+Matthew Horridge, Sean Bechhofer, and Olaf Noppens.
+\newblock Igniting the {OWL} 1.1 touch paper: The {OWL} {API}.
+\newblock In {\em Proceedings of OWLED 2007: Third International Workshop on
+  OWL Experiences and Directions}, 2007.
+
+\bibitem{Kazakov2011}
+Yevgeny Kazakov, Markus Kr{\"o}tzsch, and Franti\v{s}ek Siman\v{c}\'{i}k.
+\newblock Unchain my $\mathcal{EL}$ reasoner.
+\newblock In {\em Proceedings of the 23rd International Workshop on Description
+  Logics (DL'10)}, CEUR Workshop Proceedings. CEUR-WS.org, 2011.
+
+\bibitem{Lock2010}
+John~G Lock and Staffan Strömblad.
+\newblock Systems microscopy: an emerging strategy for the life sciences.
+\newblock {\em Experimental Cell Research}, 316(8):1438--1444, 2010.
+
+\bibitem{Fuchs2010}
+Florian Fuchs, Gregoire Pau, Dominique Kranz, Oleg Sklyar, Christoph Budjan,
+  Sandra Steinbrink, Thomas Horn, Angelika Pedal, Wolfgang Huber, and Michael
+  Boutros.
+\newblock Clustering phenotype populations by genome-wide {RNAi} and
+  multiparametric imaging.
+\newblock {\em Molecular Systems Biology}, 6, June 2010.
+
+\bibitem{Neumann2010}
+Beate Neumann, Thomas Walter, Jean-Karim Hériché, Jutta Bulkescher, Holger
+  Erfle, Christian Conrad, Phill Rogers, Ina Poser, Michael Held, Urban Liebel,
+  and et~al.
+\newblock Phenotypic profiling of the human genome by time-lapse microscopy
+  reveals cell division genes.
+\newblock {\em Nature}, 464(7289):721--727, 2010.
+
+\bibitem{Schmitz2010}
+Michael H.~A. Schmitz, Michael Held, Veerle Janssens, James R.~A. Hutchins,
+  Otto Hudecz, Elitsa Ivanova, Jozef Goris, Laura Trinkle-Mulcahy, Angus~I.
+  Lamond, Ina Poser, Anthony~A. Hyman, Karl Mechtler, Jan-Michael Peters, and
+  Daniel~W. Gerlich.
+\newblock Live-cell imaging rnai screen identifies pp2a–b55$\alpha$ and
+  importin-$\beta$1 as key mitotic exit regulators in human cells.
+\newblock {\em Nature Cell Biology}, 12:886--893, 2010.
+
+\bibitem{Conrad2011}
+Christian Conrad, Annelie W\"{u}nsche, Tze Heng~H. Tan, Jutta Bulkescher, Frank
+  Sieckmann, Fatima Verissimo, Arthur Edelstein, Thomas Walter, Urban Liebel,
+  Rainer Pepperkok, and Jan Ellenberg.
+\newblock Micropilot: automation of fluorescence microscopy-based imaging for
+  systems biology.
+\newblock {\em Nature methods}, 8(3):246--249, March 2011.
+
+\bibitem{Herre2006}
+Heinrich Herre, Barbara Heller, Patryk Burek, Robert Hoehndorf, Frank Loebe,
+  and Hannes Michalek.
+\newblock {G}eneral {F}ormal {O}ntology ({GFO}) -- {A} foundational ontology
+  integrating objects and processes [{V}ersion 1.0].
+\newblock Onto-Med Report~8, IMISE, University of Leipzig, Leipzig, Germany,
+  2006.
+
+\bibitem{Gruninger2010}
+Atalay \"{O}zg\"{o}vde and Michael Gr\"{u}ninger.
+\newblock Foundational process relations in bio-ontologies.
+\newblock In {\em Proceeding of the 2010 conference on Formal Ontology in
+  Information Systems}, pages 243--256, Amsterdam, The Netherlands, The
+  Netherlands, 2010. IOS Press.
+
+\bibitem{Barwise1989}
+J.~Barwise.
+\newblock {\em The Situation in Logic}.
+\newblock CSLI, Stanford, CA, 1989.
+
+\bibitem{Hilbert1918}
+David Hilbert.
+\newblock {A}xiomatisches {D}enken.
+\newblock {\em {M}athematische {A}nnalen}, 78:405--415, 1918.
+
+\bibitem{Hoehndorf2011models}
+Robert Hoehndorf, Michel Dumontier, John~H. Gennari, Sarala Wimalaratne,
+  Bernard de~Bono, Daniel~L. Cook, and Georgios~V. Gkoutos.
+\newblock Integrating systems biology models and biomedical ontologies.
+\newblock {\em BMC Systems Biology}, 5(1):124+, August 2011.
+
+\bibitem{Hoehndorf2011incon}
+Robert Hoehndorf, Michel Dumontier, Anika Oellrich, Dietrich Rebholz-Schuhmann,
+  Paul~N. Schofield, and Georgios~V. Gkoutos.
+\newblock Interoperability between biomedical ontologies through relation
+  expansion, upper-level ontologies and automatic reasoning.
+\newblock {\em PLOS ONE}, 6(7):e22006, July 2011.
+
+\bibitem{Motik2009a}
+Boris Motik, Rob Shearer, and Ian Horrocks.
+\newblock {Hypertableau Reasoning for Description Logics}.
+\newblock {\em Journal of Artificial Intelligence Research}, 36:165--228, 2009.
+
+\bibitem{Sirin2004}
+Evren Sirin and Bijan Parsia.
+\newblock Pellet: An {OWL} {DL} reasoner.
+\newblock In Volker Haarslev and Ralf M{\"{o}}ller, editors, {\em Proceedings
+  of the 2004 International Workshop on Description Logics, DL2004, Whistler,
+  British Columbia, Canada, Jun 6-8}, volume 104 of {\em CEUR Workshop
+  Proceedings}, Aachen, Germany, 2004. CEUR-WS.org.
+
+\bibitem{Tsarkov2006}
+D.~Tsarkov and I.~Horrocks.
+\newblock {FaCT++} description logic reasoner: System description.
+\newblock {\em Lecture Notes in Computer Science (including subseries Lecture
+  Notes in Artificial Intelligence and Lecture Notes in Bioinformatics)}, 4130
+  LNAI:292--297, 2006.
+
+\bibitem{el4}
+F.~Baader, C.~Lutz, and B.~Suntisrivaraporn.
+\newblock {CEL} -- a polynomial-time reasoner for life science ontologies.
+\newblock In U.~Furbach and N.~Shankar, editors, {\em Proceedings of the 3rd
+  International Joint Conference on Automated Reasoning ({IJCAR'06})}, volume
+  4130 of {\em Lecture Notes in Artificial Intelligence}, pages 287--291.
+  Springer-Verlag, 2006.
+
+\bibitem{elvira}
+Robert Hoehndorf, Michel Dumontier, Anika Oellrich, Sarala Wimalaratne,
+  Dietrich Rebholz-Schuhmann, Paul Schofield, and Georgios~V. Gkoutos.
+\newblock A common layer of interoperability for biomedical ontologies based on
+  {OWL} {EL}.
+\newblock {\em Bioinformatics}, 27(7):1001--1008, April 2011.
+
+\bibitem{cbreasoner}
+Yevgeny Kazakov.
+\newblock Consequence-driven reasoning for {H}orn {SHIQ} ontologies.
+\newblock In {\em Proceedings of the 21st International Conference on
+  Artificial Intelligence (IJCAI 2009)}, pages 2040--2045, July 11-17 2009.
+
+\bibitem{rnao}
+Robert Hoehndorf, Colin Batchelor, Thomas Bittner, Michel Dumontier, Karen
+  Eilbeck, Rob Knight, Chris~J. Mungall, Jane~S. Richardson, Jesse Stombaugh,
+  Eric Westhof, Craig~L. Zirbel, and Neocles~B. Leontis.
+\newblock The {RNA} ontology ({RNAO}): An ontology for integrating {RNA}
+  sequence and structure data.
+\newblock {\em Applied Ontology}, 6(1):53--89, April 2011.
+
+\bibitem{Hoehndorf2009sequences}
+Robert Hoehndorf, Janet Kelso, and Heinrich Herre.
+\newblock The ontology of biological sequences.
+\newblock {\em BMC bioinformatics}, 10(1):377+, November 2009.
+
+\bibitem{Burek2006}
+Patryk Burek.
+\newblock {\em Ontology of Functions}.
+\newblock PhD thesis, University of Leipzig, Institute of Informatics (IfI),
+  2006.
+
+\bibitem{Herre2010}
+Heinrich Herre.
+\newblock {G}eneral {F}ormal {O}ntology {(GFO)}: A foundational ontology for
+  conceptual modelling.
+\newblock In Roberto Poli, Michael Healy, and Achilles Kameas, editors, {\em
+  Theory and Applications of Ontology: Computer Applications}, chapter~14,
+  pages 297{--}345. Springer, Heidelberg, 2010.
+
+\bibitem{Loebe2007}
+Frank Loebe.
+\newblock Abstract vs. social roles -- towards a general theoretical account of
+  roles.
+\newblock {\em Applied Ontology}, 2(2):127--158, 2007.
+
+\bibitem{Ashburner2000b}
+M.~Ashburner, C.~A. Ball, J.~A. Blake, D.~Botstein, H.~Butler, J.~M. Cherry,
+  A.~P. Davis, K.~Dolinski, S.~S. Dwight, J.~T. Eppig, M.~A. Harris, D.~P.
+  Hill, L.~Issel-Tarver, A.~Kasarskis, S.~Lewis, J.~C. Matese, J.~E.
+  Richardson, M.~Ringwald, G.~M. Rubin, and G.~Sherlock.
+\newblock Gene ontology: tool for the unification of biology. the gene ontology
+  consortium.
+\newblock {\em Nat Genet}, 25(1):25--29, May 2000.
+
+\bibitem{Bard2005}
+Jonathan Bard, Seung~Y. Rhee, and Michael Ashburner.
+\newblock An ontology for cell types.
+\newblock {\em Genome Biology}, 6(2), 2005.
+
+\bibitem{Eilbeck2005}
+K.~Eilbeck, S.~E. Lewis, C.~J. Mungall, M.~Yandell, L.~Stein, R.~Durbin, and
+  M.~Ashburner.
+\newblock The sequence ontology: A tool for the unification of genome
+  annotations.
+\newblock {\em Genome Biology}, 6(R55), 2005.
+
+\end{thebibliography}
+
+%\bibliography{lc}
 

### 2012-01-20 paper/drinking-bioinformatics.tex

--- 2012-01-20 trunk/paper/drinking-bioinformatics.tex
@@
 \usepackage{multirow}
-\usepackage{pifont}
-%\newcommand{\tick}{\ding{52}}
-\newcommand{\tick}{\checkmark}
-\newcommand{\tickNo}{\hspace{1pt}\ding{55}}
+\renewcommand{\cite}{\citep}
 
@@
   \centering
-  \includegraphics[width=.5\textwidth]{processpatterns.pdf}
+  \includegraphics[width=.5\textwidth]{processpatterns.eps}
   \caption{\label{fig:patterns}Six examples of processes with
@@
   \centering
-%  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
-  \includegraphics[width=1.1\textwidth]{overview.pdf}
+  \includegraphics[width=1\textwidth]{overview.eps}
+%  \includegraphics[width=1\textwidth]{overview.pdf}
   \caption{Overview over the taxonomic structure of CPO. The structure
@@
 
-\bibliographystyle{plain}
-\bibliography{/home/leechuck/Documents/papers/bibtex/lc}
+\bibliographystyle{natbib}
+\begin{thebibliography}{}
+
+\bibitem[Ashburner {\em et~al.}(2000)Ashburner, Ball, Blake, Botstein, Butler,
+  Cherry, Davis, Dolinski, Dwight, Eppig, Harris, Hill, Issel-Tarver,
+  Kasarskis, Lewis, Matese, Richardson, Ringwald, Rubin, and
+  Sherlock]{Ashburner2000b}
+Ashburner, M., Ball, C.~A., Blake, J.~A., Botstein, D., Butler, H., Cherry,
+  J.~M., Davis, A.~P., Dolinski, K., Dwight, S.~S., Eppig, J.~T., Harris,
+  M.~A., Hill, D.~P., Issel-Tarver, L., Kasarskis, A., Lewis, S., Matese,
+  J.~C., Richardson, J.~E., Ringwald, M., Rubin, G.~M., and Sherlock, G.
+  (2000).
+\newblock Gene ontology: tool for the unification of biology. the gene ontology
+  consortium.
+\newblock {\em Nat Genet\/}, {\bf 25}(1), 25--29.
+
+\bibitem[Bada and Hunter(2007)Bada and Hunter]{Bada2007a}
+Bada, M. and Hunter, L. (2007).
+\newblock Enrichment of obo ontologies.
+\newblock {\em Journal of Biomedical Informatics\/}, {\bf 40}(3), 300--315.
+
+\bibitem[Bada {\em et~al.}(2004)Bada, Stevens, Goble, Gil, Ashburner, Blake,
+  Cherry, Harris, and Lewis]{Bada2004}
+Bada, M., Stevens, R., Goble, C., Gil, Y., Ashburner, M., Blake, J.~A., Cherry,
+  M.~J., Harris, M., and Lewis, S. (2004).
+\newblock A short study on the success of the gene ontology.
+\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
+  Web\/}, {\bf 1}(2), 235--240.
+
+\bibitem[Bard {\em et~al.}(2005)Bard, Rhee, and Ashburner]{Bard2005}
+Bard, J., Rhee, S.~Y., and Ashburner, M. (2005).
+\newblock An ontology for cell types.
+\newblock {\em Genome Biology\/}, {\bf 6}(2).
+
+\bibitem[Berners-Lee {\em et~al.}(2001)Berners-Lee, Hendler, Lassila, {\em
+  et~al.}]{Berners-Lee2001}
+Berners-Lee, T., Hendler, J., Lassila, O., {\em et~al.} (2001).
+\newblock {The Semantic Web}.
+\newblock {\em Scientific American\/}, {\bf 284}(5), 28--37.
+
+\bibitem[Conrad {\em et~al.}(2011)Conrad, W\"{u}nsche, Tan, Bulkescher,
+  Sieckmann, Verissimo, Edelstein, Walter, Liebel, Pepperkok, and
+  Ellenberg]{Conrad2011}
+Conrad, C., W\"{u}nsche, A., Tan, T. H.~H., Bulkescher, J., Sieckmann, F.,
+  Verissimo, F., Edelstein, A., Walter, T., Liebel, U., Pepperkok, R., and
+  Ellenberg, J. (2011).
+\newblock Micropilot: automation of fluorescence microscopy-based imaging for
+  systems biology.
+\newblock {\em Nature methods\/}, {\bf 8}(3), 246--249.
+
+\bibitem[Cook {\em et~al.}(2011)Cook, Bookstein, and Gennari]{opb}
+Cook, D.~L., Bookstein, F.~L., and Gennari, J.~H. (2011).
+\newblock Physical properties of biological entities: An introduction to the
+  ontology of physics for biology.
+\newblock {\em PLoS ONE\/}, {\bf 6}(12), e28708.
+
+\bibitem[Eilbeck {\em et~al.}(2005)Eilbeck, Lewis, Mungall, Yandell, Stein,
+  Durbin, and Ashburner]{Eilbeck2005}
+Eilbeck, K., Lewis, S.~E., Mungall, C.~J., Yandell, M., Stein, L., Durbin, R.,
+  and Ashburner, M. (2005).
+\newblock The sequence ontology: A tool for the unification of genome
+  annotations.
+\newblock {\em Genome Biology\/}, {\bf 6}(R55).
+
+\bibitem[Engel {\em et~al.}(2010)Engel, Balakrishnan, Binkley, Christie,
+  Costanzo, Dwight, Fisk, Hirschman, Hitz, Hong, Krieger, Livstone, Miyasato,
+  Nash, Oughtred, Park, Skrzypek, Weng, Wong, Dolinski, Botstein, and
+  Cherry]{ypo}
+Engel, S.~R., Balakrishnan, R., Binkley, G., Christie, K.~R., Costanzo, M.~C.,
+  Dwight, S.~S., Fisk, D.~G., Hirschman, J.~E., Hitz, B.~C., Hong, E.~L.,
+  Krieger, C.~J., Livstone, M.~S., Miyasato, S.~R., Nash, R., Oughtred, R.,
+  Park, J., Skrzypek, M.~S., Weng, S., Wong, E.~D., Dolinski, K., Botstein, D.,
+  and Cherry, J.~M. (2010).
+\newblock {Saccharomyces Genome Database provides mutant phenotype data}.
+\newblock {\em Nucleic Acids Research\/}, {\bf 38}(suppl 1), D433--D436.
+
+\bibitem[Fuchs {\em et~al.}(2010)Fuchs, Pau, Kranz, Sklyar, Budjan, Steinbrink,
+  Horn, Pedal, Huber, and Boutros]{Fuchs2010}
+Fuchs, F., Pau, G., Kranz, D., Sklyar, O., Budjan, C., Steinbrink, S., Horn,
+  T., Pedal, A., Huber, W., and Boutros, M. (2010).
+\newblock Clustering phenotype populations by genome-wide {RNAi} and
+  multiparametric imaging.
+\newblock {\em Molecular Systems Biology\/}, {\bf 6}.
+
+\bibitem[{Gene Ontology Consortium}(2010){Gene Ontology Consortium}]{go2010}
+{Gene Ontology Consortium} (2010).
+\newblock The gene ontology in 2010: extensions and refinements.
+\newblock {\em Nucleic acids research\/}, {\bf 38}(Database issue), D331--335.
+
+\bibitem[Gkoutos and Hoehndorf(2011)Gkoutos and Hoehndorf]{obml2011h1}
+Gkoutos, G.~V. and Hoehndorf, R. (2011).
+\newblock Ontology-based cross-species integration and analysis of
+  saccharomyces cerevisiae phenotypes.
+\newblock In {\em Proceedings of the 3rd Workshop for Ontologies in Biomedicine
+  and Life sciences (OBML)\/}.
+
+\bibitem[Gkoutos {\em et~al.}(2005)Gkoutos, Green, Mallon, Hancock, and
+  Davidson]{Gkoutos2005}
+Gkoutos, G.~V., Green, E.~C., Mallon, A.-M.~M., Hancock, J.~M., and Davidson,
+  D. (2005).
+\newblock {Using ontologies to describe mouse phenotypes.}
+\newblock {\em Genome biology\/}, {\bf 6}(1).
+
+\bibitem[Gkoutos {\em et~al.}(2009)Gkoutos, Mungall, Dolken, Ashburner, Lewis,
+  Hancock, Schofield, Kohler, and Robinson]{Gkoutos2009b}
+Gkoutos, G.~V., Mungall, C., Dolken, S., Ashburner, M., Lewis, S., Hancock, J.,
+  Schofield, P., Kohler, S., and Robinson, P.~N. (2009).
+\newblock Entity/quality-based logical definitions for the human skeletal
+  phenome using {PATO}.
+\newblock {\em Annual International Conference of the IEEE Engineering in
+  Medicine and Biology Society.}, {\bf 1}, 7069--7072.
+
+\bibitem[Goble and Stevens(2008)Goble and Stevens]{goble}
+Goble, C. and Stevens, R. (2008).
+\newblock State of the nation in data integration for bioinformatics.
+\newblock {\em Journal of Biomedical Informatics\/}, {\bf 41}(5), 687--693.
+
+\bibitem[Grau {\em et~al.}(2008)Grau, Horrocks, Motik, Parsia, Patelschneider,
+  and Sattler]{Grau2008}
+Grau, B., Horrocks, I., Motik, B., Parsia, B., Patelschneider, P., and Sattler,
+  U. (2008).
+\newblock {OWL} 2: The next step for {OWL}.
+\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
+  Web\/}, {\bf 6}(4), 309--322.
+
+\bibitem[Gruber(1995)Gruber]{Gruber1995}
+Gruber, T.~R. (1995).
+\newblock Toward principles for the design of ontologies used for knowledge
+  sharing.
+\newblock {\em International Journal of Human-Computer Studies\/}, {\bf
+  43}(5-6).
+
+\bibitem[Guarino(1998)Guarino]{Guarino1998}
+Guarino, N. (1998).
+\newblock Formal ontology and information systems.
+\newblock In N.~Guarino, editor, {\em Proceedings of the 1st International
+  Conference on Formal Ontologies in Information Systems\/}, pages 3--15. IOS
+  Press.
+
+\bibitem[Herre {\em et~al.}(2006)Herre, Heller, Burek, Hoehndorf, Loebe, and
+  Michalek]{Herre2006}
+Herre, H., Heller, B., Burek, P., Hoehndorf, R., Loebe, F., and Michalek, H.
+  (2006).
+\newblock {G}eneral {F}ormal {O}ntology ({GFO}) -- {A} foundational ontology
+  integrating objects and processes [{V}ersion 1.0].
+\newblock Onto-Med Report~8, IMISE, University of Leipzig, Leipzig, Germany.
+
+\bibitem[Hoehndorf {\em et~al.}(2010a)Hoehndorf, Oellrich, and
+  Rebholz-Schuhmann]{Hoehndorf2010phene}
+Hoehndorf, R., Oellrich, A., and Rebholz-Schuhmann, D. (2010a).
+\newblock Interoperability between phenotype and anatomy ontologies.
+\newblock {\em Bioinformatics\/}, {\bf 26}(24), 3112 -- 3118.
+
+\bibitem[Hoehndorf {\em et~al.}(2010b)Hoehndorf, Oellrich, Dumontier, Kelso,
+  Rebholz-Schuhmann, and Herre]{Hoehndorf2010patterns}
+Hoehndorf, R., Oellrich, A., Dumontier, M., Kelso, J., Rebholz-Schuhmann, D.,
+  and Herre, H. (2010b).
+\newblock Relations as patterns: Bridging the gap between {OBO} and {OWL}.
+\newblock {\em BMC Bioinformatics\/}, {\bf 11}(1), 441+.
+
+\bibitem[Horridge {\em et~al.}(2007)Horridge, Bechhofer, and
+  Noppens]{Horridge2007}
+Horridge, M., Bechhofer, S., and Noppens, O. (2007).
+\newblock Igniting the {OWL} 1.1 touch paper: The {OWL} {API}.
+\newblock In {\em Proceedings of OWLED 2007: Third International Workshop on
+  OWL Experiences and Directions\/}.
+
+\bibitem[Horrocks(2007)Horrocks]{Horrocks2007}
+Horrocks, I. (2007).
+\newblock {OBO} flat file format syntax and semantics and mapping to {OWL}
+  {W}eb {O}ntology {L}anguage.
+\newblock Technical report, University of Manchester.
+\newblock \url{http://www.cs.man.ac.uk/~horrocks/obo/}.
+
+\bibitem[Kazakov {\em et~al.}(2011)Kazakov, Kr{\"o}tzsch, and
+  Siman\v{c}\'{i}k]{Kazakov2011}
+Kazakov, Y., Kr{\"o}tzsch, M., and Siman\v{c}\'{i}k, F. (2011).
+\newblock Unchain my $\mathcal{EL}$ reasoner.
+\newblock In {\em Proceedings of the 23rd International Workshop on Description
+  Logics (DL'10)\/}, CEUR Workshop Proceedings. CEUR-WS.org.
+
+\bibitem[Lock and Strömblad(2010)Lock and Strömblad]{Lock2010}
+Lock, J.~G. and Strömblad, S. (2010).
+\newblock Systems microscopy: an emerging strategy for the life sciences.
+\newblock {\em Experimental Cell Research\/}, {\bf 316}(8), 1438--1444.
+
+\bibitem[Mungall {\em et~al.}(2010a)Mungall, Gkoutos, Smith, Haendel, Lewis,
+  and Ashburner]{Mungall2010}
+Mungall, C., Gkoutos, G., Smith, C., Haendel, M., Lewis, S., and Ashburner, M.
+  (2010a).
+\newblock Integrating phenotype ontologies across multiple species.
+\newblock {\em Genome Biology\/}, {\bf 11}(1), R2+.
+
+\bibitem[Mungall {\em et~al.}(2010b)Mungall, Bada, Berardini, Deegan, Ireland,
+  Harris, Hill, and Lomax]{Mungall2010go}
+Mungall, C.~J., Bada, M., Berardini, T.~Z., Deegan, J., Ireland, A., Harris,
+  M.~A., Hill, D.~P., and Lomax, J. (2010b).
+\newblock Cross-product extensions of the gene ontology.
+\newblock {\em Journal of biomedical informatics\/}.
+\newblock in press.
+
+\bibitem[Neumann {\em et~al.}(2010)Neumann, Walter, Hériché, Bulkescher,
+  Erfle, Conrad, Rogers, Poser, Held, Liebel, and et~al.]{Neumann2010}
+Neumann, B., Walter, T., Hériché, J.-K., Bulkescher, J., Erfle, H., Conrad,
+  C., Rogers, P., Poser, I., Held, M., Liebel, U., and et~al. (2010).
+\newblock Phenotypic profiling of the human genome by time-lapse microscopy
+  reveals cell division genes.
+\newblock {\em Nature\/}, {\bf 464}(7289), 721--727.
+
+\bibitem[Ogren {\em et~al.}(2004)Ogren, Cohen, Acquaah-Mensah, Eberlein, and
+  Hunter]{Ogren2004}
+Ogren, P.~V., Cohen, K.~B., Acquaah-Mensah, G.~K., Eberlein, J., and Hunter, L.
+  (2004).
+\newblock The compositional structure of gene ontology terms.
+\newblock {\em Pac Symp Biocomput\/}, pages 214--225.
+
+\bibitem[\"{O}zg\"{o}vde and Gr\"{u}ninger(2010)\"{O}zg\"{o}vde and
+  Gr\"{u}ninger]{Gruninger2010}
+\"{O}zg\"{o}vde, A. and Gr\"{u}ninger, M. (2010).
+\newblock Foundational process relations in bio-ontologies.
+\newblock In {\em Proceeding of the 2010 conference on Formal Ontology in
+  Information Systems\/}, pages 243--256, Amsterdam, The Netherlands, The
+  Netherlands. IOS Press.
+
+\bibitem[Robinson {\em et~al.}(2008)Robinson, Koehler, Bauer, Seelow, Horn, and
+  Mundlos]{Robinson2008}
+Robinson, P.~N., Koehler, S., Bauer, S., Seelow, D., Horn, D., and Mundlos, S.
+  (2008).
+\newblock The human phenotype ontology: a tool for annotating and analyzing
+  human hereditary disease.
+\newblock {\em American journal of human genetics\/}, {\bf 83}(5), 610--615.
+
+\bibitem[Schindelman {\em et~al.}(2011)Schindelman, Fernandes, Bastiani, Yook,
+  and Sternberg]{wpo}
+Schindelman, G., Fernandes, J., Bastiani, C., Yook, K., and Sternberg, P.
+  (2011).
+\newblock Worm phenotype ontology: integrating phenotype data within and beyond
+  the c. elegans community.
+\newblock {\em BMC Bioinformatics\/}, {\bf 12}(1), 32.
+
+\bibitem[Schmitz {\em et~al.}(2010)Schmitz, Held, Janssens, Hutchins, Hudecz,
+  Ivanova, Goris, Trinkle-Mulcahy, Lamond, Poser, Hyman, Mechtler, Peters, and
+  Gerlich]{Schmitz2010}
+Schmitz, M. H.~A., Held, M., Janssens, V., Hutchins, J. R.~A., Hudecz, O.,
+  Ivanova, E., Goris, J., Trinkle-Mulcahy, L., Lamond, A.~I., Poser, I., Hyman,
+  A.~A., Mechtler, K., Peters, J.-M., and Gerlich, D.~W. (2010).
+\newblock Live-cell imaging rnai screen identifies pp2a–b55$\alpha$ and
+  importin-$\beta$1 as key mitotic exit regulators in human cells.
+\newblock {\em Nature Cell Biology\/}, {\bf 12}, 886--893.
+
+\bibitem[Schofield {\em et~al.}(2011)Schofield, Sundberg, Hoehndorf, and
+  Gkoutos]{Schofield2011}
+Schofield, P.~N., Sundberg, J.~P., Hoehndorf, R., and Gkoutos, G.~V. (2011).
+\newblock New approaches to the representation and analysis of phenotype
+  knowledge in human diseases and their animal models.
+\newblock {\em Briefings in Functional Genomics\/}, {\bf 10}(5), 258--265.
+
+\bibitem[Smith {\em et~al.}(2007)Smith, Ashburner, Rosse, Bard, Bug, Ceusters,
+  Goldberg, Eilbeck, Ireland, Mungall, Leontis, Serra, Ruttenberg, Sansone,
+  Scheuermann, Shah, Whetzel, and Lewis]{Smith2007}
+Smith, B., Ashburner, M., Rosse, C., Bard, J., Bug, W., Ceusters, W., Goldberg,
+  L.~J., Eilbeck, K., Ireland, A., Mungall, C.~J., Leontis, N., Serra, P.~R.,
+  Ruttenberg, A., Sansone, S.~A., Scheuermann, R.~H., Shah, N., Whetzel, P.~L.,
+  and Lewis, S. (2007).
+\newblock The {OBO} {F}oundry: coordinated evolution of ontologies to support
+  biomedical data integration.
+\newblock {\em Nat Biotech\/}, {\bf 25}(11), 1251--1255.
+
+\bibitem[Smith {\em et~al.}(2004)Smith, Goldsmith, and Eppig]{Smith2004}
+Smith, C.~L., Goldsmith, C.-A.~W., and Eppig, J.~T. (2004).
+\newblock The mammalian phenotype ontology as a tool for annotating, analyzing
+  and comparing phenotypic information.
+\newblock {\em Genome Biology\/}, {\bf 6}(1), R7.
+
+\end{thebibliography}
+
+%\bibliography{/home/leechuck/Documents/papers/bibtex/lc}
 

### 2012-03-18 paper/drinking-bioinformatics.tex

--- 2012-01-20 trunk/paper/drinking-bioinformatics.tex
@@
 integrate biomedical data across heterogeneous information systems,
-biomedical ontologies are being developed \cite{Smith2007}. An
+biomedical ontologies are being developed \cite{Smith2007short}. An
 ontology is an explicit specification of a conceptualization of a
@@
 developed. In particular, ontologies to characterize mammalian
-\cite{Smith2004}, human \cite{Robinson2008}, yeast \cite{ypo} and worm
-\cite{wpo} phenotypes are now available, while several more phenotype
-ontologies are under development. To benefit from automated reasoning,
-integrate phenotypes across species and reuse the content of anatomy
-and process ontologies, we have defined process classes using the
-framework of the Phenotypic Attribute and Trait (PATO) ontology
-\cite{Gkoutos2005}. According to the PATO framework, a phenotype can
-be decomposed, using an Entity-Quality model, into an affected entity
-and a quality that characterizes {\em how} the entity is affected
-\cite{Gkoutos2005}. Such decompositions have been created for several
-widely used phenotype ontologies \cite{Mungall2010, Gkoutos2009b,
-  obml2011h1}, and are being applied together with methods for reusing
-knowledge contained in anatomy ontologies \cite{Mungall2010,
-  Hoehndorf2010phene}.
+\cite{Smith2004}, human \cite{Robinson2008}, yeast \cite{aposhort} and
+worm \cite{wpo} phenotypes are now available, while several more
+phenotype ontologies are under development. To benefit from automated
+reasoning, integrate phenotypes across species and reuse the content
+of anatomy and process ontologies, we have defined process classes
+using the framework of the Phenotypic Attribute and Trait (PATO)
+ontology \cite{Gkoutos2005}. According to the PATO framework, a
+phenotype can be decomposed, using an Entity-Quality model, into an
+affected entity and a quality that characterizes {\em how} the entity
+is affected \cite{Gkoutos2005}. Such decompositions have been created
+for several widely used phenotype ontologies \cite{Mungall2010,
+  Gkoutos2009b, obml2011h1}, and are being applied together with
+methods for reusing knowledge contained in anatomy ontologies
+\cite{Mungall2010, Hoehndorf2010phene}.
 
@@
 
-To illustrate these challenges, consider physiological processes of
-the heart. One of the heart's functions is {\em Heart beating}, i.e.,
-a capability that is realized through processes of the type {\em Heart
-  beating}. {\em Blood} is a participant of {\em Heart beating}
-processes.  An abnormal phenotype of an organism could be that the
-{\em rate of heart beating} is increased. The intended meaning of such
-a description is that the number of {\em Heart beating} processes in a
-given time interval is higher than normal. Another important attribute
-of heart physiology is the fluid flow rate through the heart. For an
-abnormal phenotype such as {\em increased rate of fluid flow through
-  the heart}, the intended meaning could either be that the amount of
-fluid that is moved through the heart within a single heart beating
-process is increased or that the amount of fluid that is moved through
-the heart within a period of time interval is increased.
+% To illustrate these challenges, consider physiological processes of
+% the heart. One of the heart's functions is {\em Heart beating}, i.e.,
+% a capability that is realized through processes of the type {\em Heart
+%   beating}. {\em Blood} is a participant of {\em Heart beating}
+% processes.  An abnormal phenotype of an organism could be that the
+% {\em rate of heart beating} is increased. The intended meaning of such
+% a description is that the number of {\em Heart beating} processes in a
+% given time interval is higher than normal. Another important attribute
+% of heart physiology is the fluid flow rate through the heart. For an
+% abnormal phenotype such as {\em increased rate of fluid flow through
+%   the heart}, the intended meaning could either be that the amount of
+% fluid that is moved through the heart within a single heart beating
+% process is increased or that the amount of fluid that is moved through
+% the heart within a period of time is increased.
 
-Based on these examples, we can make several observations about
-process attributes. First, for a process like {\em Heart beating}, we
-can distinguish between single occurrences and processes in which {\em
-  Heart beating} occurs multiple times. Only the latter kind of
-process may have a {\em rate of heart beating}, while {\em Heart
-  beating} processes do not have such an attribute. Second, we can
-distinguish between abnormal fluid flow rates in {\em Heart beating}
-processes and rate of fluid flow through the heart within a given
-duration. Both may have entirely different underlying causes and it is
-therefore important to distinguish between them. Finally, we may be
-able to infer some phenotypes from others, thereby limiting the number
-of phenotypes that must be experimentally observed. For example, when
-the fluid flow rate in single heart beating processes is increased and
-the {\em rate of heart beating} is increased within a process $P$,
-then the rate of fluid flow through the heart will be increased for
-$P$.
+% Based on these examples, we can make several observations about
+% process attributes. First, for a process like {\em Heart beating}, we
+% can distinguish between single occurrences and processes in which {\em
+%   Heart beating} occurs multiple times. Only the latter kind of
+% process may have a {\em rate of heart beating}, while {\em Heart
+%   beating} processes do not have such an attribute. Second, we can
+% distinguish between abnormal fluid flow rates in {\em Heart beating}
+% processes and rate of fluid flow through the heart within a given
+% duration. Both may have entirely different underlying causes and it is
+% therefore important to distinguish between them. Finally, we may be
+% able to infer some phenotypes from others, thereby limiting the number
+% of phenotypes that must be experimentally observed. For example, when
+% the fluid flow rate in single heart beating processes is increased and
+% the {\em rate of heart beating} is increased within a process $P$,
+% then the rate of fluid flow through the heart will be increased for
+% $P$.
 
 Here, we present the foundations for an ontology of process
-phenotypes. We present the outline of a theory in which several kinds
-of process attributes can be distinguished so that normal and abnormal
-physiology of biological systems can be formally characterized. We
-apply this theory to cellular processes and create the Cell Phenotype
-Ontology (CPO). The CPO is linked to reference ontologies for
-qualities, biological processes, functions and cell components, and
-its prime application is the unification of phenotypes on the cellular
-level across different species as well as for annotation of cellular
+phenotypes. We present a theory in which several kinds of process
+attributes can be distinguished so that normal and abnormal physiology
+of biological systems can be formally characterized. We apply this
+theory to cellular processes and create the Cell Phenotype Ontology
+(CPO). The CPO is linked to reference ontologies for qualities,
+biological processes, functions and cell components, and its prime
+application is the unification of phenotypes on the cellular level
+across different species as well as for annotation of cellular
 phenotypes in domains in which no such ontology exists.
@@
 various ontologies that describe the entities that have been affected,
-such as the various anatomical ontologies, GO \cite {Ashburner2000b},
+such as the various anatomical ontologies, GO \cite {Ashburner2000short},
 the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
@@
 description.
+
+\subsection{Axioms for physiology phenotypes}
+We implement our theory of physiology phenotypes using OWL, a formal
+language based on description logics. Using OWL, we formulate axioms
+that can be used by automated reasoners to infer additional
+information. Automated reasoning and the axioms we provide are
+intended to satisfy three aims. First, we use the axioms to infer
+information from background knowledge. In particular, we aim to
+automatically generate a class hierarchy of physiology phenotypes when
+an ontology of physiological processes, such as the GO, is provided as
+background knowledge. Our second aim is to provide interoperability
+with phenotype ontologies of other domains, including species-specific
+phenotype ontologies that have been formalized using the EQ method
+\cite{Mungall2010, Gkoutos2005}. Finally, our third aim is to query
+physiology phenotypes based on physiological processes that are
+affected or based on the way in which they are affected.
+
+Our three aims rely on the possibility of using automated reasoning
+over a resulting ontology of physiology phenotypes.  However, OWL is
+an expressive formal language, and automated reasoning in OWL has a
+high computational complexity (reasoning in OWL 2.0 is
+2NEXPTIME-complete). Consequently, due to the exponential increase in
+worst-case time complexity for automated reasoning in OWL, we would
+not be guaranteed to achieve our aims once we consider more than very
+few phenotype classes. In particular, using an ontology of the size of
+GO, with more than XX,XXX classes for processes, as a foundation for
+constructing an ontology of physiology phenotypes would not allow us
+to achieve our aims through automated reasoning.
+
+The OWL EL profile is a subset of OWL that significantly decreases the
+expressivity of OWL and the resulting time complexity of automated
+reasoning \cite{owlprofiles}. Highly efficient automated reasoners
+have been developed for OWL EL which are capable of processing very
+large ontologies \cite{Kazakov2011}. In order to achieve our aims and
+utilize automated reasoning for large ontologies, we limit ourselves
+to the OWL EL profile. As a consequence of the restriction to OWL EL,
+we cannot make use of negation ({\tt not}), disjunction ({\tt or}),
+universal quantification ({\tt forall}) and several other types of
+operations in our axioms \cite{owlprofiles}.
+
+The lack of expressivity in OWL EL requires a formulation of axioms so
+that the inferences we desire (i.e., the subclass relations resulting
+in the ontology's taxonomy) are maintained without using features of
+OWL that go beyond OWL EL expressivity. Consequently, we formulate
+{\em abnormality} and {\em absence} similarly to current
+formalizations of EQ-based phenotype ontologies, without the use of
+negation, disjunction or universal quantification
+\cite{Mungall2010}. A detailed description of the axioms we implement
+is available as supplementary material (Supplement 1).
 \end{methods}
@@
 functioning. For example, abnormalities in cell division resulting in
-{\em Aneuploidy} refer to abnormalities of {\em cell division}
-processes, while an {\em increased rate of cell division} refers to an
+{\em Aneuploidy} refer to abnormalities of {\em Cell division}
+processes, while an {\em Increased rate of cell division} refers to an
 abnormality in the pattern of occurrence of multiple cell division
 processes. In the CPO we follow the GO and represent abnormalities in
-the pattern of occurrence of $X$ as abnormalities of {\em regulation
-  of $X$} processes. In particular, an {\em increased rate of cell
-  division} is not an attribute of {\em cell division} processes, but
-rather an attribute of the {\em regulation of cell division}.
+the pattern of occurrence of $X$ as abnormalities of {\em Regulation
+  of $X$} processes.  In particular, an {\em Increased rate of cell
+  division} is not an attribute of {\em Cell division} processes, but
+rather arises from the collection of all {\em Cell division} processes
+that occur within an organism at a given time. In the
+GO, {\em Regulation of $X$} processes refer to those processes that
+determine how often and the way in which one or more $X$ processes
+occur. Therefore, we assign the attribute of {\em Increased rate of
+  cell division} to {\em Regulation of cell division} processes.
 
@@
 published in this field in the last few years \cite{Neumann2010,
-  Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
+  Schmitz2010short, Fuchs2010}, the need for CPO becomes evident.  In the
 three studies, live-cell imaging assays and RNAi knockdown were used
@@
 \bibliographystyle{plain}
+
 \bibliography{/home/leechuck/Documents/papers/bibtex/lc}

### 2012-03-18 paper/supplement.tex

\documentclass{bioinfo}
\copyrightyear{2005}
\pubyear{2005}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\renewcommand{\cite}{\citep}

\begin{document}
\firstpage{1}

\title[Physiology phenotypes]{Semantic integration of physiology
  phenotypes with an application to the Cellular Phenotype Ontology:
  Supplement 1}

\author[Hoehndorf \textit{et~al}]{Robert Hoehndorf$^{1}$\footnote{to
    whom correspondence should be addressed}, Midori
  A. Harris$^2$, Heinrich Herre$^3$, Gabriella Rustici$^4$ and
  Georgios V. Gkoutos$^{1}$}

\address{$^{1}$Department of Genetics, University of Cambridge,
  Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
  $^{2}$Department of Biochemistry; University of Cambridge, 80 Tennis
  Court Road, Cambridge CB2 1GA, UK\\
  $^{3}$Institute for Medical Informatics, Statistics and
  Epidemiology, University of Leipzig, Haertelstrasse 16-18, 04107
  Leipzig, Germany\\
  $^{4}$European Bioinformatics Institute, Wellcome Trust Genome
  Campus, Hinxton, Cambridge, Cambridge CB10 1SD, UK}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\maketitle

\section{Axiom patterns}

\subsection{Basic pattern}
Our basic pattern for formalizing phenotypes is derived from previous
work \cite{Hoehndorf2010phene}. We use the relation {\bf phenotype-of}
as a relation between a phenotype and some entity (the bearer of the
phenotype). We assume that the bearer of the phenotype is an
organism. Using the {\bf phenotype-of} relation, we further specify
the bearer's properties axiomatically. For example, a phenotype {\em
  Broad forehead} is a phenotype of an entity (the organism) that has
a nose as part which has a quality of being {\em Broad}:
\begin{verbatim}
'Broad forehead' EquivalentTo:
  phenotype-of some (has-part some
    Nose and has-quality some Broad)
\end{verbatim}

\subsection{Comparison to normal}
Phenotypes are often comparative to some reference that is considered
as {\em normal} within a context, while the phenotypes specify
deviations from this normality. One possibility to express deviation
from normality is to explicitly contextualize the classes that
represent the reference entity and either explicitly state or infer
that entities with abnormal phenotypes are distinct from the reference
entities \cite{Hoehndorf2010phene}. Being abnormal is then equivalent
to ``not being normal''. Through the explicit use of the {\bf
  has-part} relation, it becomes possible to formally include {\em
  absence} as a type of abnormality: an {\em Abnormality of X} would
be equivalent to {\em not having a normal X as part}, while the {\em
  Absence of X} is equivalent to {\em not having X as part}. The
resulting inference, that {\em Absence of X} is a subclass of {\em
  Abnormality of X}, is widely implemented in phenotype ontologies
and should be accommodated by our axiom patterns.

However, the explicit contextualization of classes with {\em normal}
and {\em abnormal} qualifiers requires the use of disjunction, and the
formalization of {\em Abnormality of X} as {\em not having a normal X
  as part} requires the use of negation. Neither operation is
supported in the OWL EL profile \cite{owlprofiles}, and since we limit
ourselves to the OWL EL profile due to its low computational
complexity, we cannot implement these patterns.  As an alternative, we
chose to implement abnormality and absence following formalization
patterns already implemented in phenotype ontologies
\cite{Mungall2010, Hoehndorf2011phenome}, since these patterns can be
expressed in OWL EL. As results, we can perform automated reasoning
over our formalizaton and benefit from interoperability with previous
work on formalizing phenotype ontologies. On the other hand, the use
of these patterns may lead to unintended consequences for some queries
\cite{Boeker2011}.  Furthermore, although we could distinguish between
{\em normal} and {\em abnormal} properties of entities using the PATO
framework, we can also omit this distinction when we build an ontology
that is solely composed of abnormal phenotypes or in which entities
only have abnormal attributes.  Based on these considerations, an
example formalization for {\em Abnormality of cell cycle} is
\begin{verbatim}
'Abnormality of X' EquivalentTo:
  phenotype-of some ( has-part some
    ( part-of some 'Cell cycle' and
      has-quality some Quality))
\end{verbatim}
The use of the class {\tt part-of some 'Cell cycle'} instead of {\tt
  'Cell cycle'} allows us to reuse the {\bf part-of} relation to infer
that abnormalities of parts of the cell cycle become subclasses of
abnormalities of the cell cycle.

\subsection{Definition of complex properties and classes}
We use several relations in order to construct complex classes. In
particular, we use the {\bf regulates} relation from the GO to define
classes of regulation processes, and we use the {\bf towards} to
specify a required second argument in some qualities. 
For example, we define the class {\em Decreased frequency of DNA
  ligation} as:
\begin{verbatim}
'Decreased frequency of DNA ligation' EquivalentTo:
  phenotype-of some (has-part some (participates-in some 
    ((has-quality some 
        ('decreased frequency'
         and (towards some 'DNA ligation')))
     and (regulates some 'DNA ligation'))))
\end{verbatim}
This definition states that a {\em Decreased frequency of DNA
  ligation} is a phenotype of organisms that have a part that
participates in a {\em Regulation of DNA ligation} process which has a
{\em Decreased frequency of DNA ligation} ({\tt 'decreased frequency'
  and (towards some 'DNA ligation')}) as quality. In this definition,
we refer to the class {\em Regulation of DNA ligation} based on its
definition {\tt regulates some 'DNA ligation'}, and to the class {\em
  Decreased frequency of DNA ligation} using the class description
{\tt 'decreased frequency' and towards some 'DNA ligation'}. Although
the use of the {\bf towards} relation in combination with a quality
such as {\em Decreased frequency} has several problems
\cite{obml2011h3}, it will lead to interoperability with the large
number of phenotype ontologies that follow a similar pattern
\cite{Mungall2010}.

\subsection{Chemical substances}
We use the ChEBI ontology \cite{Degtyarenko2007} to refer to classes
that represent chemical entities or substances. In ChEBI, no
distinction is made explicit between chemical entities, in the sense
of individual molecules, and substances composed of these entities.
In some applications, it can be necessary to distinguish between both
in ontology-based information systems or tasks that react differently
to both types (e.g., that distinguish between molecular weights and
melting points) \cite{}. However, we reuse GO axioms that utilize
ChEBI, and the GO does not make a distinction between the molecular
entity and the substance composed of it. Similarly, we do not require
such a distinction to achieve the aims of our
axiomatization. Therefore, we implicitly treat classes in ChEBI as either
molecular entities or substances composed thereof. If this distinction
is made explicit, we can define each class that we currently use from
ChEBI as a disjunction of the class representing a molecular entity
and the substance composed of the entity.

\subsection{Detailed axiom patterns}

Phenotype of $X$
\begin{verbatim}
'X phenotype' EquivalentTo:
  phenotype-of some (has-part some
    (part-of some X))
\end{verbatim}

Abnormality of $X$
\begin{verbatim}
'X phenotype' EquivalentTo:
  phenotype-of some (has-part some
    (part-of some (X and has-quality some Quality)))
\end{verbatim}

Abnormal $X$ morphology:
\begin{verbatim}
'X phenotype' EquivalentTo:
  phenotype-of some (has-part some
    (X and has-quality some Morphology))
\end{verbatim}

Abnormal $X$ physiology:
\begin{verbatim}
'X phenotype' EquivalentTo:
  phenotype-of some (has-part some
    (part-of some (X and has-quality some Functionality)))
\end{verbatim}

\bibliographystyle{plain}
\bibliography{/home/leechuck/Documents/papers/bibtex/lc}

\end{document}

### 2012-03-24 trunk/paper/supplement.tex

--- 2012-03-18 paper/supplement.tex
@@
-\documentclass{bioinfo}
-\copyrightyear{2005}
-\pubyear{2005}
-\usepackage{url}
-\usepackage[normalem]{ulem}
-\usepackage{multirow}
-\renewcommand{\cite}{\citep}
+\documentclass{article}
+\begin{document}
 
-\begin{document}
-\firstpage{1}
+\title{Supplement to Semantic integration of physiology phenotypes with an
+  application to the Cellular Phenotype Ontology}
 
-\title[Physiology phenotypes]{Semantic integration of physiology
-  phenotypes with an application to the Cellular Phenotype Ontology:
-  Supplement 1}
+\author{Robert Hoehndorf$^{1}$\footnote{to whom correspondence should
+    be addressed}, Midori A. Harris$^2$, Heinrich Herre$^3$, \\Gabriella
+  Rustici$^4$ and Georgios V. Gkoutos$^{1}$}
 
-\author[Hoehndorf \textit{et~al}]{Robert Hoehndorf$^{1}$\footnote{to
-    whom correspondence should be addressed}, Midori
-  A. Harris$^2$, Heinrich Herre$^3$, Gabriella Rustici$^4$ and
-  Georgios V. Gkoutos$^{1}$}
-
-\address{$^{1}$Department of Genetics, University of Cambridge,
+\date{$^{1}$Department of Genetics, University of Cambridge,
   Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
@@
   Campus, Hinxton, Cambridge, Cambridge CB10 1SD, UK}
-
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
-
-\editor{Associate Editor: XXXXXXX}
 
@@
 \subsection{Detailed axiom patterns}
+Here, we present the precise formulation of axiom patterns. We use $X$
+and $A$ as variables in these patterns that are replaced with specific
+classes from the GO or ChEBI ontologies.
 
-Phenotype of $X$
+\paragraph{Phenotype of X:}
+If $X$ is a material object, we assert:
 \begin{verbatim}
@@
 
-Abnormality of $X$
+If $X$ is a cellular biological process, we assert:
+\begin{verbatim}
+'X phenotype' EquivalentTo:
+  phenotype-of some (has-part some
+    (participates-in some (part-of some X)))
+\end{verbatim}
+
+\paragraph{Abnormality of X:}
+If $X$ is a material object:
 \begin{verbatim}
@@
 
+If $X$ is a process, we create a class that does not have any formal
+constraints. In CPO, abnormalities of $X$ processes include both
+abnormalities of $X$ and abnormality of regulation of $X$. A
+formulation of this constraint requires the use of disjunction. Since
+we limit ourselves to the use of OWL EL and disjunction is not a valid
+logical operator in OWL EL, we do not add any axiom to CPO in this case.
+
+\paragraph{Abnormal morphology and physiology, and absence:}
 Abnormal $X$ morphology:
@@
 
+Formalizing absence of $X$ requires the use of negation. Since
+negation is not allowed in OWL EL, we create a class for absence that
+we directly assert to be a subclass of {\em Abnormality of X}, without
+adding further axioms.
+
+\paragraph{Single occurrence abnormalities:}
+Abnormality of single occurrence of $X$:
+\begin{verbatim}
+'Abnormality of single occurrence of X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (part-of some X and
+has-quality some Quality)))
+\end{verbatim}
+
+Increased duration of X:
+\begin{verbatim}
+'Increased duration of X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (X and
+has-quality some 'Increased duration')))
+\end{verbatim}
+
+Decreased duration of X:
+\begin{verbatim}
+'Decreased duration of X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (X and
+has-quality some 'Decreased duration')))
+\end{verbatim}
+
+\paragraph{Abnormalities of X regulation:}
+Abnormality of regulation of $X$:
+\begin{verbatim}
+'Abnormality of regulation of X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (regulates some X and
+has-quality some (Quality and towards some X))))
+\end{verbatim}
+
+Increased frequency of $X$:
+\begin{verbatim}
+'Increased frequency of X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (regulates some X and
+has-quality some ('Increased frequency' and towards some X))))
+\end{verbatim}
+
+Decreased frequency of $X$:
+\begin{verbatim}
+'Decreased frequency of X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (regulates some X and
+has-quality some ('Decreased frequency' and towards some X))))
+\end{verbatim}
+
+Late onset of $X$:
+\begin{verbatim}
+'Late onset of X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (regulates some X and
+has-quality some (Delayed and towards some X))))
+\end{verbatim}
+
+Early onset of $X$:
+\begin{verbatim}
+'Early onset of X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (regulates some X and
+has-quality some (Premature and towards some X))))
+\end{verbatim}
+
+\paragraph{Abnormality of inputs and outputs (rates):}
+If, by virtue of GO-XP definitions, we are able to identify inputs and
+outputs of $X$, we define the following classes:
+
+Increased mass of $A$ as output in regulation of $X$:
+\begin{verbatim}
+'Decreased mass of A as output in regulation of X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (regulates some X and
+has-output some (A and has-quality some 'Increased mass'))))
+\end{verbatim}
+
+Decreased mass of $A$ as output in regulation of $X$:
+\begin{verbatim}
+'Decreased mass of A as output in regulation of X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (regulates some X and
+has-output some (A and has-quality some 'Decreased mass'))))
+\end{verbatim}
+
+Increased mass of $A$ as output in $X$:
+\begin{verbatim}
+'Increased mass of A as output in X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (X and
+has-output some (A and has-quality some 'Increased mass'))))
+\end{verbatim}
+
+Decreased mass of $A$ as output in $X$:
+\begin{verbatim}
+'Decreased mass of A as output in X' EquivalentTo:
+phenotype-of some (has-part some (
+participates-in some (X and
+has-output some (A and has-quality some 'Decreased mass'))))
+\end{verbatim}
+
+We assert similar axioms for the inputs of $X$.
+
 \bibliographystyle{plain}

### 2012-03-25 paper/drinking-bioinformatics.tex

@@
 morphological and physiological phenotypic characteristics of cells,
-cell components and cellular processes. Its prime application is the
-unification of cellular phenotype descriptions across species by
-providing terms and uniform definition patterns. The CPO can further
-be used for the annotation of observed abnormalities in domains, such
-as systems microscopy, in which cellular abnormalities are observed
-and for which no phenotype ontology has been created.
+cell components and cellular processes. Its prime application is to
+provide terms and uniform definition patterns for the annotation of
+cellular phenotypes. The CPO can be used for the annotation of
+observed abnormalities in domains, such as systems microscopy, in
+which cellular abnormalities are observed and for which no phenotype
+ontology has been created.
 
@@
 instance, to describe the decrease in the length of the sexual cycle
-of female animals, we can combine the PATO term \textit{decreased
+of female animals, we can combine the PATO term \textit{Decreased
   duration} ({\tt PATO:0000499}) with the Gene Ontology term
-\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
+\textit{Estrous cycle} ({\tt GO:0042698}), whilst if such a term
 existed in a pre-composed ontology (for example the {\tt MP:0009007}
@@
 few phenotype classes. In particular, using an ontology of the size of
-GO, with more than XX,XXX classes for processes, as a foundation for
+GO, with more than 22,000 classes for processes, as a foundation for
 constructing an ontology of physiology phenotypes would not allow us

### 2012-03-29 trunk/paper/drinking-bioinformatics.tex

@@
 description.
+
+\subsection{Axioms for physiology phenotypes}
+We implement our theory of physiology phenotypes using OWL, a formal
+language based on description logics. Using OWL, we formulate axioms
+that can be used by automated reasoners to infer additional
+information. Automated reasoning and the axioms we provide are
+intended to satisfy three aims. First, we use the axioms to infer
+information from background knowledge. In particular, we aim to
+automatically generate a class hierarchy of physiology phenotypes when
+an ontology of physiological processes, such as the GO, is provided as
+background knowledge. Our second aim is to provide interoperability
+with phenotype ontologies of other domains, including species-specific
+phenotype ontologies that have been formalized using the EQ method
+\cite{Mungall2010, Gkoutos2005}. Finally, our third aim is to query
+physiology phenotypes based on physiological processes that are
+affected or based on the way in which they are affected.
+
+Our three aims rely on the possibility of using automated reasoning
+over a resulting ontology of physiology phenotypes.  However, OWL is
+an expressive formal language, and automated reasoning in OWL has a
+high computational complexity (reasoning in OWL 2.0 is
+2NEXPTIME-complete). Consequently, due to the exponential increase in
+worst-case time complexity for automated reasoning in OWL, we would
+not be guaranteed to achieve our aims once we consider more than very
+few phenotype classes. In particular, using an ontology of the size of
+GO, with more than 22,000 classes for processes, as a foundation for
+constructing an ontology of physiology phenotypes would not allow us
+to achieve our aims through automated reasoning.
+
+The OWL EL profile is a subset of OWL that significantly decreases the
+expressivity of OWL and the resulting time complexity of automated
+reasoning \cite{owlprofiles}. Highly efficient automated reasoners
+have been developed for OWL EL which are capable of processing very
+large ontologies \cite{Kazakov2011}. In order to achieve our aims and
+utilize automated reasoning for large ontologies, we limit ourselves
+to the OWL EL profile. As a consequence of the restriction to OWL EL,
+we cannot make use of negation ({\tt not}), disjunction ({\tt or}),
+universal quantification ({\tt forall}) and several other types of
+operations in our axioms \cite{owlprofiles}.
+
+The lack of expressivity in OWL EL requires a formulation of axioms so
+that the inferences we desire (i.e., the subclass relations resulting
+in the ontology's taxonomy) are maintained without using features of
+OWL that go beyond OWL EL expressivity. Consequently, we formulate
+{\em abnormality} and {\em absence} similarly to current
+formalizations of EQ-based phenotype ontologies, without the use of
+negation, disjunction or universal quantification
+\cite{Mungall2010}. A detailed description of the axioms we implement
+is available as supplementary material (Supplement 1).
 \end{methods}
@@
 \section{Acknowledgement}
+We thank Michel Dumontier for valuable discussions about the formal
+representation of the axioms.
+
 Funding for RH is provided by the European Commission's 7th Framework

### 2012-04-02 paper/drinking-bioinformatics.tex

@@
 rather arises from the collection of all {\em Cell division} processes
-that occur within an organism at a given time. In the
-GO, {\em Regulation of $X$} processes refer to those processes that
-determine how often and the way in which one or more $X$ processes
-occur. Therefore, we assign the attribute of {\em Increased rate of
-  cell division} to {\em Regulation of cell division} processes.
+that occur within an organism at a given time. In the GO, {\em
+  Regulation of $X$} processes refer to those processes that determine
+how often and in which way one or more $X$ processes occur. Therefore,
+we assign the attribute of {\em Increased rate of cell division} to
+{\em Regulation of cell division} processes.
 

### 2012-04-02 trunk/paper/response-bioinformatics.txt

Dear Dr. Bateman,

Thank you for providing us with the opportunity to revise our
manuscript. We have now prepared a revision in which we have addressed
all reviewers' comments. In response to both reviewers, we have
removed the discussion of the heart beating example from the
introduction and now provide a discussion of the axioms in the
manuscript as well as a detailed presentation of the axioms as
supplementary material. We have included a response to the individual
points raised by the reviewers below.  We would like to thank the
reviewers for their valuable comments which allowed us to improve our
manuscript.

Please do not hesitate to contact us if you require any further
information.

With best wishes,
Robert.

Responses to reviewers
----------------------

Reviewer 1:
"Introduction: I find the heart beating example difficult to follow,
primarily because it lacks simple easy-to-understand definitions - so
either this section is removed or needs to be rewritten (the result
section, however, is much clearer)."

Response: We have removed the section on heart beating from the
introduction, as recommended by the reviewer.

Reviewer 1:
p4 - On the one hand, I agree with "abnormalities in cell division
resulting in Aneuploidy refer to abnormalities of cell division
processes, while an increased rate of cell division refers to an
abnormality in the pattern of occurrence of multiple cell division
processes". However, you then state "an increased rate of cell
division is not an attribute of cell division processes, but rather an
attribute of the regulation of cell division"
two questions: 
1. is the regulation of a process(es) not different from that
collection of processes?
2. is it not the case that a collection of processes has attributes
based on its parts?

Response:
We agree with the reviewer that "regulation" processes do not always
(or commonly) include the processes that are being regulated as
parts, and we have modified our manuscript in this regard, providing
more details on regulation processes in GO and their relation to
collection processes.

Reviewer 1:
This paper needs to discuss the axiom patterns that generates the ontology.

Response:
We agree with the reviewer, and now provide information about the
axioms both in the manuscript (new section 2.3 "Axioms for physiology
phenotypes") and full details as supplementary material.

Reviewer 1:
abstract: suggests that its application is in the unification of
cellular phenotypes across species; but it's not clear where
species-specific differentiation would come into play. it also
suggests (and is further outlined in the discussion) that it could be
ussed for systems microscopy, but no evidence is provided to support
the claim. And the latter, really must be further developed, to
demonstrate that the ontology can be used for some application.

Response: 
The Cell Phenotype Ontology, which we have built by applying our
ontology-patterns for representing physiology phenotypes, is currently
being applied for the annotation of cellular observations in systems
microscopy. For this purpose, CPO provides the necessary terms (in the
sense of a controlled vocabulary) as well as a taxonomic structure
that is used for basic retrieval operations.

We agree with the reviewers that an application of our work in a
bioinformatics analysis would be highly beneficial. For example, once
sufficient data is annotated with the CPO, we can use its taxonomic
structure for enrichment analyses. Similarly, once more cellular
phenotypes are being annotated in various model organism databases, we
can utilize CPO for integrative phenotype analyses across
species. However, at the moment, no comprehensive ontology of
physiology phenotypes is available, partly due to the complexity of
the domain and partly due to limitations in available ontology
frameworks. We see our main contribution in the development of a
theory that allows to capture physiology phenotypes; we applied this
theory to construct (largely automatically) the CPO, which is now
being used to annotate cellular phenotypes in the systems microscopy
domain. Based on our contributions, it is now possible to annotate
and eventually integrate (through the use of a controlled vocabulary
of physiology phenotypes, of which the CPO provides a first example)
data in the domain of physiology. At the same time, we maintain
compatibility with GO (on which a majority of physiology phenotypes
are based) and other PATO-based phenotype ontologies, thereby filling
a gap that has limited data integration in the physiology domain so
far.

Reviewer 2:
A theory of physiology related phenotypes and an implementation based
on GO, GO cross-products and PATO is presented. An outline of a theory
of phenotypes is introduced which participate-in a set of pathological
processes which can be qualified by a small set of PATO qualities. At
least, for a majority of heterogeneous phenotype descriptions this can
be an approach for unification. In future work a sound formal
characterization has and an evaluation of the practical value of the
approach has to be delivered.

Response:
We agree, and have added a detailed formal presentation of our axioms
as supplementary material.

### 2012-04-02 trunk/paper/drinking-bioinformatics.tex

@@
 first are process attributes that arise directly from processes and
-include {\em duration} and {\em temporal location}; the second are
+include {\em Duration} and {\em Temporal location}; the second are
 attributes that arise from processes and their temporal parts and
-include {\em frequency} and {\em onset}; and the third are attributes
+include {\em Frequency} and {\em Onset}; and the third are attributes
 that arise from processes and qualities of their participants, and
-include {\em flow rates}.
+include {\em Flow rate}s.
 
@@
 relate attributes of process participants with the duration of a
-process. In particular, a {\em rate} typically refers to an attribute
+process. In particular, a {\em Rate} typically refers to an attribute
 of some entity {\em with respect to an attribute of another entity},
@@
 process participant with respect to the duration of the process. For
-example, a {\em mass flow rate} refers to the {\em Mass} of a process
-participant with respect to the duration of the process, i.e., how
-much matter is moved (from one point to another) through the process.
-As a more complex example, a {\em rate of change of position} refers
-to the {\em distance} that an object is moved with respect to the
-duration of the process.
+example, a {\em Mass flow rate} refers to the {\em Mass} of a process
+participant with respect to the {\em Duration} of the process, i.e.,
+how much matter is moved (from one point to another) through the
+process.  As a more complex example, a {\em Rate of change of
+  position} refers to the {\em Distance} that an object is moved with
+respect to the {\em Duration} of the process.
 
 However, not all rates of a process depend on attributes of the
-process participants. In particular, a {\em frequency of occurrence}
-or {\em event rate} refers to the number of occurrences of a type of
-process during a reference process. For example, a {\em rate of heart
+process participants. In particular, a {\em Frequency of occurrence}
+or {\em Event rate} refers to the number of occurrences of a type of
+process during a reference process. For example, a {\em Rate of heart
   beating} refers to the number of {\em Heart beating} processes that
@@
 a particular type are distributed within a reference process. For
-example, the heart may beat {\em rhythmically} or {\em arrhythmically}
-within a period of time (see Figure \ref{fig:patterns}).
+example, the heart may beat rhythmically or arrhythmically within a
+period of time (see Figure \ref{fig:patterns}).
 \begin{figure}
@@
     non-comparative attributes. The first process has an attribute of
-    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
+    {\em Rhythmic occurrence of P} because the instances of $P$ are
     temporally equidistantly distributed. The second example shows an
-    {\em arrhythmic} occurrence of $P$, and the third examples shows
-    an {\em increasing frequency} (of $P$). A regulation process with
-    an increasing frequency (of $P$) is a process in which the
-    frequency of occurrences of $P$ is lower in the first half of the
-    process than in the second half. The right side of the figure
-    illustrates comparative phenotypic descriptions of processes. On
-    the upper right, the {\em normal} reference is shown. The second
-    example illustrates a {\em late onset} of $P$, i.e., the attribute
-    that $P$ processes begin later than {\em normal}. Finally, the
-    lower right illustrates a {\em decreased frequency} (of $P$),
-    since fewer processes of the type $P$ occur within the reference
-    process than {\em normal}.}
+    {\em Arrhythmic occurrence of P}, and the third examples shows an
+    {\em Increasing frequency of P}. A regulation process with an {\em
+      Increasing frequency of P} attribute is a process in which the
+    value of the {\em Frequency of occurrences of P} attribute is
+    lower in the first half of the process than in the second
+    half. The right side of the figure illustrates comparative
+    phenotypic descriptions of processes. On the upper right, the {\em
+      normal} reference process is shown. The second example
+    illustrates a {\em Late onset of P}, i.e., the attribute that $P$
+    processes begin later than {\em normal} processes. Finally, the
+    lower right illustrates a {\em Decreased frequency of P}, since
+    fewer processes of the type $P$ occur within the reference process
+    than {\em normal}.}
 \end{figure}
@@
 % all temporal parts of $p$ are also instances of $C$. 
-Examples of processes that can be divided include {\em continuous
-  movements} or {\em mass flow} processes, for which all parts have a
-{\em speed} or {\em flow rate} attribute. On the other hand, some
+Examples of processes that can be divided include {\em Continuous
+  movement} or {\em Mass flow} processes, for which all parts have a
+{\em Speed} or {\em Flow rate} attribute. On the other hand, some
 processes can be subdivided into stages of activity and stages of
-inactivity and cannot arbitrarily be divided. For example, a process
-of {\em heart beating} has periods of activity (a single heart beat)
-and inactivity. Consequently, not all parts of the process have a {\em
-  heart rate} attribute.
+inactivity (with respect to a particular process type) and cannot
+arbitrarily be divided. For example, a process of {\em Heart beating}
+has periods of activity in which a heart beat occurs and periods in
+which no heart beat process occurs. Consequently, not all parts of the
+process have a {\em Heart rate} ({\em Rate of heart beating})
+attribute.
 
-We may further attribute a {\em frequency} or {\em rate} to an object
+We may further attribute a {\em Frequency} or {\em Rate} to an object
 instead of a process. For example, a heart that beats {\em now} with a
@@
 as observed with a speed camera) can be considered attributes of the
-objects (the heart or the car), not attributes of the processes in
+objects (the heart, or the car), not attributes of the processes in
 which the objects participate. However, these are {\em different}
@@
 participates. Such a reference process is necessary in order to obtain
-a value for a frequency even when no {\em heart beating} process is
+a value for a frequency even when no {\em Heart beating} process is
 occurring. However, the frequency is only an attribute of the heart in
-virtue of such a reference process in which {\em heart beating} is
+virtue of such a reference process in which {\em Heart beating} is
 actually occurring.  This reference process can be uniquely determined
-for processes such as {\em continuous movement}, where the rate of an
+for processes such as {\em Continuous movement}, where the rate of an
 object at a time $t$ is the rate of the infinitesimally small process
 that occurs around $t$. The reference process is ambiguous for
-processes such as {\em heart beating}, and the reference process must
+processes such as {\em Heart beating}, and the reference process must
 be explicitly stated.
@@
 Amongst the abnormal phenotypes that we include for all cell
-components listed in GO-CC, we distinguish {\em abnormal morphology}
-and {\em abnormal physiology} phenotypes. An {\em Abnormal morphology
+components listed in GO-CC, we distinguish {\em Abnormal morphology}
+and {\em Abnormal physiology} phenotypes. An {\em Abnormal morphology
   of $C$} is either the (abnormal) absence of required parts of $C$,
 the (abnormal) presence of additional parts, or abnormal qualities of
-$C$ \cite{Hoehndorf2010phene}. For example, an absence of caveolae
-({\tt MP:0004150}) would be a subclass of {\em Abnormal morphology of
-  plasma membrane} in virtue of caveolae necessarily being part of the
-{\em Plasma membrane} ({\tt GO:0005886}).
+$C$ \cite{Hoehndorf2010phene}. For example, an {\em Absence of
+  caveolae} ({\tt MP:0004150}) would be a subclass of {\em Abnormal
+  morphology of plasma membrane} in virtue of caveolae necessarily
+being part of the {\em Plasma membrane} ({\tt GO:0005886}).
 
@@
 %
-First, common to all processes is the quality of {\em duration} and
+First, common to all processes is the quality of {\em Duration} and
 consequently, each process can have an {\em abnormal} (increased or
@@
 process participants in relation to the duration of the process. These
-include all kinds of {\em rates} such as {\em mass flow rate}, {\em
-  energy flow rate} and {\em velocity} (the rate of change of
+include all kinds of {\em rates} such as {\em Mass flow rate}, {\em
+  Energy flow rate} and {\em Velocity} (the rate of change of
 position). In each of these cases, an object participates in a process
@@
   streaming} can be defined as an increased amount of inputs or an
-increased amount of outputs of a {\em cytoplasmic streaming} process.
+increased amount of outputs of a {\em Cytoplasmic streaming} process.
 
@@
 convention, we call an abnormality of multiple occurrences of $X$ {\em
-  abnormality of the regulation of $X$}.
+  Abnormality of the regulation of $X$}.
 
-\begin{table*}
-  \centering
-  \begin{tabular}{p{2.8cm}|p{4cm}|p{4cm}|p{4cm}}
-    & Increased flow rate & Normal flow rate & Decreased
-    flow rate \\
-    \hline
-    Increased frequency &increased total flow rate &increased total
-    flow rate &?\\
-    Normal frequency &increased total flow rate &normal total flow
-    rate &decreased total flow rate\\
-    Decreased frequency &?&decreased total flow rate &decreased total
-    flow rate\\
-    \hline
-  \end{tabular}
-  \caption{\label{tbl:flow}Interdependency for the attribute {\em
-      Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
-      rate} is an attribute of {\em Regulation of cytoplasmic
-      streaming} processes, while {\em Cytoplasmic flow rate} is an
-    attribute of individual {\em cytoplasmic streaming}
-    processes. Depending both on whether the cytoplasmic flow rate in
-    individual {\em cytoplasmic streaming} processes is increased or
-    decreased and whether the frequency of occurrence of {\em
-      cytoplasmic streaming} is increased or decreased, the total
-    cytoplasmic flow rate can be increased or decreased.}
-\end{table*}
+% \begin{table*}
+%   \centering
+%   \begin{tabular}{p{2.8cm}|p{4cm}|p{4cm}|p{4cm}}
+%     & Increased flow rate & Normal flow rate & Decreased
+%     flow rate \\
+%     \hline
+%     Increased frequency &increased total flow rate &increased total
+%     flow rate &?\\
+%     Normal frequency &increased total flow rate &normal total flow
+%     rate &decreased total flow rate\\
+%     Decreased frequency &?&decreased total flow rate &decreased total
+%     flow rate\\
+%     \hline
+%   \end{tabular}
+%   \caption{\label{tbl:flow}Interdependency for the attribute {\em
+%       Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
+%       rate} is an attribute of {\em Regulation of cytoplasmic
+%       streaming} processes, while {\em Cytoplasmic flow rate} is an
+%     attribute of individual {\em cytoplasmic streaming}
+%     processes. Depending both on whether the cytoplasmic flow rate in
+%     individual {\em cytoplasmic streaming} processes is increased or
+%     decreased and whether the frequency of occurrence of {\em
+%       cytoplasmic streaming} is increased or decreased, the total
+%     cytoplasmic flow rate can be increased or decreased.}
+% \end{table*}
 
@@
 defined as an abnormality of {\em Regulation of apoptosis} ({\tt
-  GO:0042981}) with respect to the {\em frequency} ({\tt
+  GO:0042981}) with respect to the {\em Frequency} ({\tt
   PATO:0000044}) of {\em Apoptosis} ({\tt GO:0006915}) occurrences.
@@
   onset} of a process. With respect to a reference process, a
-particular kind of process may be {\em delayed} ({\tt PATO:0000502})
-or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
+particular kind of process may be {\em Delayed} ({\tt PATO:0000502})
+or {\em Premature} ({\tt PATO:0000694}). For example, {\em Delayed
   apoptosis} refers to an abnormality of the {\em Regulation of
   apoptosis} in which apoptosis is induced later than normal.  We use
-the PATO quality {\em onset} ({\tt PATO:0002325}) and its children
-{\em delayed} and {\em premature} to define these types of regulatory
-abnormality. Similarly, we use PATO's {\em offset} ({\tt
+the PATO quality {\em Onset} ({\tt PATO:0002325}) and its children
+{\em Delayed} and {\em Premature} to define these types of regulatory
+abnormality. Similarly, we use PATO's {\em Offset} ({\tt
   PATO:0002324}) quality and its children to characterize regulatory
@@
 regulated. For example, a cytoplasmic flow rate can be increased or
-decreased not within a single {\em cytoplasmic streaming} process but
+decreased not within a single {\em Cytoplasmic streaming} process but
 rather the total cytoplasmic flow rate, as a summation over all
@@
 particular anatomical location), is increased or decreased. While a
-flow rate of a single {\em cytoplasmic streaming} process is a quality
+flow rate of a single {\em Cytoplasmic streaming} process is a quality
 of that process, an increased {\em total} cytoplasmic flow rate is a
-quality of the regulation of {\em cytoplasmic streaming}. In
+quality of the regulation of {\em Cytoplasmic streaming}. In
 particular, it is possible for an organism to have a normal --- or
@@
 increased cytoplasmic flow rate in each individual cytoplasmic
-streaming process. Table \ref{tbl:flow} illustrates the dependencies
-between rates of individual processes, their frequency of occurrence
-and the total rate of these processes. We include total rates as
-regulatory abnormalities in the CPO since these are the attributes of
-processes that are often measured or observed, while the rates of
-individual processes are inferred following a schema such as Table
-\ref{tbl:flow}.
+streaming process. 
+% A {\em Total cytoplasmic flow rate} is an attribute of {\em Regulation
+%   of cytoplasmic streaming} processes, while {\em Cytoplasmic flow
+%   rate} is an attribute of individual {\em cytoplasmic streaming}
+% processes. Depending both on whether the cytoplasmic flow rate in
+% individual {\em cytoplasmic streaming} processes is increased or
+% decreased and whether the frequency of occurrence of {\em cytoplasmic
+%   streaming} is increased or decreased, the total cytoplasmic flow
+% rate can be increased or decreased.
+We include total rates as subclasses of regulatory abnormalities in
+the CPO since these are the attributes of processes that are often
+measured or observed, while the rates of individual processes are
+inferred.
 
@@
 qualities in PATO, including {\em normal}, {\em abnormal}, and several
-process qualities including {\em increased duration} and {\em
-  decreased occurrence}. FYPO will thus fit neatly under the CPO
+process qualities including {\em Increased duration} and {\em
+  Decreased occurrence}. FYPO will thus fit neatly under the CPO
 umbrella, and stands to benefit from the automated synchronization
@@
 integration can be achieved by further developing CPO.
-
 This ontology will be used to integrate phenotypes' definitions across
@@
 We thank Michel Dumontier for valuable discussions about the formal
-representation of the axioms.
+representation of the axioms and the different possible
+interpretations of the {\bf regulates} relation.
 

### 2012-04-02 paper/response-bioinformatics.txt

--- 2012-04-02 trunk/paper/response-bioinformatics.txt
@@
 Response:
-We agree, and have added a detailed formal presentation of our axioms
-as supplementary material.
+We have now added a detailed formal presentation of our axioms as
+supplementary material. An evaluation of the practical utility depends
+on data that is being annotated with physiology phenotypes. Currently,
+no comprehensive ontology of physiology phenotypes is available. Our
+work brings the immediate benefit that the availability of CPO enables
+annotation of cellular phenotypes in the systems microscropy
+community, and our method can be applied to physiology phenotypes in
+other domains, in particular computational modelling.
+
+Reviewer 2:
+Minor revision (page 2 line 27): You introduce a Heart beating example
+to illustrate the challenges in modeling processes and their
+regulation. [...] The Heart example is illustrative but limited with
+respect to your ontology. There, both the the function of the single
+entity can be impaired and the regulation of multiple occurences of
+processes (where the example ends due to the single heartedness of
+many animals).
+
+Response: 
+In response to this comment and reviewer 1's comment, we have removed
+the Heart beating example from the introduction and focus on the
+detailed presentation of the processes involved in the Results section
+of our manuscript.
+
+Reviewer 2:
+Minor revision (page 3 second column line 33) see as above on the
+phases of a single Heart action. The heart has no phase of
+“inactivity” (that would be death), only a number of phases of which
+the filling/ ejection phases are the least granular.
+
+Response: We have changed our manuscript
+
+Reviewer 2:
+Minor revision (page 5 Table 1): make it clear that you refer to the
+product of row and column quantitites in the cells. I find it
+difficult to understand this easy relation from your legend. Do you
+really need this table?
+
+Response:
+We have removed the table from our manuscript, and discuss the
+content of the table now in the main text.
+
+Reviewer 2:
+Class names should be consistently written italic and uppercase
+throughout the manuscript (e.g. Heart beating and Rate of heart
+beating)
+
+Response:
+We have made this change throughout our manuscript.

### 2012-04-04 trunk/paper/response-bioinformatics.txt

@@
 Response:
-We agree, and have added a detailed formal presentation of our axioms
-as supplementary material.
+We have now added a detailed formal presentation of our axioms as
+supplementary material. An evaluation of the practical utility depends
+on data that is being annotated with physiology phenotypes. Currently,
+no comprehensive ontology of physiology phenotypes is available. Our
+work brings the immediate benefit that the availability of CPO enables
+annotation of cellular phenotypes in the systems microscropy
+community, and our method can be applied to physiology phenotypes in
+other domains, in particular computational modelling.
+
+Reviewer 2:
+Minor revision (page 2 line 27): You introduce a Heart beating example
+to illustrate the challenges in modeling processes and their
+regulation. [...] The Heart example is illustrative but limited with
+respect to your ontology. There, both the the function of the single
+entity can be impaired and the regulation of multiple occurences of
+processes (where the example ends due to the single heartedness of
+many animals).
+
+Response: 
+In response to this comment and reviewer 1's comment, we have removed
+the Heart beating example from the introduction and focus on the
+detailed presentation of the processes involved in the Results section
+of our manuscript.
+
+Reviewer 2:
+Minor revision (page 3 second column line 33) see as above on the
+phases of a single Heart action. The heart has no phase of
+“inactivity” (that would be death), only a number of phases of which
+the filling/ ejection phases are the least granular.
+
+Response: We have changed our manuscript, according to the reviewer's
+suggestion.
+
+Reviewer 2:
+Minor revision (page 5 Table 1): make it clear that you refer to the
+product of row and column quantitites in the cells. I find it
+difficult to understand this easy relation from your legend. Do you
+really need this table?
+
+Response:
+We have removed the table from our manuscript, and discuss the
+content of the table now in the main text.
+
+Reviewer 2:
+Class names should be consistently written italic and uppercase
+throughout the manuscript (e.g. Heart beating and Rate of heart
+beating)
+
+Response:
+We have made this change throughout our manuscript.

### 2012-04-04 trunk/paper/drinking-bioinformatics.tex

@@
 \bibliographystyle{plain}
-
-\bibliography{/home/leechuck/Documents/papers/bibtex/lc}
+\begin{thebibliography}{10}
+
+\bibitem{Ashburner2000short}
+Michael Ashburner et~al.
+\newblock Gene ontology: tool for the unification of biology.
+\newblock {\em Nature Genetics}, 25(1), May 2000.
+
+\bibitem{Bada2007a}
+M.~Bada and L.~Hunter.
+\newblock Enrichment of obo ontologies.
+\newblock {\em Journal of Biomedical Informatics}, 40(3):300--315, June 2007.
+
+\bibitem{Bada2004}
+Michael Bada, Robert Stevens, Carole Goble, Yolanda Gil, Michael Ashburner,
+  Judith~A. Blake, Michael~J. Cherry, Midori Harris, and Suzanna Lewis.
+\newblock A short study on the success of the gene ontology.
+\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
+  Web}, 1(2):235--240, February 2004.
+
+\bibitem{Bard2005}
+Jonathan Bard, Seung~Y. Rhee, and Michael Ashburner.
+\newblock An ontology for cell types.
+\newblock {\em Genome Biology}, 6(2), 2005.
+
+\bibitem{Berners-Lee2001}
+T.~Berners-Lee, J.~Hendler, O.~Lassila, et~al.
+\newblock {The Semantic Web}.
+\newblock {\em Scientific American}, 284(5):28--37, 2001.
+
+\bibitem{Conrad2011}
+Christian Conrad, Annelie W\"{u}nsche, Tze Heng~H. Tan, Jutta Bulkescher, Frank
+  Sieckmann, Fatima Verissimo, Arthur Edelstein, Thomas Walter, Urban Liebel,
+  Rainer Pepperkok, and Jan Ellenberg.
+\newblock Micropilot: automation of fluorescence microscopy-based imaging for
+  systems biology.
+\newblock {\em Nature methods}, 8(3):246--249, March 2011.
+
+\bibitem{opb}
+Daniel~L. Cook, Fred~L. Bookstein, and John~H. Gennari.
+\newblock Physical properties of biological entities: An introduction to the
+  ontology of physics for biology.
+\newblock {\em PLoS ONE}, 6(12):e28708, 12 2011.
+
+\bibitem{Eilbeck2005}
+K.~Eilbeck, S.~E. Lewis, C.~J. Mungall, M.~Yandell, L.~Stein, R.~Durbin, and
+  M.~Ashburner.
+\newblock The sequence ontology: A tool for the unification of genome
+  annotations.
+\newblock {\em Genome Biology}, 6(R55), 2005.
+
+\bibitem{aposhort}
+Stacia~R. Engel et~al.
+\newblock Saccharomyces genome database provides mutant phenotype data.
+\newblock {\em Nucleic acids research}, 38(Database issue), January 2010.
+
+\bibitem{Fuchs2010}
+Florian Fuchs, Gregoire Pau, Dominique Kranz, Oleg Sklyar, Christoph Budjan,
+  Sandra Steinbrink, Thomas Horn, Angelika Pedal, Wolfgang Huber, and Michael
+  Boutros.
+\newblock Clustering phenotype populations by genome-wide {RNAi} and
+  multiparametric imaging.
+\newblock {\em Molecular Systems Biology}, 6, June 2010.
+
+\bibitem{go2010}
+{Gene Ontology Consortium}.
+\newblock The gene ontology in 2010: extensions and refinements.
+\newblock {\em Nucleic acids research}, 38(Database issue):D331--335, January
+  2010.
+
+\bibitem{Gkoutos2005}
+Georgios~V. Gkoutos, Eain~C. Green, Ann-Marie~M. Mallon, John~M. Hancock, and
+  Duncan Davidson.
+\newblock {Using ontologies to describe mouse phenotypes.}
+\newblock {\em Genome biology}, 6(1), 2005.
+
+\bibitem{obml2011h1}
+Georgios~V. Gkoutos and Robert Hoehndorf.
+\newblock Ontology-based cross-species integration and analysis of
+  saccharomyces cerevisiae phenotypes.
+\newblock In {\em Proceedings of the 3rd Workshop for Ontologies in Biomedicine
+  and Life sciences (OBML)}, October 2011.
+
+\bibitem{Gkoutos2009b}
+Georgios~V. Gkoutos, Chris Mungall, Sandra Dolken, Michael Ashburner, Suzanna
+  Lewis, John Hancock, Paul Schofield, Sebastian Kohler, and Peter~N. Robinson.
+\newblock Entity/quality-based logical definitions for the human skeletal
+  phenome using {PATO}.
+\newblock {\em Annual International Conference of the IEEE Engineering in
+  Medicine and Biology Society.}, 1:7069--7072, 2009.
+
+\bibitem{goble}
+C.~Goble and R.~Stevens.
+\newblock State of the nation in data integration for bioinformatics.
+\newblock {\em Journal of Biomedical Informatics}, 41(5):687--693, 10 2008.
+
+\bibitem{Grau2008}
+B.~Grau, I.~Horrocks, B.~Motik, B.~Parsia, P.~Patelschneider, and U.~Sattler.
+\newblock {OWL} 2: The next step for {OWL}.
+\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
+  Web}, 6(4):309--322, November 2008.
+
+\bibitem{Gruber1995}
+Thomas~R. Gruber.
+\newblock Toward principles for the design of ontologies used for knowledge
+  sharing.
+\newblock {\em International Journal of Human-Computer Studies}, 43(5-6), 1995.
+
+\bibitem{Guarino1998}
+Nicola Guarino.
+\newblock Formal ontology and information systems.
+\newblock In Nicola Guarino, editor, {\em Proceedings of the 1st International
+  Conference on Formal Ontologies in Information Systems}, pages 3--15. IOS
+  Press, 1998.
+
+\bibitem{Herre2006}
+Heinrich Herre, Barbara Heller, Patryk Burek, Robert Hoehndorf, Frank Loebe,
+  and Hannes Michalek.
+\newblock {G}eneral {F}ormal {O}ntology ({GFO}) -- {A} foundational ontology
+  integrating objects and processes [{V}ersion 1.0].
+\newblock Onto-Med Report~8, IMISE, University of Leipzig, Leipzig, Germany,
+  2006.
+
+\bibitem{Hoehndorf2010patterns}
+Robert Hoehndorf, Anika Oellrich, Michel Dumontier, Janet Kelso, Dietrich
+  Rebholz-Schuhmann, and Heinrich Herre.
+\newblock Relations as patterns: Bridging the gap between {OBO} and {OWL}.
+\newblock {\em BMC Bioinformatics}, 11(1):441+, 2010.
+
+\bibitem{Hoehndorf2010phene}
+Robert Hoehndorf, Anika Oellrich, and Dietrich Rebholz-Schuhmann.
+\newblock Interoperability between phenotype and anatomy ontologies.
+\newblock {\em Bioinformatics}, 26(24):3112 -- 3118, 10 2010.
+
+\bibitem{Horridge2007}
+Matthew Horridge, Sean Bechhofer, and Olaf Noppens.
+\newblock Igniting the {OWL} 1.1 touch paper: The {OWL} {API}.
+\newblock In {\em Proceedings of OWLED 2007: Third International Workshop on
+  OWL Experiences and Directions}, 2007.
+
+\bibitem{Horrocks2007}
+Ian Horrocks.
+\newblock {OBO} flat file format syntax and semantics and mapping to {OWL}
+  {W}eb {O}ntology {L}anguage.
+\newblock Technical report, University of Manchester, March 2007.
+\newblock \url{http://www.cs.man.ac.uk/~horrocks/obo/}.
+
+\bibitem{Kazakov2011}
+Yevgeny Kazakov, Markus Kr{\"o}tzsch, and Franti\v{s}ek Siman\v{c}\'{i}k.
+\newblock Unchain my $\mathcal{EL}$ reasoner.
+\newblock In {\em Proceedings of the 23rd International Workshop on Description
+  Logics (DL'10)}, CEUR Workshop Proceedings. CEUR-WS.org, 2011.
+
+\bibitem{Lock2010}
+John~G Lock and Staffan Strömblad.
+\newblock Systems microscopy: an emerging strategy for the life sciences.
+\newblock {\em Experimental Cell Research}, 316(8):1438--1444, 2010.
+
+\bibitem{owlprofiles}
+Boris Motik, Bernardo~Cuenca Grau, Ian Horrocks, Zhe Wu, Achille Fokoue, and
+  Carsten Lutz.
+\newblock Owl 2 web ontology language: Profiles.
+\newblock Recommendation, World Wide Web Consortium (W3C), 2009.
+
+\bibitem{Mungall2010}
+Christopher Mungall, Georgios Gkoutos, Cynthia Smith, Melissa Haendel, Suzanna
+  Lewis, and Michael Ashburner.
+\newblock Integrating phenotype ontologies across multiple species.
+\newblock {\em Genome Biology}, 11(1):R2+, 2010.
+
+\bibitem{Mungall2010go}
+Christopher~J. Mungall, Michael Bada, Tanya~Z. Berardini, Jennifer Deegan,
+  Amelia Ireland, Midori~A. Harris, David~P. Hill, and Jane Lomax.
+\newblock Cross-product extensions of the gene ontology.
+\newblock {\em Journal of biomedical informatics}, February 2010.
+\newblock in press.
+
+\bibitem{Neumann2010}
+Beate Neumann, Thomas Walter, Jean-Karim Hériché, Jutta Bulkescher, Holger
+  Erfle, Christian Conrad, Phill Rogers, Ina Poser, Michael Held, Urban Liebel,
+  and et~al.
+\newblock Phenotypic profiling of the human genome by time-lapse microscopy
+  reveals cell division genes.
+\newblock {\em Nature}, 464(7289):721--727, 2010.
+
+\bibitem{Ogren2004}
+P.~V. Ogren, K.~B. Cohen, G.~K. Acquaah-Mensah, J.~Eberlein, and L.~Hunter.
+\newblock The compositional structure of gene ontology terms.
+\newblock {\em Pac Symp Biocomput}, pages 214--225, 2004.
+
+\bibitem{Gruninger2010}
+Atalay \"{O}zg\"{o}vde and Michael Gr\"{u}ninger.
+\newblock Foundational process relations in bio-ontologies.
+\newblock In {\em Proceeding of the 2010 conference on Formal Ontology in
+  Information Systems}, pages 243--256, Amsterdam, The Netherlands, The
+  Netherlands, 2010. IOS Press.
+
+\bibitem{Robinson2008}
+P.~N. Robinson, S.~Koehler, S.~Bauer, D.~Seelow, D.~Horn, and S.~Mundlos.
+\newblock The human phenotype ontology: a tool for annotating and analyzing
+  human hereditary disease.
+\newblock {\em American journal of human genetics}, 83(5):610--615, 2008.
+
+\bibitem{wpo}
+Gary Schindelman, Jolene Fernandes, Carol Bastiani, Karen Yook, and Paul
+  Sternberg.
+\newblock Worm phenotype ontology: integrating phenotype data within and beyond
+  the c. elegans community.
+\newblock {\em BMC Bioinformatics}, 12(1):32, 2011.
+
+\bibitem{Schmitz2010short}
+Michael H.~A. Schmitz et~al.
+\newblock Live-cell imaging rnai screen identifies pp2a–b55$\alpha$ and
+  importin-$\beta$1 as key mitotic exit regulators in human cells.
+\newblock {\em Nature Cell Biology}, 12:886--893, 2010.
+
+\bibitem{Schofield2011}
+Paul~N. Schofield, John~P. Sundberg, Robert Hoehndorf, and Georgios~V. Gkoutos.
+\newblock New approaches to the representation and analysis of phenotype
+  knowledge in human diseases and their animal models.
+\newblock {\em Briefings in Functional Genomics}, 10(5):258--265, 2011.
+
+\bibitem{Smith2007short}
+Barry Smith et~al.
+\newblock The {OBO} {F}oundry: coordinated evolution of ontologies to support
+  biomedical data integration.
+\newblock {\em Nat Biotech}, 25(11):1251--1255, 2007.
+
+\bibitem{Smith2004}
+Cynthia~L. Smith, Carroll-Ann~W. Goldsmith, and Janan~T. Eppig.
+\newblock The mammalian phenotype ontology as a tool for annotating, analyzing
+  and comparing phenotypic information.
+\newblock {\em Genome Biology}, 6(1):R7, 2004.
+
+\end{thebibliography}
+
+% \bibliography{/home/leechuck/Documents/papers/bibtex/lc}
 

### 2012-04-04 paper/response-bioinformatics.txt

@@
 
-Thank you for providing us with the opportunity to revise our
+Thank you very much for providing us with the opportunity to revise our
 manuscript. We have now prepared a revision in which we have addressed
@@
 
-With best wishes,
-Robert.
+With kind regards,
+Robert
 
@@
 (or commonly) include the processes that are being regulated as
-parts, and we have modified our manuscript in this regard, providing
+parts. We have modified our manuscript in this regard and we now provide
 more details on regulation processes in GO and their relation to
@@
 Response:
-We agree with the reviewer, and now provide information about the
+We agree with the reviewer and in response we now provide information about the
 axioms both in the manuscript (new section 2.3 "Axioms for physiology
-phenotypes") and full details as supplementary material.
+phenotypes"). We also provide the full details of all axioms as supplementary material.
 
@@
 
-We agree with the reviewers that an application of our work in a
+We agree with the reviewer that an application of our work in a
 bioinformatics analysis would be highly beneficial. For example, once
@@
 are based) and other PATO-based phenotype ontologies, thereby filling
-a gap that has limited data integration in the physiology domain so
+a gap that has limited data integration in the cellular physiology domain so
 far.
@@
 Response: 
-In response to this comment and reviewer 1's comment, we have removed
+In response to this comment as well as the similar comment reviewer 1 made, we have removed
 the Heart beating example from the introduction and focus on the
@@
 
-Response: We have changed our manuscript
+Response: We agree with the reviewer and we have changed our manuscript, according to his/her
+suggestion.
 
@@
 Response:
-We have removed the table from our manuscript, and discuss the
-content of the table now in the main text.
+We thank the reviewer for pointing this out and in response, we have removed the table from our manuscript. In the revised manuscript, we discuss the
+content of the table in the main text.
 

### 2012-04-05 paper/drinking-bioinformatics.tex

@@
 % all temporal parts of $p$ are also instances of $C$. 
-Examples of processes that can be divided include {\em continuous
-  movements} or {\em mass flow} processes, for which all parts have a
-{\em speed} or {\em flow rate} attribute. On the other hand, some
+Examples of processes that can be divided include {\em Continuous
+  movement} or {\em Mass flow} processes, for which all parts have a
+{\em Speed} or {\em Flow rate} attribute. On the other hand, some
 processes can be subdivided into stages of activity and stages of
-inactivity and cannot arbitrarily be divided. For example, a process
-of {\em heart beating} has periods of activity (a single heart beat)
-and inactivity. Consequently, not all parts of the process have a {\em
-  heart rate} attribute.
-
-We may further attribute a {\em frequency} or {\em rate} to an object
+inactivity (with respect to a particular process type) and cannot
+arbitrarily be divided. For example, a process of {\em Heart beating}
+has periods of activity in which a heart beat occurs and periods in
+which no heart beat process occurs. Consequently, not all parts of the
+process have a {\em Heart rate} ({\em Rate of heart beating})
+attribute.
+
+We may further attribute a {\em Frequency} or {\em Rate} to an object
 instead of a process. For example, a heart that beats {\em now} with a
@@
 convention, we call an abnormality of multiple occurrences of $X$ {\em
-  abnormality of the regulation of $X$}.
-
-\begin{table*}
-  \centering
-  \begin{tabular}{p{2.8cm}|p{4cm}|p{4cm}|p{4cm}}
-    & Increased flow rate & Normal flow rate & Decreased
-    flow rate \\
-    \hline
-    Increased frequency &increased total flow rate &increased total
-    flow rate &?\\
-    Normal frequency &increased total flow rate &normal total flow
-    rate &decreased total flow rate\\
-    Decreased frequency &?&decreased total flow rate &decreased total
-    flow rate\\
-    \hline
-  \end{tabular}
-  \caption{\label{tbl:flow}Interdependency for the attribute {\em
-      Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
-      rate} is an attribute of {\em Regulation of cytoplasmic
-      streaming} processes, while {\em Cytoplasmic flow rate} is an
-    attribute of individual {\em cytoplasmic streaming}
-    processes. Depending both on whether the cytoplasmic flow rate in
-    individual {\em cytoplasmic streaming} processes is increased or
-    decreased and whether the frequency of occurrence of {\em
-      cytoplasmic streaming} is increased or decreased, the total
-    cytoplasmic flow rate can be increased or decreased.}
-\end{table*}
+  Abnormality of the regulation of $X$}.
+
+% \begin{table*}
+%   \centering
+%   \begin{tabular}{p{2.8cm}|p{4cm}|p{4cm}|p{4cm}}
+%     & Increased flow rate & Normal flow rate & Decreased
+%     flow rate \\
+%     \hline
+%     Increased frequency &increased total flow rate &increased total
+%     flow rate &?\\
+%     Normal frequency &increased total flow rate &normal total flow
+%     rate &decreased total flow rate\\
+%     Decreased frequency &?&decreased total flow rate &decreased total
+%     flow rate\\
+%     \hline
+%   \end{tabular}
+%   \caption{\label{tbl:flow}Interdependency for the attribute {\em
+%       Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
+%       rate} is an attribute of {\em Regulation of cytoplasmic
+%       streaming} processes, while {\em Cytoplasmic flow rate} is an
+%     attribute of individual {\em cytoplasmic streaming}
+%     processes. Depending both on whether the cytoplasmic flow rate in
+%     individual {\em cytoplasmic streaming} processes is increased or
+%     decreased and whether the frequency of occurrence of {\em
+%       cytoplasmic streaming} is increased or decreased, the total
+%     cytoplasmic flow rate can be increased or decreased.}
+% \end{table*}
 
@@
 
-\bibliographystyle{plain}
-
-\bibliography{/home/leechuck/Documents/papers/bibtex/lc}
+% \bibliographystyle{natbib}
+
+% \bibliography{/home/leechuck/Documents/papers/bibtex/lc}
+\begin{thebibliography}{}
+
+\bibitem[Ashburner {\em et~al.}(2000)Ashburner {\em
+  et~al.}]{Ashburner2000short}
+Ashburner, M. {\em et~al.} (2000).
+\newblock Gene ontology: tool for the unification of biology.
+\newblock {\em Nature Genetics\/}, {\bf 25}(1).
+
+\bibitem[Bada and Hunter(2007)Bada and Hunter]{Bada2007a}
+Bada, M. and Hunter, L. (2007).
+\newblock Enrichment of obo ontologies.
+\newblock {\em Journal of Biomedical Informatics\/}, {\bf 40}(3), 300--315.
+
+\bibitem[Bada {\em et~al.}(2004)Bada, Stevens, Goble, Gil, Ashburner, Blake,
+  Cherry, Harris, and Lewis]{Bada2004}
+Bada, M., Stevens, R., Goble, C., Gil, Y., Ashburner, M., Blake, J.~A., Cherry,
+  M.~J., Harris, M., and Lewis, S. (2004).
+\newblock A short study on the success of the gene ontology.
+\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
+  Web\/}, {\bf 1}(2), 235--240.
+
+\bibitem[Bard {\em et~al.}(2005)Bard, Rhee, and Ashburner]{Bard2005}
+Bard, J., Rhee, S.~Y., and Ashburner, M. (2005).
+\newblock An ontology for cell types.
+\newblock {\em Genome Biology\/}, {\bf 6}(2).
+
+\bibitem[Berners-Lee {\em et~al.}(2001)Berners-Lee, Hendler, Lassila, {\em
+  et~al.}]{Berners-Lee2001}
+Berners-Lee, T., Hendler, J., Lassila, O., {\em et~al.} (2001).
+\newblock {The Semantic Web}.
+\newblock {\em Scientific American\/}, {\bf 284}(5), 28--37.
+
+\bibitem[Conrad {\em et~al.}(2011)Conrad, W\"{u}nsche, Tan, Bulkescher,
+  Sieckmann, Verissimo, Edelstein, Walter, Liebel, Pepperkok, and
+  Ellenberg]{Conrad2011}
+Conrad, C., W\"{u}nsche, A., Tan, T. H.~H., Bulkescher, J., Sieckmann, F.,
+  Verissimo, F., Edelstein, A., Walter, T., Liebel, U., Pepperkok, R., and
+  Ellenberg, J. (2011).
+\newblock Micropilot: automation of fluorescence microscopy-based imaging for
+  systems biology.
+\newblock {\em Nature methods\/}, {\bf 8}(3), 246--249.
+
+\bibitem[Cook {\em et~al.}(2011)Cook, Bookstein, and Gennari]{opb}
+Cook, D.~L., Bookstein, F.~L., and Gennari, J.~H. (2011).
+\newblock Physical properties of biological entities: An introduction to the
+  ontology of physics for biology.
+\newblock {\em PLoS ONE\/}, {\bf 6}(12), e28708.
+
+\bibitem[Eilbeck {\em et~al.}(2005)Eilbeck, Lewis, Mungall, Yandell, Stein,
+  Durbin, and Ashburner]{Eilbeck2005}
+Eilbeck, K., Lewis, S.~E., Mungall, C.~J., Yandell, M., Stein, L., Durbin, R.,
+  and Ashburner, M. (2005).
+\newblock The sequence ontology: A tool for the unification of genome
+  annotations.
+\newblock {\em Genome Biology\/}, {\bf 6}(R55).
+
+\bibitem[Engel {\em et~al.}(2010)Engel {\em et~al.}]{aposhort}
+Engel, S.~R. {\em et~al.} (2010).
+\newblock Saccharomyces genome database provides mutant phenotype data.
+\newblock {\em Nucleic acids research\/}, {\bf 38}(Database issue).
+
+\bibitem[Fuchs {\em et~al.}(2010)Fuchs, Pau, Kranz, Sklyar, Budjan, Steinbrink,
+  Horn, Pedal, Huber, and Boutros]{Fuchs2010}
+Fuchs, F., Pau, G., Kranz, D., Sklyar, O., Budjan, C., Steinbrink, S., Horn,
+  T., Pedal, A., Huber, W., and Boutros, M. (2010).
+\newblock Clustering phenotype populations by genome-wide {RNAi} and
+  multiparametric imaging.
+\newblock {\em Molecular Systems Biology\/}, {\bf 6}.
+
+\bibitem[{Gene Ontology Consortium}(2010){Gene Ontology Consortium}]{go2010}
+{Gene Ontology Consortium} (2010).
+\newblock The gene ontology in 2010: extensions and refinements.
+\newblock {\em Nucleic acids research\/}, {\bf 38}(Database issue), D331--335.
+
+\bibitem[Gkoutos and Hoehndorf(2011)Gkoutos and Hoehndorf]{obml2011h1}
+Gkoutos, G.~V. and Hoehndorf, R. (2011).
+\newblock Ontology-based cross-species integration and analysis of
+  saccharomyces cerevisiae phenotypes.
+\newblock In {\em Proceedings of the 3rd Workshop for Ontologies in Biomedicine
+  and Life sciences (OBML)\/}.
+
+\bibitem[Gkoutos {\em et~al.}(2005)Gkoutos, Green, Mallon, Hancock, and
+  Davidson]{Gkoutos2005}
+Gkoutos, G.~V., Green, E.~C., Mallon, A.-M.~M., Hancock, J.~M., and Davidson,
+  D. (2005).
+\newblock {Using ontologies to describe mouse phenotypes.}
+\newblock {\em Genome biology\/}, {\bf 6}(1).
+
+\bibitem[Gkoutos {\em et~al.}(2009)Gkoutos, Mungall, Dolken, Ashburner, Lewis,
+  Hancock, Schofield, Kohler, and Robinson]{Gkoutos2009b}
+Gkoutos, G.~V., Mungall, C., Dolken, S., Ashburner, M., Lewis, S., Hancock, J.,
+  Schofield, P., Kohler, S., and Robinson, P.~N. (2009).
+\newblock Entity/quality-based logical definitions for the human skeletal
+  phenome using {PATO}.
+\newblock {\em Annual International Conference of the IEEE Engineering in
+  Medicine and Biology Society.}, {\bf 1}, 7069--7072.
+
+\bibitem[Goble and Stevens(2008)Goble and Stevens]{goble}
+Goble, C. and Stevens, R. (2008).
+\newblock State of the nation in data integration for bioinformatics.
+\newblock {\em Journal of Biomedical Informatics\/}, {\bf 41}(5), 687--693.
+
+\bibitem[Grau {\em et~al.}(2008)Grau, Horrocks, Motik, Parsia, Patelschneider,
+  and Sattler]{Grau2008}
+Grau, B., Horrocks, I., Motik, B., Parsia, B., Patelschneider, P., and Sattler,
+  U. (2008).
+\newblock {OWL} 2: The next step for {OWL}.
+\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
+  Web\/}, {\bf 6}(4), 309--322.
+
+\bibitem[Gruber(1995)Gruber]{Gruber1995}
+Gruber, T.~R. (1995).
+\newblock Toward principles for the design of ontologies used for knowledge
+  sharing.
+\newblock {\em International Journal of Human-Computer Studies\/}, {\bf
+  43}(5-6).
+
+\bibitem[Guarino(1998)Guarino]{Guarino1998}
+Guarino, N. (1998).
+\newblock Formal ontology and information systems.
+\newblock In N.~Guarino, editor, {\em Proceedings of the 1st International
+  Conference on Formal Ontologies in Information Systems\/}, pages 3--15. IOS
+  Press.
+
+\bibitem[Herre {\em et~al.}(2006)Herre, Heller, Burek, Hoehndorf, Loebe, and
+  Michalek]{Herre2006}
+Herre, H., Heller, B., Burek, P., Hoehndorf, R., Loebe, F., and Michalek, H.
+  (2006).
+\newblock {G}eneral {F}ormal {O}ntology ({GFO}) -- {A} foundational ontology
+  integrating objects and processes [{V}ersion 1.0].
+\newblock Onto-Med Report~8, IMISE, University of Leipzig, Leipzig, Germany.
+
+\bibitem[Hoehndorf {\em et~al.}(2010a)Hoehndorf, Oellrich, and
+  Rebholz-Schuhmann]{Hoehndorf2010phene}
+Hoehndorf, R., Oellrich, A., and Rebholz-Schuhmann, D. (2010a).
+\newblock Interoperability between phenotype and anatomy ontologies.
+\newblock {\em Bioinformatics\/}, {\bf 26}(24), 3112 -- 3118.
+
+\bibitem[Hoehndorf {\em et~al.}(2010b)Hoehndorf, Oellrich, Dumontier, Kelso,
+  Rebholz-Schuhmann, and Herre]{Hoehndorf2010patterns}
+Hoehndorf, R., Oellrich, A., Dumontier, M., Kelso, J., Rebholz-Schuhmann, D.,
+  and Herre, H. (2010b).
+\newblock Relations as patterns: Bridging the gap between {OBO} and {OWL}.
+\newblock {\em BMC Bioinformatics\/}, {\bf 11}(1), 441+.
+
+\bibitem[Horridge {\em et~al.}(2007)Horridge, Bechhofer, and
+  Noppens]{Horridge2007}
+Horridge, M., Bechhofer, S., and Noppens, O. (2007).
+\newblock Igniting the {OWL} 1.1 touch paper: The {OWL} {API}.
+\newblock In {\em Proceedings of OWLED 2007: Third International Workshop on
+  OWL Experiences and Directions\/}.
+
+\bibitem[Horrocks(2007)Horrocks]{Horrocks2007}
+Horrocks, I. (2007).
+\newblock {OBO} flat file format syntax and semantics and mapping to {OWL}
+  {W}eb {O}ntology {L}anguage.
+\newblock Technical report, University of Manchester.
+
+\bibitem[Kazakov {\em et~al.}(2011)Kazakov, Kr{\"o}tzsch, and
+  Siman\v{c}\'{i}k]{Kazakov2011}
+Kazakov, Y., Kr{\"o}tzsch, M., and Siman\v{c}\'{i}k, F. (2011).
+\newblock Unchain my $\mathcal{EL}$ reasoner.
+\newblock In {\em Proceedings of the 23rd International Workshop on Description
+  Logics (DL'10)\/}, CEUR Workshop Proceedings. CEUR-WS.org.
+
+\bibitem[Lock and Strömblad(2010)Lock and Strömblad]{Lock2010}
+Lock, J.~G. and Strömblad, S. (2010).
+\newblock Systems microscopy: an emerging strategy for the life sciences.
+\newblock {\em Experimental Cell Research\/}, {\bf 316}(8), 1438--1444.
+
+\bibitem[Motik {\em et~al.}(2009)Motik, Grau, Horrocks, Wu, Fokoue, and
+  Lutz]{owlprofiles}
+Motik, B., Grau, B.~C., Horrocks, I., Wu, Z., Fokoue, A., and Lutz, C. (2009).
+\newblock Owl 2 web ontology language: Profiles.
+\newblock Recommendation, World Wide Web Consortium (W3C).
+
+\bibitem[Mungall {\em et~al.}(2010a)Mungall, Gkoutos, Smith, Haendel, Lewis,
+  and Ashburner]{Mungall2010}
+Mungall, C., Gkoutos, G., Smith, C., Haendel, M., Lewis, S., and Ashburner, M.
+  (2010a).
+\newblock Integrating phenotype ontologies across multiple species.
+\newblock {\em Genome Biology\/}, {\bf 11}(1), R2+.
+
+\bibitem[Mungall {\em et~al.}(2010b)Mungall, Bada, Berardini, Deegan, Ireland,
+  Harris, Hill, and Lomax]{Mungall2010go}
+Mungall, C.~J., Bada, M., Berardini, T.~Z., Deegan, J., Ireland, A., Harris,
+  M.~A., Hill, D.~P., and Lomax, J. (2010b).
+\newblock Cross-product extensions of the gene ontology.
+\newblock {\em Journal of biomedical informatics\/}.
+\newblock in press.
+
+\bibitem[Neumann {\em et~al.}(2010)Neumann, Walter, Hériché, Bulkescher,
+  Erfle, Conrad, Rogers, Poser, Held, Liebel, and et~al.]{Neumann2010}
+Neumann, B., Walter, T., Hériché, J.-K., Bulkescher, J., Erfle, H., Conrad,
+  C., Rogers, P., Poser, I., Held, M., Liebel, U., and et~al. (2010).
+\newblock Phenotypic profiling of the human genome by time-lapse microscopy
+  reveals cell division genes.
+\newblock {\em Nature\/}, {\bf 464}(7289), 721--727.
+
+\bibitem[Ogren {\em et~al.}(2004)Ogren, Cohen, Acquaah-Mensah, Eberlein, and
+  Hunter]{Ogren2004}
+Ogren, P.~V., Cohen, K.~B., Acquaah-Mensah, G.~K., Eberlein, J., and Hunter, L.
+  (2004).
+\newblock The compositional structure of gene ontology terms.
+\newblock {\em Pac Symp Biocomput\/}, pages 214--225.
+
+\bibitem[\"{O}zg\"{o}vde and Gr\"{u}ninger(2010)\"{O}zg\"{o}vde and
+  Gr\"{u}ninger]{Gruninger2010}
+\"{O}zg\"{o}vde, A. and Gr\"{u}ninger, M. (2010).
+\newblock Foundational process relations in bio-ontologies.
+\newblock In {\em Proceeding of the 2010 conference on Formal Ontology in
+  Information Systems\/}, pages 243--256, Amsterdam, The Netherlands, The
+  Netherlands. IOS Press.
+
+\bibitem[Robinson {\em et~al.}(2008)Robinson, Koehler, Bauer, Seelow, Horn, and
+  Mundlos]{Robinson2008}
+Robinson, P.~N., Koehler, S., Bauer, S., Seelow, D., Horn, D., and Mundlos, S.
+  (2008).
+\newblock The human phenotype ontology: a tool for annotating and analyzing
+  human hereditary disease.
+\newblock {\em American journal of human genetics\/}, {\bf 83}(5), 610--615.
+
+\bibitem[Schindelman {\em et~al.}(2011)Schindelman, Fernandes, Bastiani, Yook,
+  and Sternberg]{wpo}
+Schindelman, G., Fernandes, J., Bastiani, C., Yook, K., and Sternberg, P.
+  (2011).
+\newblock Worm phenotype ontology: integrating phenotype data within and beyond
+  the c. elegans community.
+\newblock {\em BMC Bioinformatics\/}, {\bf 12}(1), 32.
+
+\bibitem[Schmitz {\em et~al.}(2010)Schmitz {\em et~al.}]{Schmitz2010short}
+Schmitz, M. H.~A. {\em et~al.} (2010).
+\newblock Live-cell imaging rnai screen identifies pp2a–b55$\alpha$ and
+  importin-$\beta$1 as key mitotic exit regulators in human cells.
+\newblock {\em Nature Cell Biology\/}, {\bf 12}, 886--893.
+
+\bibitem[Schofield {\em et~al.}(2011)Schofield, Sundberg, Hoehndorf, and
+  Gkoutos]{Schofield2011}
+Schofield, P.~N., Sundberg, J.~P., Hoehndorf, R., and Gkoutos, G.~V. (2011).
+\newblock New approaches to the representation and analysis of phenotype
+  knowledge in human diseases and their animal models.
+\newblock {\em Briefings in Functional Genomics\/}, {\bf 10}(5), 258--265.
+
+\bibitem[Smith {\em et~al.}(2007)Smith {\em et~al.}]{Smith2007short}
+Smith, B. {\em et~al.} (2007).
+\newblock The {OBO} {F}oundry: coordinated evolution of ontologies to support
+  biomedical data integration.
+\newblock {\em Nat Biotech\/}, {\bf 25}(11), 1251--1255.
+
+\bibitem[Smith {\em et~al.}(2004)Smith, Goldsmith, and Eppig]{Smith2004}
+Smith, C.~L., Goldsmith, C.-A.~W., and Eppig, J.~T. (2004).
+\newblock The mammalian phenotype ontology as a tool for annotating, analyzing
+  and comparing phenotypic information.
+\newblock {\em Genome Biology\/}, {\bf 6}(1), R7.
+
+\end{thebibliography}
 
